---
title: "OMOP Image Occurrence"
execute:
  echo: false
---
```{r setup, include=FALSE}

# Your existing initialization
rm(list=ls())
source("/workspaces/starr-oncology-data-lake-arpah/src/R/all_function.R", encoding = "UTF-8")
yaml_file_path <- "/workspaces/starr-oncology-data-lake-arpah/src/sql_params.yml"
```

```{r, message=FALSE, warning=FALSE, results='hide'}
#| context: global
install.packages("forcats")
library(forcats)
sql_file_path_image_mod <- "/workspaces/starr-oncology-data-lake-arpah/src/sql/image_occ/image_occ_modality.sql"
res_img_mod <- fetch_data_from_sql_file(sql_file_path_image_mod, yaml_file_path)
head(res_img_mod)
res_img_mod$variable_name<-NULL
colnames(res_img_mod)=c("modality_type", "series_count", "person_count")
```

```{r,  warning=FALSE, results='hide'}
#| context: global

Value <- c("AR", "AS", "ASMT", "AU", "BDUS", "BI", "BMD", "CD", "CF", "CP", 
           "CR", "CS", "CT", "DD", "DF", "DG", "DM", "DOC", "DS", "DX", 
           "EC", "ECG", "EPS", "ES", "FA", "FID", "FS", "GM", "HC", "HD", 
           "IO", "IOL", "IVOCT", "IVUS", "KER", "KO", "LEN", "LP", "LS", 
           "MA", "MG", "MR", "MS", "NM", "OAM", "OCT", "OP", "OPM", "OPR", 
           "OPT", "OPV", "OSS", "OT", "PLAN", "PR", "PT", "PX", "REG", 
           "RESP", "RF", "RG", "RTDOSE", "RTIMAGE", "RTPLAN", "RTRECORD", 
           "RTSTRUCT", "RWV", "SEG", "SM", "SMR", "SR", "SRF", "ST", 
           "STAIN", "TG", "US", "VA", "VF", "XA", "XC")

Description <- c("Autorefraction", "Angioscopy", "Content Assessment Results", 
                 "Audio", "Bone Densitometry (ultrasound)", "Biomagnetic imaging", 
                 "Bone Densitometry (X-Ray)", "Color flow Doppler", 
                 "Cinefluorography", "Colposcopy", "Computed Radiography", 
                 "Cystoscopy", "Computed Tomography", "Duplex Doppler", 
                 "Digital fluoroscopy", "Diaphanography", "Digital microscopy", 
                 "Document", "Digital Subtraction Angiography", "Digital Radiography", 
                 "Echocardiography", "Electrocardiography", 
                 "Cardiac Electrophysiology", "Endoscopy", 
                 "Fluorescein angiography", "Fiducials", "Fundoscopy", 
                 "General Microscopy", "Hard Copy", "Hemodynamic Waveform", 
                 "Intra-Oral Radiography", "Intraocular Lens Data", 
                 "Intravascular Optical Coherence Tomography", 
                 "Intravascular Ultrasound", "Keratometry", 
                 "Key Object Selection", "Lensometry", "Laparoscopy", 
                 "Laser surface scan", "Magnetic resonance angiography", 
                 "Mammography", "Magnetic Resonance", 
                 "Magnetic resonance spectroscopy", "Nuclear Medicine", 
                 "Ophthalmic Axial Measurements", 
                 "Optical Coherence Tomography (non-Ophthalmic)", 
                 "Ophthalmic Photography", "Ophthalmic Mapping", 
                 "Ophthalmic Refraction", "Ophthalmic Tomography", 
                 "Ophthalmic Visual Field", "Optical Surface Scan", 
                 "Other", "Plan", "Presentation State", 
                 "Positron emission tomography (PET)", "Panoramic X-Ray", 
                 "Registration", "Respiratory Waveform", 
                 "Radio Fluoroscopy", "Radiographic imaging (conventional film/screen)", 
                 "Radiotherapy Dose", "Radiotherapy Image", 
                 "Radiotherapy Plan", "RT Treatment Record", 
                 "Radiotherapy Structure Set", "Real World Value Map", 
                 "Segmentation", "Slide Microscopy", "Stereometric Relationship", 
                 "SR Document", "Subjective Refraction", 
                 "Single-photon emission computed tomography (SPECT)", 
                 "Automated Slide Stainer", "Thermography", 
                 "Ultrasound", "Visual Acuity", "Videofluorography", 
                 "X-Ray Angiography", "External-camera Photography")

retired <- c(NA, "Retired", NA, NA, NA, NA, NA, "Retired", "Retired", 
             "Retired", NA, "Retired", NA, "Retired", "Retired", NA, 
             "Retired", NA, NA, NA, "Retired", NA, NA, NA, "Retired", 
             NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, "Retired", 
             "Retired", NA, NA, "Retired", NA, "Retired", NA, 
             "Retired", NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, 
             NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, 
             NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, 
             "Retired", NA)

# Create the data frame
dat_modality <- data.frame(Value, Description, retired, stringsAsFactors = FALSE)

```
### Summary:

In the OMOP Common Data Model (CDM), the "Image_occurrence" table provides a structured way to represent medical imaging events, bridging the gap between imaging and observational research by integrating image-based measurements into the CDM. 

The OMOP image_occurrence table captures records of medical imaging events at the series level, representing events where one or more related images are acquired as part of a clinical procedure. 

Each record includes metadata such as the modality (e.g., CT, MRI) and the anatomic site of the image. The table maintains foreign key links to the procedure_occurrence and visit_occurrence tables, enabling contextual association with the related clinical procedure and patient encounter. 

At Stanford, this table has been extended to include a link to the note table, allowing integration of imaging data with relevant clinical documentation.

```{r, message=FALSE, warning=FALSE, results='hide'}
## number of pts and number of series 
path_img <- "/workspaces/starr-oncology-data-lake-arpah/src/sql/image_occ/image_occ_person.sql"
pt_img <- fetch_data_from_sql_file(path_img, yaml_file_path)
pt_img
den_series=as.numeric(pt_img$counts)[2]
den_pts=as.numeric(pt_img$counts)[1]
```

As of May 2025, this table included images for  `r format(pt_img$counts[1], big.mark = ",")` patients with `r format(pt_img$counts[2], big.mark = ",")` series. **Series** are defined as individual imaging acquisitions within a study.


```{r, message=FALSE, warning=FALSE, results='hide'}
## basic counts 
sql_file_path_image<- "/workspaces/starr-oncology-data-lake-arpah/src/sql/image_occ/image_occ_basic.sql"
res<- fetch_data_from_sql_file(sql_file_path_image, yaml_file_path)
head(res)


##- **>`r format(res$counts[res$flag=="size_gb"], big.mark = ",")` GB** of imaging data 
```
::: {.columns}

::: {.column width="40%"}
### Data Volume 
- Patient Count: `r format(pt_img$counts[1], big.mark = ",")`
- Series Count: `r format(pt_img$counts[2], big.mark = ",")`
- Study UID: `r format(res$counts[res$flag=="counts_study"], big.mark = ",")`
:::

::: {.column width="60%"}
### Data Components 
- `r format(res$counts[res$flag=="counts_modality_type"], big.mark = ",")` Modality types 
- `r format(res$counts[res$flag=="counts_anatomic_type"], big.mark = ",")` Anatomic sites 
:::



## 📊 Imaging Summary Results {.tabset}
This visualization summarizes distribution of Modality categories and Anatomic sites by series and patient counts for all cancer types and for thoracic patients. 
Note: We expect the number of patients with imaging data to increase per release. 

::: {.panel-tabset}

### All Cancer Types 
## Modality Types
Top modality types for series with 5000 and more.
```{r, message=FALSE, warning=FALSE, results='hide'}
colnames(dat_modality)[1]<-"modality_type"
res_img_mod=res_img_mod %>% left_join(dat_modality,by="modality_type" )
res_img_mod=res_img_mod %>% filter(person_count>10)
res_img_mod$retired<-NULL
```

```{r, message=FALSE, warning=FALSE}
## Number of series in each modality - bar plot to keep 
library(plotly)
df_bar <- res_img_mod %>% filter(series_count>5000)%>%
  arrange(desc(series_count)) %>% mutate(modality_type = fct_reorder(modality_type, series_count))

plot_ly(df_bar,
        y = ~modality_type,
        x = ~series_count,
        type = "bar",
        color = ~modality_type,
        colors = "Set3") %>%
  layout(xaxis = list(title = "Series Count"),
         yaxis = list(title = ""))%>%layout (title = "Number of Series per Modality")

```

```{r, message=FALSE, warning=FALSE, results='hide'}

## hiding the pie ## 
df_pie <- res_img_mod %>%
  group_by(modality_type) %>%  filter(series_count>5000)%>%
  arrange(desc(person_count)) 

plot_ly(
  df_pie,
  labels = ~modality_type,
  values = ~series_count,
  type = "pie",
  text = ~paste("Modality:", modality_type, "<br>Series Count:", format(series_count, big.mark = ",")),
  hoverinfo = "text",
  marker = list(colors = RColorBrewer::brewer.pal(n = max(3, nrow(df_pie)), "Set3")),
  textinfo = "label+percent"
) %>%
  layout(
    title = list(text = "Number of Series per Modality", x = 0.5),
    legend = list(
      orientation = "v",
      x = -0.1,     
      y = 0.5,     
      xanchor = "left",
      font = list(size = 12)
    )
  )
```


## Patient Population 
```{r, message=FALSE, warning=FALSE}
df_bar <- res_img_mod %>% filter(person_count>1000)%>%
  arrange(desc(person_count)) %>% mutate(modality_type = fct_reorder(modality_type, person_count))

plot_ly(df_bar,
        y = ~modality_type,
        x = ~person_count,
        type = "bar",
        color = ~modality_type,
        colors = "Set3") %>%
  layout(xaxis = list(title = "Person Count"),
         yaxis = list(title = ""))%>%layout (title = "Number of Patients per Modality")
```
## Modality Descriptions and Frequency by Series and Patients 
Please note that the modality descriptions are listed here https://www.dicomlibrary.com/dicom/modality/.

```{r, message=FALSE, warning=FALSE}
library(DT)
res_img_mod=res_img_mod %>% mutate(modality_description = paste0(Description, " (", modality_type, ")"))
res_img_mod$Description[is.na(res_img_mod$Description)]<-"Not Available"
#res_img_mod$modality_type<-NULL
res_img_mod<-res_img_mod %>% select (Description,modality_type,series_count,person_count)
DT::datatable(as.data.frame(res_img_mod), options = list(scrollX = TRUE, pageLength = 10))%>%
  formatRound(columns = c("person_count", "series_count"), digits = 0)

## https://www.dicomlibrary.com/dicom/modality/

## drop or regrp the documents and patterns by pixel data 
## round to 1 decimal not 2
```


## Anatomic Sites
Top anatomic sites for series with 5000 and more.

```{r, message=FALSE, warning=FALSE, results='hide'}
#| context: global

sql_file_path_image_an <- "/workspaces/starr-oncology-data-lake-arpah/src/sql/image_occ/image_occ_anatomic.sql"
res_ana <- fetch_data_from_sql_file(sql_file_path_image_an, yaml_file_path)
head(res_ana)
res_ana$variable_name<-NULL
colnames(res_ana)=c("anatomic_site", "series_count", "person_count")
res_ana=res_ana %>% filter(person_count>10)
```
```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(plotly)
library(DT)

df_bar_sr <- res_ana %>% filter(series_count>5000)%>%
  arrange(desc(series_count)) %>% mutate(anatomic_site = forcats::fct_reorder(anatomic_site, series_count))

plot_ly(df_bar_sr,
        y = ~anatomic_site,
        x = ~series_count,
        type = "bar",
        color = ~anatomic_site,
        colors = "Set3") %>%
  layout(xaxis = list(title = "Series Count"),
         yaxis = list(title = "")) %>%layout (title = "Number of Series per Anatomic Site")

```
## Patient Population 

```{r, message=FALSE, warning=FALSE}
install.packages("forcats")
library(forcats)
df_bar <- res_ana %>% filter(person_count>1000)%>%
  arrange(desc(person_count)) %>% mutate(anatomic_site = forcats::fct_reorder(anatomic_site, person_count))

plot_ly(df_bar,
        y = ~anatomic_site,
        x = ~person_count,
        type = "bar",
        color = ~anatomic_site,
        colors = "Set3") %>%
  layout(xaxis = list(title = "Person Count"),
         yaxis = list(title = "")) %>%layout (title = "Number of Patients per Anatomic Site")
```
## Anatomic Site Descriptions and Frequency by Series and Patients 

```{r, message=FALSE, warning=FALSE}
library(DT)

res_ana=res_ana %>% select (anatomic_site,series_count,person_count)

DT::datatable(as.data.frame(res_ana), options = list(scrollX = TRUE, pageLength = 10))%>%
  formatRound(columns = c("person_count", "series_count"), digits = 0)

```
### Thoracic Cancer 
```{r, message=FALSE, warning=FALSE, results='hide'}
## basic counts 
sql_thr<- "/workspaces/starr-oncology-data-lake-arpah/src/sql/image_occ/image_occ_thoracic/image_occ_person_thr.sql"
res_den_thr<- fetch_data_from_sql_file(sql_thr, yaml_file_path)
head(res_den_thr)
```
The Thoracic Cancer cohort includes patients diagnosed with primary malignancies located in the thoracic cavity, as defined by the International Classification of Diseases for Oncology, 3rd Edition (ICD-O-3) primary site codes. Specifically, the cohort is defined by ICD-O-3 topography codes (C34 and C37). 
As of May 2025, From `r format(res_den_thr$counts[1], big.mark = ",")` patients with thoracic cancer, `r format(res_den_thr$counts[2], big.mark = ",")` had imaging data. 



## Modality Types
Top modality types for series above 500. 
```{r, message=FALSE, warning=FALSE, results='hide'}
## modality
sql_thr_mod<- "/workspaces/starr-oncology-data-lake-arpah/src/sql/image_occ/image_occ_thoracic/image_occ_modality_thr.sql"
res_thr<- fetch_data_from_sql_file(sql_thr_mod, yaml_file_path)
head(res_thr)
res_thr
```

```{r, message=FALSE, warning=FALSE}
# Prepare the data
res_thr <- res_thr %>% filter (series_count>500) %>%
  arrange(desc(series_count)) %>%
  mutate(
    modality_source_value = factor(modality_source_value, levels = unique(modality_source_value)),
    # Calculate percentage of patients
    patient_percentage = round((unique_person_count / max(unique_person_count)) * 100, 1)
  )

# Create the plot
plot <- plot_ly() %>%
  # Add bars for series count
  add_trace(
    data = res_thr,
    x = ~modality_source_value,
    y = ~series_count,
    type = "bar",
    name = "Series Count",
    marker = list(color =  "#8DD3C7")
  ) %>%
  # Add line for person count
  add_trace(
    data = res_thr,
    x = ~modality_source_value,
    y = ~unique_person_count,
    type = "scatter",
    mode = "lines+markers",
    name = "Person Count",
    yaxis = "y2",
    line = list(color = "#BEBADA", width = 2),
    marker = list(color = "#BEBADA", size = 8)
  ) %>%
  layout(
    title = "Imaging Modalities: Series Count vs Patient Count",
    xaxis = list(
      title = "Modality Type",
      tickangle = 45
    ),
    yaxis = list(
      title = "Series Count",
      side = "left"
    ),
    yaxis2 = list(
      title = "",
      side = "right",
      overlaying = "y"
    ),
    legend = list(x = 0.8, y = 0.9),
    margin = list(b = 100),  # Add bottom margin for rotated labels
    hovermode = "x unified"
  )

plot
```
## Anatomic Sites 
```{r, message=FALSE, warning=FALSE, results='hide'}
## anatomic
sql_thr_ana<- "/workspaces/starr-oncology-data-lake-arpah/src/sql/image_occ/image_occ_thoracic/image_occ_anatomic._thr.sql"
res_ana_thr<- fetch_data_from_sql_file(sql_thr_ana, yaml_file_path)
head(res_ana_thr)

```
```{r, message=FALSE, warning=FALSE}
# Prepare the data
res_ana_thr <- res_ana_thr %>% filter (series_count>500) %>%
  arrange(desc(series_count)) %>%
  mutate(
    anatomic_site_source_value = factor(anatomic_site_source_value, levels = unique(anatomic_site_source_value)),
    # Calculate percentage of patients
    patient_percentage = round((unique_person_count / max(unique_person_count)) * 100, 1)
  )

# Create the plot
plot <- plot_ly() %>%
  # Add bars for series count
  add_trace(
    data = res_ana_thr,
    x = ~anatomic_site_source_value,
    y = ~series_count,
    type = "bar",
    name = "Series Count",
    marker = list(color =  "#8DD3C7")
  ) %>%
  # Add line for person count
  add_trace(
    data = res_ana_thr,
    x = ~anatomic_site_source_value,
    y = ~unique_person_count,
    type = "scatter",
    mode = "lines+markers",
    name = "Person Count",
    yaxis = "y2",
    line = list(color = "#BEBADA", width = 2),
    marker = list(color = "#BEBADA", size = 8)
  ) %>%
  layout(
    title = "Imaging Anatomic Site: Series Count vs Patient Count",
    xaxis = list(
      title = "Anatomic Sites",
      tickangle = 45
    ),
    yaxis = list(
      title = "Series Count",
      side = "left"
    ),
    yaxis2 = list(
      title = "",
      side = "right",
      overlaying = "y"
    ),
    legend = list(x = 0.8, y = 0.9),
    margin = list(b = 100),  # Add bottom margin for rotated labels
    hovermode = "x unified"
  )

plot
```
:::

