---
title: "Neural Frame"
execute:
  echo: false
---
```{r, message=FALSE, warning=FALSE, results='hide'}
#| context: global

rm(list=ls())
source("/workspaces/starr-oncology-data-lake-arpah/src/R/all_function.R", encoding = "UTF-8")
yaml_file_path <-  "/workspaces/starr-oncology-data-lake-arpah/src/sql_params.yml"

```


```{r, message=FALSE, warning=FALSE, results='hide'}
#| context: global

sql_file_path_image_mod <- "/workspaces/starr-oncology-data-lake-arpah/src/sql/image_occ/image_occ_modality.sql"
res_img_mod <- fetch_data_from_sql_file(sql_file_path_image_mod, yaml_file_path)
head(res_img_mod)
res_img_mod$variable_name<-NULL
colnames(res_img_mod)=c("modality_type", "series_count", "person_count")
```
### Neural Frame:
Class of Case divides cases into two groups. Analytic cases (codes 00-22) are those that are required by CoC to be abstracted because of the program’s primary responsibility in managing the cancer. Analytic cases are grouped according to the location of diagnosis and treatment. Treatment and outcome reports may be limited to analytic cases. Nonanalytic cases (codes 30-49 and 99) may be abstracted by the facility to meet central registry requirements or because of a request by the facility’s cancer program. Nonanalytic cases are grouped according to the reason a patient who received care at the facility is nonanalytic, or the reason a patient who never received care at the facility may have been abstracted.