---
title: "Neural Frame Metrics"
execute:
  echo: false
---
```{r load functions, message=FALSE, warning=FALSE, results='hide'}
#| context: global

rm(list=ls())
source("/workspaces/starr-oncology-data-lake-arpah/src/R/all_function.R", encoding = "UTF-8")
yaml_file_path <-  "/workspaces/starr-oncology-data-lake-arpah/src/sql_params.yml"
library(RColorBrewer)  # Add RColorBrewer package
```

```{r nf denominator, message=FALSE, warning=FALSE, results='hide'}
sql_den<- "/workspaces/starr-oncology-data-lake-arpah/src/sql/scr/thoracic/scr_den.sql"
nf_den <- fetch_data_from_sql_file(sql_den, yaml_file_path)
nf_den
```

```{r case status all ,message=FALSE, warning=FALSE, results='hide'}
sql_nf<- "/workspaces/starr-oncology-data-lake-arpah/src/sql/scr/thoracic/scr_case_status.sql"
case <- fetch_data_from_sql_file(sql_nf, yaml_file_path)
head(case)
case$nfcasestatus[case$nfcasestatus=="Incomplete - Edit"]<-"Incomplete"
case$flag="All Cancer"
```
### Summary:
The Stanford Cancer Registry data is integrated with the California Cancer Registry and updated monthly, encompassing approximately 800 variables related to various cancer. As of January 2022, the KACI® application from NeuralFrame was approved for abstracting and transmitting diagnosed cases, with Stanford adopting it in 2023. Neural Frame connects to EPIC to extract relevant data, including pathology notes.

Within Neural Frame, Class of Case (CoC) divides cases into two groups: cancer cases are categorized as **Analytic** (reportable with complete data) or **Non-analytic** (not reportable by law, with complete data only from 2022 onward).

 - **Analytic**: Analytic cases include patients with initial diagnoses and full or partial treatments at Stanford. Analytic cases (codes 00-22) are those that are required by CoC to be abstracted because of the program’s primary responsibility in managing the cancer. Analytic cases are grouped according to the location of diagnosis and treatment. 

 - **Non-analytic**: Non-analytic cases include recurrences, consults only, and in-transit patients. Non-analytic cases (codes 30-49 and 99) can be abstracted by the facility to fulfill central registry requirements or upon request from the facility’s cancer program. The grouping of non-analytic cases is based on the reason for their classification, whether they received care at the facility or were abstracted for other reasons. 

::: {.columns}

::: {.column width="50%"}
### 📊 Data Volume 
- Patients Count:  `r format(nf_den$unique_person_count[1], big.mark = ",")`
- Patients with Completed Cases: `r format(case$unique_person_count_case[case$nfcasestatus=="Completed"], big.mark = ",")`
:::

::: {.column width="50%"}
### 🧬 Data Components 
- Outcome
- Diagnoses 
- Treatment
- Miscellaneous
:::

## Thoracic Cancer Cohort
Thoracic cancer patients are identified based on their primary site descriptions in the Neural Frame diagnoses data, which include diagnoses of lung, bronchus, or thymus cancer.
The thoracic cancer cohort includes patients diagnosed with primary malignancies located in the thoracic cavity, as defined by the International Classification of Diseases for Oncology, 3rd Edition (ICD-O-3) primary site codes. Specifically, the cohort is defined by the following ICD-O-3 topography codes:

 - C34.0 – Main bronchus
 - C34.1 – Upper lobe, lung
 - C34.2 – Middle lobe, lung
 - C34.3 – Lower lobe, lung
 - C34.8 – Overlapping lesion of lung
 - C34.9 – Lung, not otherwise specified (NOS)
 - C37.9 – Thymus


```{r case status thoracic, message=FALSE, warning=FALSE, results='hide'}
sql_nf<- "/workspaces/starr-oncology-data-lake-arpah/src/sql/scr/thoracic/scr_case_status.sql"
case <- fetch_data_from_sql_file(sql_nf, yaml_file_path)
head(case)
case$nfcasestatus[case$nfcasestatus=="Incomplete - Edit"]<-"Incomplete"
case$flag="All Cancer"

# Calculate counts and percentages using unique person counts
case_summary <- aggregate(unique_person_count_case ~ nfcasestatus, data = case, sum)
case_df <- data.frame(
  Status = case_summary$nfcasestatus,
  Count = case_summary$unique_person_count_case,
  Percentage = round(case_summary$unique_person_count_case / (nf_den$unique_person_count) * 100, 1)
)
## code this number ##
# Sort by percentage in descending order
case_df <- case_df[order(-case_df$Percentage), ]
```




## Multiple Primary 
 
Within the thoracic cancer cohort, a subset of patients presents with multiple primary thoracic malignancies. This designation refers to the presence of two or more histologically distinct malignant tumors arising independently within the thoracic cavity, rather than as metastases from a single primary site. These neoplasms may occur synchronously, identified concurrently or within a short time interval, or metachronously, developing at distinct time points.

## Thoracic Cancer Metrics {.tabset}
This visualization summarizes the characteristics of thoracic cancer cases by:

 - Cases, Diagnoses, Staging
 - Treatment
 - Social History
 - Follow up Outcome

::: {.panel-tabset}

### Cases & Diagnoses

## Case Status
Cases are defined as completed cases
```{r, message=FALSE, warning=FALSE, results='hide'}
sql_thr_case<- "/workspaces/starr-oncology-data-lake-arpah/src/sql/scr/thoracic/scr_thoracic_case_status_person_metrics.sql"
thr_case <- fetch_data_from_sql_file(sql_thr_case, yaml_file_path)
head(thr_case)
thr_case$flag="Thoracic Cancer"
thr_case$nfcasestatus[thr_case$nfcasestatus=="Incomplete - Edit"]<-"Incomplete"


sql_thr<-"/workspaces/starr-oncology-data-lake-arpah/src/sql/scr/thoracic/scr_thoracic_denominator.sql"

thr_den=fetch_data_from_sql_file(sql_thr, yaml_file_path)
thr_den
```
```{r,message=FALSE, warning=FALSE, results='hide'}
# Calculate counts and percentages using unique person counts
case_summary_thr <- aggregate(patient_count ~ nfcasestatus, data = thr_case, sum)
case_thr <- data.frame(
  Status = case_summary_thr$nfcasestatus,
  Count = case_summary_thr$patient_count,
  Percentage = round(case_summary_thr$patient_count / (thr_den$unique_thoracic_cancer_pts) * 100)
)

```
```{r denominators, message=FALSE, warning=FALSE, results='hide'}
## thoracic cancer with complete cases 
sql_thr_c<-"/workspaces/starr-oncology-data-lake-arpah/src/sql/scr/thoracic/scr_thoracic_denominator_complete.sql"

thr_den_c=fetch_data_from_sql_file(sql_thr_c, yaml_file_path)
thr_den_c
## denominators ##
sum_thr_c=thr_den_c$unique_thoracic_cancer_pts
sum_thr=thr_den$unique_thoracic_cancer_pts
```
As of May 2025, there are  `r format(thr_den$unique_thoracic_cancer_pts, big.mark = ",")` unique thoracic patients with `r format(case_thr$Percentage[case_thr$Status=="Completed"], big.mark = ",")` % completed cases. 



```{r, message=FALSE, warning=FALSE}
# Create pie chart
library(ggplot2)
ggplot(case_thr, aes(x = "", y = Percentage, fill = Status)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  theme_minimal() +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.border = element_blank(),
    panel.grid = element_blank(),
    axis.ticks = element_blank(),
    axis.text.x = element_blank()
  ) +
  geom_text(aes(label = paste0(Status, "\n", Percentage, "%")), 
            position = position_stack(vjust = 0.5)) +
  ggtitle("Case Status Distribution of Thoracic Cancer Patients") +
  scale_fill_brewer(palette = "Set3")
```

## Patient Summary by Diagnosis Year 
The following distribution summarizes the number of diagnosed thoracic patients over time. 

```{r dx year calculate, message=FALSE, warning=FALSE, results='hide'}
sql_thr_yr<- "/workspaces/starr-oncology-data-lake-arpah/src/sql/scr/thoracic/scr_thoracic_completed_cases_dx_year_person_metrics.sql"
thr_yr <- fetch_data_from_sql_file(sql_thr_yr, yaml_file_path)

# Group years before 2000 and calculate percentages ##
thr_yr_grouped <- thr_yr %>%
  mutate(dx_year = as.character(ifelse(dx_year < 2000, "<2000", dx_year))) %>%
  group_by(dx_year) %>%
  summarize(patient_count = sum(patient_count)) %>%
  mutate(percentage = round(100 * patient_count / sum(patient_count), 1),
         label = paste0(percentage)) %>%
  arrange(dx_year)
```
```{r dx year barplot, message=FALSE, warning=FALSE}
library(plotly)

# Create an interactive bar plot using plotly
p <- plot_ly(data = thr_yr_grouped,
             x = ~dx_year,
             y = ~patient_count,
             type = "bar",
             text = ~label,
             textposition = "auto",
             marker = list(color ="#8B0000" ))  

# Add layout settings
p <- p %>% 
  layout(
    title = list(text = "Number of Thoracic Cancer Patients by Diagnosis Year"),
    xaxis = list(
      title = "Diagnosis Year",
      tickangle = 45
    ),
    yaxis = list(
      title = "Unique Patient Count"
    ),
    showlegend = FALSE
  )

# Display the plot
p
```

## Histology Distribution & ICD-O-3-Codes
Thoracic cancer encompasses a diverse group of histologic subtypes, with Non-Small Cell Lung Cancer **(NSCLC)** representing the majority of cases and Small Cell Lung Cancer **(SCLC)**.

Thoracic cancer is categorized into detailed histologic groups using ICD-O-3 codes and then further grouped into four overarching categories for simplified analysis. The main histology groups include:

- **Non-Small Cell Lung Carcinoma (NSCLC):** This group encompasses the majority of lung cancer and includes subtypes such as:
  - Adenocarcinoma (8140–8147)
  - Squamous Cell Carcinoma (8070–8080)
  - Large Cell Carcinoma (8010–8015)
  - Non-Small Cell Carcinoma NOS (8046)

- **Small Cell Lung Carcinoma (SCLC):** Represented by ICD-O-3 codes 8041–8045, this group is characterized by more aggressive, rapidly progressing tumors.

- **Other Specified Histologies:** These include a range of less common thoracic cancer such as:
  - Carcinoid Tumors (8240–8249)
  - Sarcoma (8800–8806)
  - Mesothelioma (9050–9053)
  - Acinar Cell Carcinoma (8550–8551)
  - Bronchiolo-alveolar Carcinoma (8250–8257)
  - Neoplasms (8000–8005)
  - Papillary Adenocarcinoma (8260–8265)
  - Papillary Carcinoma (8050–8052)

- **Other/Unspecified Histologies:** Any histology not falling into the categories above.



```{r icdo mapping, message=FALSE, warning=FALSE, results='hide'}
# Load required libraries
library(plotly)
library(dplyr)
library(RColorBrewer)
library(htmlwidgets)

sql_thr_his <- "/workspaces/starr-oncology-data-lake-arpah/src/sql/scr/thoracic/scr_thoracic_completed_cases_histology_person_metrics.sql"
thr_his <- fetch_data_from_sql_file(sql_thr_his, yaml_file_path)

# Add histology grouping
thr_his <- thr_his %>%
  mutate(histology_group = case_when(
    # Non-Small Cell Lung Cancer (NSCLC) Groups
    between(as.numeric(histologicTypeIcdO3), 8140, 8147) ~ 'Adenocarcinoma',
    between(as.numeric(histologicTypeIcdO3), 8070, 8080) ~ 'Squamous Cell Carcinoma',
    between(as.numeric(histologicTypeIcdO3), 8012, 8014) ~ 'Large Cell Carcinoma',
    histologicTypeIcdO3 == '8046' ~ 'Non-Small Cell Carcinoma NOS',
    
    # Small Cell Lung Cancer (SCLC)
    between(as.numeric(histologicTypeIcdO3), 8041, 8045) ~ 'Small Cell Carcinoma',
    
    # Neuroendocrine Tumors
    between(as.numeric(histologicTypeIcdO3), 8240, 8249) ~ 'Carcinoid Tumor',
    
    # Other specified types
    between(as.numeric(histologicTypeIcdO3), 8800, 8806) ~ 'Sarcoma',
    between(as.numeric(histologicTypeIcdO3), 9050, 9053) ~ 'Mesothelioma',
    between(as.numeric(histologicTypeIcdO3), 8550, 8551) ~ 'Acinar Cell Carcinoma',
    between(as.numeric(histologicTypeIcdO3), 8250, 8257) ~ 'Bronchiolo-alveolar Carcinoma',
    between(as.numeric(histologicTypeIcdO3), 8000, 8005) ~ 'Neoplasm' ,
    between (as.numeric(histologicTypeIcdO3), 8260, 8265) ~ 'Papillary Adenocarcinoma',
    between (as.numeric(histologicTypeIcdO3), 8050, 8052)~ 'Papillary Carcinoma',
    # NOS and others
    TRUE ~ 'Other/Not Specified'
  ))


  # Add simplified 4-group histology classification
thr_his <- thr_his %>%
  mutate(histology = case_when(
    # NSCLC group
    histology_group %in% c('Adenocarcinoma', 'Squamous Cell Carcinoma', 
                          'Large Cell Carcinoma', 'Non-Small Cell Carcinoma NOS') ~ 'NSCLC',
    
    # SCLC group
    histology_group == 'Small Cell Carcinoma' ~ 'SCLC',
    
    # Other Specified group (known histology types that aren't NSCLC or SCLC)
    histology_group %in% c('Carcinoid Tumor', 'Sarcoma', 'Bronchiolo-alveolar Carcinoma',
    'Acinar Cell Carcinoma', 'Neoplasm' ,'Papillary Adenocarcinoma', 'Papillary Carcinoma',
                           'Mesothelioma') ~ 'Other Specified',
    
    # Other Unspecified group
    TRUE ~ 'Other Unspecified'
  ))

  #table(thr_his$histology, thr_his$histology_group)
  #thr_his=thr_his %>% filter(histology_group!="Other/Not Specified")
  tab_res=thr_his %>% group_by(histology)%>% summarise(total=sum(patient_count))
  tab_res=tab_res %>% mutate(perc=round(100*(total/sum_thr_c),1)) %>%  mutate(count_perc = sprintf("%s (%s%%)", format(total, big.mark = ","), perc))
```




## Thoracic Cancer Histology Distribution
From `r format(thr_den_c$unique_thoracic_cancer_pts[1], big.mark = ",")` thoracic patients with completed cases, the majority `r format(tab_res$total[tab_res$histology=="NSCLC"], big.mark = ",")` were diagnosed with Non-Small Cell Lung Carcinoma (NSCLC). 

::: {style="text-align: center"}
```{r histology output, message=FALSE, warning=FALSE}
labels <- c(
  "histology"="Histology",
  "count_perc" = "N(%) of Patients"
)
create_gt_table_v3(
  data = tab_res, 
  columns = c( "histology", "count_perc"), 
  labels = labels, 
   subtitle_text = "",
  footnote_text = "Note: Patients could have multiple histologies")
```
:::



## Distribution of Thoracic Cancer Cases by Histology Types

**Note:** The following analysis excluded unspecified histology for simplicity and filtered out any groups with less than 20 cases. 

::: {.panel-tabset}

### Bar Plot
The following horizontal bar chart presents the hierarchical distribution of thoracic cancer cases, showing both major categories (NSCLC, SCLC, and Other Specified) and their corresponding histological subtypes. 

- Click legend items to show/hide specific main categories (NSCLC, SCLC, or Other Specified)
- Hover over bars to see detailed information
- Double-click legend items to isolate a single category and  double-click again to restore all categories

```{r histology barplot, message=FALSE, warning=FALSE}
  #
thr_his=thr_his %>% filter (patient_count>20)

summary_plot <- thr_his %>%
  # First group by main category
  group_by(histology) %>%
  mutate(total_in_category = sum(patient_count)) %>%
  group_by(histology, histology_group) %>%
  summarise(
    count = sum(patient_count),
    pct_of_total = round(100 * count / sum_thr_c, 1),
    .groups = 'drop'
  ) %>%
  arrange(desc(count)) %>%
  mutate(
    # Create ordered factors based on count
    histology = factor(histology, levels = unique(histology[order(desc(count))])),
    histology_group = factor(histology_group, levels = unique(histology_group[order(desc(count))])),
    hover_text = sprintf(
      "Category: %s<br>Subtype: %s<br>Count: %d<br>Percent of Total: %.1f%%",
      histology, histology_group, count, pct_of_total
    )
  )

plot_ly(summary_plot,
        x = ~count,
        y = ~histology_group,
        color = ~histology,
        type = "bar",
        orientation = 'h',
        text = ~paste0(count, " (", pct_of_total, "%)"),
        textposition = "auto",
        hovertext = ~hover_text,
        hoverinfo = "text",
        colors = brewer.pal(4, "Set3")
) %>%
  layout(
    title = "",
    xaxis = list(title = "Patient Count"),
    yaxis = list(
      title = "",
      automargin = TRUE,
      categoryorder = "total ascending"  # This will sort the y-axis by total values
    ),
    barmode = "stack",
    showlegend = TRUE,
    legend = list(
      title = list(text = "Histology"),
      x = 0.7,
      y = 0.9
    )
    #margin = list(l = 150, r = 50, t = 50, b = 50),
    #height = 600
  )
```


## Distribution of Thoracic Cancer Cases by Histology Types
The following dual-layered classification system by specific histology and then by broader clinical grouping (NSCLC, SCLC, and Other Specified) represents the thoracic cancer distribution by histology grouping. 

### Sunburst Plot

```{r histology sunburst,message=FALSE, warning=FALSE}
  thr_his=thr_his %>% filter(histology_group!="Other/Not Specified")

# Create hierarchical data for sunburst chart
sunburst_data <- thr_his %>% filter (patient_count>20)%>%
  group_by(histology, histology_group) %>%
  summarize(value = sum(patient_count)) %>%
  ungroup()

# Calculate total for percentages
total_count <- sum_thr_c

# Create the labels and parents vectors for sunburst
labels <- c(unique(sunburst_data$histology), sunburst_data$histology_group)
parents <- c(rep("", length(unique(sunburst_data$histology))), sunburst_data$histology)
values <- c(tapply(sunburst_data$value, sunburst_data$histology, sum), sunburst_data$value)

# Create color palette with enough colors
colors <- colorRampPalette(brewer.pal(8, "Set2"))(length(labels))

# Create sunburst chart
p2 <- plot_ly(
  labels = labels,
  parents = parents,
  values = values,
  type = "sunburst",
  branchvalues = "total",
  hovertemplate = ~paste(
    "<b>%{label}</b><br>",
    "Count: %{value}<br>",
    "<extra></extra>"
  ),
  textposition = "auto",
  insidetextorientation = "horizontal",
  textfont = list(size = 14),
  marker = list(
    colors = colors,
    line = list(color = '#FFFFFF', width = 1)
  )
) %>%
  layout(
    title = "(Click to explore subcategories)",
    width = 600,  # Further increased width for better label spacing
    height = 400, # Further increased height for better label spacing
    sunburstcolorway = colors,
    margin = list(l = 50, r = 50, t = 50, b = 50),  # Added margins for label space
    annotations = list(
      text = "Click segments to explore subcategories",
      x = 0.5,
      y = -0.1,
      showarrow = FALSE,
      xref = 'paper',
      yref = 'paper'
    ),
    showlegend = FALSE
  )

# Display the sunburst chart
p2
```

:::

```{r histology table, message=FALSE, warning=FALSE}
thr_his=thr_his %>% filter (patient_count>20)
library(DT)

# Create summary table
histology_summary <- thr_his %>%
  group_by(histology, histology_group) %>%
  summarize(
    Patient_Count = sum(patient_count),
    Percentage = round(sum(patient_count) / sum(thr_his$patient_count) * 100, 2)
  ) %>%
  arrange(desc(Patient_Count))

# Display interactive table
datatable(histology_summary,
          colnames = c("Histology", "Histology Group", "Patient Count", "Percentage (%)"),
          options = list(
            pageLength = 10,
            dom = 'Bfrtip',
            ordering = TRUE
          ),
          rownames = FALSE) %>%
  formatStyle('Percentage',
             background = styleColorBar(histology_summary$Percentage, "lightgray"),
             backgroundSize = '98% 88%',
             backgroundRepeat = 'no-repeat',
             backgroundPosition = 'center')
```


## Detailed Distribution of Thoracic Cancer Subcategories
```{r piechart with barplots combined, message=FALSE, warning=FALSE}
# Create main summary
thr_his_res=thr_his %>% filter (patient_count>20)
hist_summary <- thr_his_res %>%
  group_by(histology_group) %>%
  summarize(patient_count = sum(patient_count)) %>%
  mutate(percentage = round(100 * patient_count/sum(patient_count), 1)) %>%
  arrange(desc(patient_count))

# Create the main pie chart
p <- plot_ly() %>%
  add_pie(data = hist_summary,
          labels = ~histology_group,
          values = ~patient_count,
          textposition = 'inside',
          textinfo = 'label+percent',
          hoverinfo = 'text',
          text = ~paste(histology_group, 
                       '\nCount:', patient_count,
                       '\nPercentage:', percentage, '%'),
          marker = list(colors = brewer.pal(12, "Set3"))) %>%
  layout(title = "Distribution of Thoracic Cancer Histology Types",
         showlegend = TRUE,
         legend = list(orientation = "h", x = 0.5, xanchor = "center", y = -0.1))

# Create detailed subtype plot (initially hidden)
subtype_data <- thr_his %>% filter (patient_count>20) %>%
  group_by(histology_group, histologicTypeIcdO3Description) %>%
  summarize(patient_count = sum(patient_count)) %>%
  mutate(percentage = round(100 * patient_count/sum(patient_count), 1))

# Add event handling for switching between views
p <- onRender(p, "
  function(el, x) {
    var mainPlot = x.mainPlot;
    var currentView = 'pie';
    
    el.on('plotly_click', function(d) {
      var pt = d.points[0];
      if (currentView === 'pie') {
        var selectedGroup = pt.label;
        var filteredData = x.subgroups[selectedGroup];
        
        if (filteredData) {
          currentView = 'bar';
          Plotly.newPlot(el, [{
            type: 'bar',
            x: filteredData.x,
            y: filteredData.y,
            text: filteredData.text,
            textposition: 'auto',
            marker: {color: x.colors[0]}
          }], {
            title: 'Detailed View: ' + selectedGroup + ' (Refresh the page to return)',
            xaxis: {
              title: '',
              tickangle: ''
            },
            yaxis: {title: 'Patient Count'},
            showlegend: false
          });
        }
      } else {
        // Return to main pie chart view
        currentView = 'pie';
        Plotly.newPlot(el, mainPlot.data, mainPlot.layout);
      }
    });
  }
")

# Prepare subgroup data
subgroups <- lapply(unique(thr_his$histology_group), function(group) {
  data <- subtype_data %>%
    filter(histology_group == group) %>%
    arrange(desc(patient_count))
  
  list(
    x = data$histologicTypeIcdO3Description,
    y = data$patient_count,
    text = paste0(data$patient_count, " (", data$percentage, "%)")
  )
})
names(subgroups) <- unique(thr_his$histology_group)

# Add data to the plot's x parameter
p$x$subgroups <- subgroups
p$x$colors <- brewer.pal(12, "Set3")
p$x$mainPlot <- list(
  data = p$x$data,
  layout = p$x$layout
)

# Display the interactive plot
p
```


```{r icdo code table, message=FALSE, warning=FALSE}
# Create detailed summary table
thr_his_summary <- thr_his_res %>%filter (patient_count>20)%>%
  group_by(histology, histology_group, histologicTypeIcdO3Description) %>%
  summarize(
    patient_count = sum(patient_count),
    histology_codes = paste(sort(unique(histologicTypeIcdO3)), collapse = ", ")
  ) %>%
  mutate(percentage = round(100 * patient_count/sum(patient_count), 1)) %>%
  arrange(desc(patient_count))

tab_his<- thr_his_summary %>% select (histology_group, histologicTypeIcdO3Description, histology_codes, patient_count)


# Display using DT for interactive table
DT::datatable(tab_his,
              colnames = c("Histology","Histology Group", "Original Histology", "ICD-O-3", "Patient Count"),
              options = list(
                pageLength = 10,
                searching = TRUE,
                ordering = TRUE,
                dom = 'f<"clear">lrtip',  #  empty search box
                language = list(
                  search = ""  # This removes the "Search:" label
                )
              ),
              rownames = FALSE) 
```
## Cancer Staging 
Cancer staging is categorized as:

- **Stage 0**: Carcinoma in situ (neoplastic cells confined to the epithelium without invasion) with no evidence of invasion beyond the basement membrane.
- **Occult**: A malignancy is confirmed by cytology or metastatic presentation without a known primary, meaning that the primary tumor is not clinically or radiologically identifiable.
- **Stage I**: A localized tumor confined to the organ of origin, with no regional lymph node involvement or distant metastasis.
- **Stage II**: A tumor of larger size or with limited regional extension that may involve adjacent structures or limited regional lymph nodes.
- **Stage III**: Locally advanced disease characterized by involvement of regional lymph nodes and/or extension to nearby tissues.
- **Stage IV**: Distant metastatic disease where the tumor has spread to distant organs or non-regional lymph nodes.

```{r staging, message=FALSE, warning=FALSE, include=FALSE}
# Create the staging data frame
staging <- data.frame(
  Stage = c("Stage 0", "Occult", "Stage I", "Stage II", "Stage III", "Stage IV"),
  Description = c(
    "Carcinoma in situ; neoplastic cells confined to the epithelium without invasion",
    "Malignancy confirmed by cytology or metastatic presentation without known primary",
    "Localized tumor confined to the organ of origin",
    "Larger tumor size or limited regional extension" ,
    "Locally advanced disease" ,
    "Distant metastatic disease"
  ), 
  `Extent of Disease Spread` =c ("No evidence of invasion beyond the basement membrane.", "Primary tumor not clinically or radiologically identifiable.", "No regional lymph node involvement or distant metastasis.", "May involve adjacent structures or limited regional lymph nodes.", "Involvement of regional lymph nodes and/or extension to nearby tissues.", "Tumor has spread to distant organs or non-regional lymph nodes."),

  stringsAsFactors = FALSE
)

# Display the table interactively with DT and without pagination/searching
DT::datatable(
  staging,
  colnames = c("Stage", "Description",  "Extent of Disease Spread"),
  options = list(
    pageLength = 10,
    searching = FALSE,
    ordering = FALSE,
    dom = 't'
  ),
  rownames = FALSE
)
```
## Summary of Staging of Thoracic Cancer Patients
The following table summarizes the staging status of thoracic patients with completed cases. 
```{r staging counts, message=FALSE, warning=FALSE, results='hide'}
sql_stg <- "/workspaces/starr-oncology-data-lake-arpah/src/sql/scr/thoracic/scr_thoracic_stage_person_metrics.sql"
thr_stg <- fetch_data_from_sql_file(sql_stg, yaml_file_path)
thr_stg

thr_stg = thr_stg %>% mutate(perc=100*(patient_count/sum(patient_count)) ) %>% arrange(desc(patient_count))


thr_stg <- thr_stg %>%
  mutate(
    full_stage = case_when(
      derived_stage == "0" ~ "Stage 0: Carcinoma in situ",
      derived_stage == "Occult"  ~ "Occult: Malignancy confirmed by cytology or metastatic presentation without known primary",
      derived_stage == "1" ~ "Stage I: Localized tumor confined to the organ of origin",
      derived_stage == "2" ~ "Stage II: Larger tumor size or limited regional extension",
      derived_stage == "3"~ "Stage III: Locally advanced disease",
      derived_stage == "4" ~ "Stage IV: Distant metastatic disease",
      TRUE ~ derived_stage
    )
  )
```
```{r staging table, message=FALSE, warning=FALSE}
#thr_stg %>% filter(full_stage="Unknown")
labels <- c(
  "full_stage"="Staging",
  "patient_count" = "Patient Count",
   "perc" = "Percent"
)

create_gt_table_v1(
  data = thr_stg, 
  columns = c( "full_stage","patient_count", "perc"), 
  labels = labels, 
   subtitle_text = "",
  footnote_text = "")
```


### Treatment
The following table summarizes the distribution of thoracic cancer treatment types including chemotherapy, radiation therapy, and surgery across the patient cohort with completed cases. 

```{r, message=FALSE, warning=FALSE, results='hide'}
sql_tx <- "/workspaces/starr-oncology-data-lake-arpah/src/sql/scr/thoracic/scr_thoracic_completed_cases_tx_summary_person_metrics.sql"
thr_tx <- fetch_data_from_sql_file(sql_tx, yaml_file_path)


library(dplyr)
library(tidyr)

thr_tx_long <- thr_tx %>%
  pivot_longer(
    cols = c(chemo_flag, radiation_flag, surgery_flag),
    names_to = "treatment_type",
    values_to = "treatment_flag"
  ) %>%
  group_by(treatment_type, treatment_flag) %>%
  summarize(
    total_patients = sum(patient_count),
    .groups = "drop"
  ) %>%
  group_by(treatment_type) %>%
  mutate(
    percentage = round(100 * total_patients /sum_thr_c, 1),
    treatment_name = case_when(
      treatment_type == "chemo_flag" ~ "Chemotherapy",
      treatment_type == "radiation_flag" ~ "Radiation Therapy",
      treatment_type == "surgery_flag" ~ "Surgery",
      TRUE ~ treatment_type
    )
  ) %>% 
  filter(treatment_flag=="Yes") %>% 
  arrange(desc(percentage))

thr_tx_long$treatment_type<-NULL
```
```{r, message=FALSE, warning=FALSE}

tx_thr=thr_tx_long %>% 
              select(treatment_name, total_patients, percentage)

labels <- c(
  "Description"="Treatment",
  "Patinet_Counts" = "Patient Count",
  "percentage"="Percentage (%)"
)

create_gt_table_v1(
  data = tx_thr, 
  columns = c( "treatment_name", "total_patients", "percentage"), 
  labels = labels, 
   subtitle_text = "",
  footnote_text = "")
```
### Social History 
From  `r format(thr_den$unique_thoracic_cancer_pts, big.mark = ",")` unique thoracic cancer patients, the majority didn't have any records of smoking status and alcohol use in Neural Frame. 

```{r load soh, message=FALSE, warning=FALSE, results='hide'}
sql_soh <- "/workspaces/starr-oncology-data-lake-arpah/src/sql/scr/thoracic/scr_thoracic_social_history.sql"
thr_soh <- fetch_data_from_sql_file(sql_soh, yaml_file_path)

colnames(thr_soh)<-c("Description", "Patinet_Counts", "Status" )

# Split data into smoking and alcohol
smoking_data <- thr_soh %>% filter(Status=="smoking status")
alcohol_data <-  thr_soh %>% filter(Status=="alcohol status")
## den is all thoracic pts 
```


#### Smoking Status (Neural Frame)
```{r smoking, message=FALSE, warning=FALSE}
smoking_data <- smoking_data %>%
  mutate(Description = replace_na(Description, "Not Available"))

smoking_data=smoking_data %>% select (Description , Patinet_Counts)%>% arrange(desc(Patinet_Counts))
labels <- c(
  "Description"="Description",
  "Patinet_Counts" = "Patient Count"
)

create_gt_table_v1(
  data = smoking_data, 
  columns = c( "Description","Patinet_Counts"), 
  labels = labels, 
   subtitle_text = "",
  footnote_text = "Note: sum of sungroups exceeds the denominator due to multiple responses by patients")
```

#### Smoking Status (OMOP)
The information regarding the most recent smoking status of patients with thoracic cancer was generated by querying the OMOP database. For patients diagnosed with lung, bronchus, or thymus cancer, relevant patient records were linked to their smoking status in OMOP observation table by filtering out the smoking concept ID (43054909), which included multiple entries over time. By ranking these entries by date, the analysis focused on retrieving only the most recent smoking status for each patient as follow. 

```{r , message=FALSE, warning=FALSE, results='hide'}
sql_sm_omop <- "/workspaces/starr-oncology-data-lake-arpah/src/sql/scr/thoracic/scr_thoracic_smoking_omop.sql"
thr_smk <- fetch_data_from_sql_file(sql_sm_omop, yaml_file_path)
smk_na= sum_thr - sum(thr_smk$unique_person_count)
```


#### Alcohol Use (Neural Frame)
```{r, message=FALSE, warning=FALSE, results='hide'}
alcohol_data <- alcohol_data %>%
  mutate(Description = replace_na(Description, "Not Available"))

alcohol_data=alcohol_data %>% select (Description , Patinet_Counts)%>% arrange(desc(Patinet_Counts))
labels <- c(
  "Description"="Description",
  "Patinet_Counts" = "Patient Count"
)
```
```{r alcohol,message=FALSE, warning=FALSE }
create_gt_table_v1(
  data = alcohol_data, 
  columns = c( "Description","Patinet_Counts"), 
  labels = labels, 
   subtitle_text = "",
  footnote_text = "Note: sum of sungroups exceeds the denominator due to multiple responses by patients")
```




```{r smoking omop,message=FALSE, warning=FALSE }

smoking_omop=thr_smk %>% select (smoking_status , unique_person_count)%>% arrange(desc(unique_person_count))
labels <- c(
  "smoking_status"="Description",
  "unique_person_count" = "Patient Count"
)

create_gt_table_v1(
  data = smoking_omop, 
  columns = c( "smoking_status" , "unique_person_count"), 
  labels = labels, 
   subtitle_text = "",
  footnote_text = paste0("Note: ", format(smk_na, big.mark = ","), " patients didn't have smoking concept ID in the observation table"))
```
### Follow-up Outcome 
The following table summarizes the distribution of thoracic cancer recurrence status with completed cases. 
```{r load recurrence, message=FALSE, warning=FALSE, include='FALSE' }
library(readr)

recurrence <- read_csv("/workspaces/starr-oncology-data-lake-arpah/src/may_2025/data/recurrenceType_category.csv")

# Preview the data
head(recurrence)
recurrence$patient_count<-NULL
colnames(recurrence)<-c("recurrencetype1st","recurrencetype1stdescription", "recurrence_Type_category")
```
```{r thoracic recurrence, message=FALSE, warning=FALSE, results='hide', include='FALSE'}
sql_rec <- "/workspaces/starr-oncology-data-lake-arpah/src/sql/scr/thoracic/scr_thoracic_recurrence.sql"
thr_rec <- fetch_data_from_sql_file(sql_rec, yaml_file_path)
thr_rec=thr_rec %>% inner_join(recurrence, by="recurrencetype1stdescription")
thr_rec = thr_rec %>% select (-recurrenceType1st) %>% group_by(recurrence_Type_category) %>% summarise(total=sum(n_pts))%>% select(recurrence_Type_category, total)
```
```{r recurrence type, message=FALSE, warning=FALSE }
thr_rec=thr_rec %>% arrange(desc(total))

labels <- c(
  "recurrence_Type_category"="Recurrence Type Category",
  "total" = "Patient Count"
)

create_gt_table_v1(
  data = thr_rec, 
  columns = c( "recurrence_Type_category" , "total"), 
  labels = labels, 
   subtitle_text = "",
  footnote_text ="")

```

:::

:::