---
title: "Neural Frame Metrics"
execute:
  echo: false
---
```{r, message=FALSE, warning=FALSE, results='hide'}
#| context: global

rm(list=ls())
source("/workspaces/starr-oncology-data-lake-arpah/src/R/all_function.R", encoding = "UTF-8")
yaml_file_path <-  "/workspaces/starr-oncology-data-lake-arpah/src/sql_params.yml"
library(RColorBrewer)  # Add RColorBrewer package
```
```{r, message=FALSE, warning=FALSE, results='hide'}
sql_den<- "/workspaces/starr-oncology-data-lake-arpah/src/sql/scr/thoracic/scr_den.sql"
nf_den <- fetch_data_from_sql_file(sql_den, yaml_file_path)
nf_den
```

```{r,message=FALSE, warning=FALSE, results='hide'}
sql_nf<- "/workspaces/starr-oncology-data-lake-arpah/src/sql/scr/thoracic/scr_case_status.sql"
case <- fetch_data_from_sql_file(sql_nf, yaml_file_path)
head(case)
case$nfcasestatus[case$nfcasestatus=="Incomplete - Edit"]<-"Incomplete"
case$flag="All Cancer"
```
### Summary:
The Stanford Cancer Registry data is integrated with the California Cancer Registry and updated monthly, encompassing approximately 800 variables related to various cancers. As of January 2022, the KACI® application from NeuralFrame was approved for abstracting and transmitting diagnosed cases, with Stanford adopting it in 2023. NeuralFrame connects to EPIC to extract relevant data, including pathology notes.

Within Neural Frame, Class of Case (CoC) divides cases into two groups: cancer cases are categorized as **Analytic** (reportable with complete data) or **Non-analytic** (not reportable by law, with complete data only from 2022 onward).

 - **Analytic**: Analytic cases include patients with initial diagnoses and full or partial treatments at Stanford. Analytic cases (codes 00-22) are those that are required by CoC to be abstracted because of the program’s primary responsibility in managing the cancer. Analytic cases are grouped according to the location of diagnosis and treatment. Treatment and outcome reports may be limited to analytic cases. 

 - **Non-analytic**: Non-analytic cases include recurrences, consults only, and in-transit patients. Non-analytic cases (codes 30-49 and 99) may be abstracted by the facility to meet central registry requirements or because of a request by the facility’s cancer program. Non-analytic cases are grouped according to the reason a patient who received care at the facility is Non-analytic, or the reason a patient who never received care at the facility may have been abstracted.

::: {.columns}

::: {.column width="50%"}
### 📊 Data Volume 
- Patients Count:  `r format(nf_den$unique_person_count[1], big.mark = ",")`
- Patients with Complete Cases: `r format(case$unique_person_count_case[case$nfcasestatus=="Completed"], big.mark = ",")`
:::

::: {.column width="50%"}
### 🧬 Data Components 
- Outcome
- Diagnoses 
- Treatment
- Miscellaneous
:::

```{r, message=FALSE, warning=FALSE, results='hide'}
sql_nf<- "/workspaces/starr-oncology-data-lake-arpah/src/sql/scr/thoracic/scr_case_status.sql"
case <- fetch_data_from_sql_file(sql_nf, yaml_file_path)
head(case)
case$nfcasestatus[case$nfcasestatus=="Incomplete - Edit"]<-"Incomplete"
case$flag="All Cancer"

# Calculate counts and percentages using unique person counts
case_summary <- aggregate(unique_person_count_case ~ nfcasestatus, data = case, sum)
case_df <- data.frame(
  Status = case_summary$nfcasestatus,
  Count = case_summary$unique_person_count_case,
  Percentage = round(case_summary$unique_person_count_case / 197000 * 100, 1)
)

# Sort by percentage in descending order
case_df <- case_df[order(-case_df$Percentage), ]
```
```{r, message=FALSE, warning=FALSE, results='hide'}
# Create interactive table using DT
library(DT)
library(ggplot2)

# Display table
#DT::datatable(case_df, 
 #        colnames = c("Case Status", "Unique Person Count", "Percentage (%)"),
    #       options = list(
      #       pageLength = 25,  # increased to show more rows per page
        #     searching = FALSE,
        #     ordering = TRUE,
        #     dom = 'rtip'
        #   ),
        #   rownames = FALSE
 #)
```





## Thoracic Cancer Metrics {.tabset}
This visualization summarizes the characteristics of complete cases for Thoracic Cancer by:

 - Cases and Diagnoses 
 - Treatment
 - Social History

::: {.panel-tabset}

### Cases & Diagnoses
Thoracic cancer patients are identified based on their primary site descriptions in the Neural Frame diagnoses data, which include diagnoses of lung, bronchus, or thymus cancers.
The thoracic cancer cohort includes patients diagnosed with primary malignancies located in the thoracic cavity, as defined by the International Classification of Diseases for Oncology, 3rd Edition (ICD-O-3) primary site codes. Specifically, the cohort is defined by the following ICD-O-3 topography codes:

 - C34.0 – Main bronchus
 - C34.1 – Upper lobe, lung
 - C34.2 – Middle lobe, lung
 - C34.3 – Lower lobe, lung
 - C34.8 – Overlapping lesion of lung
 - C34.9 – Lung, not otherwise specified (NOS)
 - C37.9 – Thymus



## Case Status
```{r, message=FALSE, warning=FALSE, results='hide'}
sql_thr_case<- "/workspaces/starr-oncology-data-lake-arpah/src/sql/scr/thoracic/scr_thoracic_case_status_person_metrics.sql"
thr_case <- fetch_data_from_sql_file(sql_thr_case, yaml_file_path)
head(thr_case)
thr_case$flag="Thoracic Cancer"
thr_case$nfcasestatus[thr_case$nfcasestatus=="Incomplete - Edit"]<-"Incomplete"


sql_thr<-"/workspaces/starr-oncology-data-lake-arpah/src/sql/scr/thoracic/scr_thoracic_denominator.sql"

thr_den=fetch_data_from_sql_file(sql_thr, yaml_file_path)
thr_den
```
```{r,message=FALSE, warning=FALSE, results='hide'}
# Calculate counts and percentages using unique person counts
case_summary_thr <- aggregate(patient_count ~ nfcasestatus, data = thr_case, sum)
case_thr <- data.frame(
  Status = case_summary_thr$nfcasestatus,
  Count = case_summary_thr$patient_count,
  Percentage = round(case_summary_thr$patient_count / (thr_den$unique_thoracic_cancer_pts) * 100)
)

```
```{r, message=FALSE, warning=FALSE, results='hide'}
## thoracic cancer with complete cases 
sql_thr_c<-"/workspaces/starr-oncology-data-lake-arpah/src/sql/scr/thoracic/scr_thoracic_denominator_complete.sql"

thr_den_c=fetch_data_from_sql_file(sql_thr_c, yaml_file_path)
thr_den_c
## denominators ##
sum_thr_c=thr_den_c$unique_thoracic_cancer_pts
sum_thr=thr_den$unique_thoracic_cancer_pts
```
As of May 2025, there are  `r format(thr_den$unique_thoracic_cancer_pts, big.mark = ",")` unique thoracic patients with `r format(case_thr$Percentage[case_thr$Status=="Completed"], big.mark = ",")` % complete cases. 



```{r, message=FALSE, warning=FALSE}
# Create pie chart
ggplot(case_thr, aes(x = "", y = Percentage, fill = Status)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  theme_minimal() +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.border = element_blank(),
    panel.grid = element_blank(),
    axis.ticks = element_blank(),
    axis.text.x = element_blank()
  ) +
  geom_text(aes(label = paste0(Status, "\n", Percentage, "%")), 
            position = position_stack(vjust = 0.5)) +
  ggtitle("Case Status Distribution") +
  scale_fill_brewer(palette = "Set3")
```

## Patient Summary by Diagnosis Year 
```{r, message=FALSE, warning=FALSE, results='hide'}
sql_thr_yr<- "/workspaces/starr-oncology-data-lake-arpah/src/sql/scr/thoracic/scr_thoracic_completed_cases_dx_year_person_metrics.sql"
thr_yr <- fetch_data_from_sql_file(sql_thr_yr, yaml_file_path)

# Group years before 2000 and calculate percentages
thr_yr_grouped <- thr_yr %>%
  mutate(dx_year = as.character(ifelse(dx_year < 2000, "<2000", dx_year))) %>%
  group_by(dx_year) %>%
  summarize(patient_count = sum(patient_count)) %>%
  mutate(percentage = round(100 * patient_count / sum(patient_count), 1),
         label = paste0(percentage, "%")) %>%
  arrange(dx_year)
```
```{r, message=FALSE, warning=FALSE}
library(plotly)

# Create an interactive bar plot using plotly
p <- plot_ly(data = thr_yr_grouped,
             x = ~dx_year,
             y = ~patient_count,
             type = "bar",
             text = ~label,
             textposition = "auto",
             marker = list(color ="#FB8072" ))  

# Add layout settings
p <- p %>% 
  layout(
    title = list(text = "Number of Thoracic Cancer Patients by Diagnosis Year"),
    xaxis = list(
      title = "Diagnosis Year",
      tickangle = 45
    ),
    yaxis = list(
      title = "Patient Count"
    ),
    showlegend = FALSE
  )

# Display the plot
p
```

## Histology Distribution & ICD-O-3-Codes
Thoracic cancer encompasses a diverse group of histologic subtypes, with Non-Small Cell Lung Cancer **(NSCLC)** representing the majority of cases and Small Cell Lung Cancer **(SCLC)**.

The Thoracic Cancer cohort includes patients diagnosed with primary malignancies located in the thoracic cavity, as defined by the International Classification of Diseases for Oncology, 3rd Edition (ICD-O-3) primary site codes. 

Thoracic cancers are categorized into detailed histologic groups using ICD-O-3 codes and then further grouped into four overarching categories for simplified analysis. The main histology groups include:

- **Non-Small Cell Lung Carcinoma (NSCLC):** This group encompasses the majority of lung cancers and includes subtypes such as:
  - Adenocarcinoma (8140–8147)
  - Squamous Cell Carcinoma (8070–8080)
  - Large Cell Carcinoma (8010–8015)
  - Non-Small Cell Carcinoma NOS (8046)

- **Small Cell Lung Carcinoma (SCLC):** Represented by ICD-O-3 codes 8041–8045, this group is characterized by more aggressive, rapidly progressing tumors.

- **Other Specified Histologies:** These include a range of less common thoracic cancers such as:
  - Carcinoid Tumors (8240–8249)
  - Sarcoma (8800–8806)
  - Mesothelioma (9050–9053)
  - Acinar Cell Carcinoma (8550–8551)
  - Bronchiolo-alveolar Carcinoma (8250–8257)
  - Neoplasms (8000–8005)
  - Papillary Adenocarcinoma (8260–8265)
  - Papillary Carcinoma (8050–8052)

- **Other/Unspecified Histologies:** Any histology not falling into the categories above.

Note: The following analysis excluded unspecified histology for simplicity and filtered out any groups with less than 20 cases. 


```{r, message=FALSE, warning=FALSE, results='hide'}
# Load required libraries
library(plotly)
library(dplyr)
library(RColorBrewer)
library(htmlwidgets)

sql_thr_his <- "/workspaces/starr-oncology-data-lake-arpah/src/sql/scr/thoracic/scr_thoracic_completed_cases_histology_person_metrics.sql"
thr_his <- fetch_data_from_sql_file(sql_thr_his, yaml_file_path)

# Add histology grouping
thr_his <- thr_his %>%
  mutate(histology_group = case_when(
    # Non-Small Cell Lung Cancer (NSCLC) Groups
    between(as.numeric(histologicTypeIcdO3), 8140, 8147) ~ 'Adenocarcinoma',
    between(as.numeric(histologicTypeIcdO3), 8070, 8080) ~ 'Squamous Cell Carcinoma',
    between(as.numeric(histologicTypeIcdO3), 8012, 8014) ~ 'Large Cell Carcinoma',
    histologicTypeIcdO3 == '8046' ~ 'Non-Small Cell Carcinoma NOS',
    
    # Small Cell Lung Cancer (SCLC)
    between(as.numeric(histologicTypeIcdO3), 8041, 8045) ~ 'Small Cell Carcinoma',
    
    # Neuroendocrine Tumors
    between(as.numeric(histologicTypeIcdO3), 8240, 8249) ~ 'Carcinoid Tumor',
    
    # Other specified types
    between(as.numeric(histologicTypeIcdO3), 8800, 8806) ~ 'Sarcoma',
    between(as.numeric(histologicTypeIcdO3), 9050, 9053) ~ 'Mesothelioma',
    between(as.numeric(histologicTypeIcdO3), 8550, 8551) ~ 'Acinar Cell Carcinoma',
    between(as.numeric(histologicTypeIcdO3), 8250, 8257) ~ 'Bronchiolo-alveolar Carcinoma',
    between(as.numeric(histologicTypeIcdO3), 8000, 8005) ~ 'Neoplasm' ,
    between (as.numeric(histologicTypeIcdO3), 8260, 8265) ~ 'Papillary Adenocarcinoma',
    between (as.numeric(histologicTypeIcdO3), 8050, 8052)~ 'Papillary Carcinoma',
    # NOS and others
    TRUE ~ 'Other/Not Specified'
  ))


  # Add simplified 4-group histology classification
thr_his <- thr_his %>%
  mutate(histology = case_when(
    # NSCLC group
    histology_group %in% c('Adenocarcinoma', 'Squamous Cell Carcinoma', 
                          'Large Cell Carcinoma', 'Non-Small Cell Carcinoma NOS') ~ 'NSCLC',
    
    # SCLC group
    histology_group == 'Small Cell Carcinoma' ~ 'SCLC',
    
    # Other Specified group (known histology types that aren't NSCLC or SCLC)
    histology_group %in% c('Carcinoid Tumor', 'Sarcoma', 'Bronchiolo-alveolar Carcinoma',
    'Acinar Cell Carcinoma', 'Neoplasm' ,'Papillary Adenocarcinoma', 'Papillary Carcinoma',
                           'Mesothelioma') ~ 'Other Specified',
    
    # Other Unspecified group
    TRUE ~ 'Other Unspecified'
  ))

  #table(thr_his$histology, thr_his$histology_group)
  thr_his=thr_his %>% filter(histology_group!="Other/Not Specified")
  tab_res=thr_his %>% filter (patient_count>20)%>% group_by(histology)%>% summarise(total=sum(patient_count))
  tab_res=tab_res %>% mutate(perc=round(100*(total/sum_thr_c),1)) %>%  mutate(count_perc = sprintf("%s (%s%%)", format(total, big.mark = ","), perc))
```




## Thoracic Cancer Histology Distribution
From `r format(thr_den_c$unique_thoracic_cancer_pts[1], big.mark = ",")` thoracic patients with complete cases, the majority `r format(tab_res$total[tab_res$histology=="NSCLC"], big.mark = ",")` were diagnosed with Non-Small Cell Lung Carcinoma (NSCLC). 

::: {style="text-align: center"}
```{r, message=FALSE, warning=FALSE}
labels <- c(
  "histology"="Histology",
  "count_perc" = "N(%) of Patients"
)
create_gt_table_v3(
  data = tab_res, 
  columns = c( "histology", "count_perc"), 
  labels = labels, 
   subtitle_text = "",
  footnote_text = "")
```
:::



The following horizontal bar chart presents the hierarchical distribution of thoracic cancer cases, showing both major categories (NSCLC, SCLC, and Other Specified) and their corresponding histological subtypes. Each bar represents patient counts and percentages, with color-coding to distinguish between major types.


This interactive horizontal bar chart shows the distribution of thoracic cancer types and their subtypes. To interact with the visualization:

- Click legend items to show/hide specific main categories (NSCLC, SCLC, or Other Specified)
- Hover over bars to see detailed information
- Double-click legend items to isolate a single category and  double-click again to restore all categories


```{r, message=FALSE, warning=FALSE}
  #
thr_his=thr_his %>% filter (patient_count>20)

summary_plot <- thr_his %>%
  # First group by main category
  group_by(histology) %>%
  mutate(total_in_category = sum(patient_count)) %>%
  group_by(histology, histology_group) %>%
  summarise(
    count = sum(patient_count),
    pct_of_total = round(100 * count / sum_thr_c, 1),
    .groups = 'drop'
  ) %>%
  arrange(desc(count)) %>%
  mutate(
    # Create ordered factors based on count
    histology = factor(histology, levels = unique(histology[order(desc(count))])),
    histology_group = factor(histology_group, levels = unique(histology_group[order(desc(count))])),
    hover_text = sprintf(
      "Category: %s<br>Subtype: %s<br>Count: %d<br>Percent of Total: %.1f%%",
      histology, histology_group, count, pct_of_total
    )
  )

plot_ly(summary_plot,
        x = ~count,
        y = ~histology_group,
        color = ~histology,
        type = "bar",
        orientation = 'h',
        text = ~paste0(count, " (", pct_of_total, "%)"),
        textposition = "auto",
        hovertext = ~hover_text,
        hoverinfo = "text",
        colors = brewer.pal(4, "Set3")
) %>%
  layout(
    title = "Distribution of Thoracic Cancer Histology Types",
    xaxis = list(title = "Patient Count"),
    yaxis = list(
      title = "",
      automargin = TRUE,
      categoryorder = "total ascending"  # This will sort the y-axis by total values
    ),
    barmode = "stack",
    showlegend = TRUE,
    legend = list(
      title = list(text = "Histology"),
      x = 0.7,
      y = 0.9
    )
    #margin = list(l = 150, r = 50, t = 50, b = 50),
    #height = 600
  )
```


## Distribution of Thoracic Cancer by Histology Types
The following dual-layered classification system by specific histology and then by broader clinical grouping (NSCLC, SCLC, and Other Specified) represents the thoracic cancer distribution by histology grouping. 


```{r,message=FALSE, warning=FALSE}
  thr_his=thr_his %>% filter(histology_group!="Other/Not Specified")

# Create hierarchical data for sunburst chart
sunburst_data <- thr_his %>% filter (patient_count>20)%>%
  group_by(histology, histology_group) %>%
  summarize(value = sum(patient_count)) %>%
  ungroup()

# Calculate total for percentages
total_count <- sum_thr_c

# Create the labels and parents vectors for sunburst
labels <- c(unique(sunburst_data$histology), sunburst_data$histology_group)
parents <- c(rep("", length(unique(sunburst_data$histology))), sunburst_data$histology)
values <- c(tapply(sunburst_data$value, sunburst_data$histology, sum), sunburst_data$value)

# Create color palette with enough colors
colors <- colorRampPalette(brewer.pal(8, "Set2"))(length(labels))

# Create sunburst chart
p2 <- plot_ly(
  labels = labels,
  parents = parents,
  values = values,
  type = "sunburst",
  branchvalues = "total",
  hovertemplate = ~paste(
    "<b>%{label}</b><br>",
    "Count: %{value}<br>",
    "Percentage: %{percentParent:.1%}<br>",
    "<extra></extra>"
  ),
  textposition = "auto",
  insidetextorientation = "horizontal",
  textfont = list(size = 14),
  marker = list(
    colors = colors,
    line = list(color = '#FFFFFF', width = 1)
  )
) %>%
  layout(
    title = "(Click to explore subcategories)",
    width = 600,  # Further increased width for better label spacing
    height = 400, # Further increased height for better label spacing
    sunburstcolorway = colors,
    margin = list(l = 50, r = 50, t = 50, b = 50),  # Added margins for label space
    annotations = list(
      text = "Click segments to explore subcategories",
      x = 0.5,
      y = -0.1,
      showarrow = FALSE,
      xref = 'paper',
      yref = 'paper'
    ),
    showlegend = FALSE
  )

# Display the sunburst chart
p2
```
```{r, message=FALSE, warning=FALSE}
thr_his=thr_his %>% filter (patient_count>20)

# Create summary table
histology_summary <- thr_his %>%
  group_by(histology, histology_group) %>%
  summarize(
    Patient_Count = sum(patient_count),
    Percentage = round(sum(patient_count) / sum(thr_his$patient_count) * 100, 2)
  ) %>%
  arrange(desc(Patient_Count))

# Display interactive table
datatable(histology_summary,
          colnames = c("Histology", "Histology Group", "Patient Count", "Percentage (%)"),
          options = list(
            pageLength = 10,
            dom = 'Bfrtip',
            ordering = TRUE
          ),
          rownames = FALSE) %>%
  formatStyle('Percentage',
             background = styleColorBar(histology_summary$Percentage, '#FB8072'),
             backgroundSize = '98% 88%',
             backgroundRepeat = 'no-repeat',
             backgroundPosition = 'center')
```


## Detailed Distribution of Thoracic Cancer Subcategories
```{r, message=FALSE, warning=FALSE}
# Create main summary
thr_his_res=thr_his %>% filter (patient_count>20)
hist_summary <- thr_his_res %>%
  group_by(histology_group) %>%
  summarize(patient_count = sum(patient_count)) %>%
  mutate(percentage = round(100 * patient_count/sum(patient_count), 1)) %>%
  arrange(desc(patient_count))

# Create the main pie chart
p <- plot_ly() %>%
  add_pie(data = hist_summary,
          labels = ~histology_group,
          values = ~patient_count,
          textposition = 'inside',
          textinfo = 'label+percent',
          hoverinfo = 'text',
          text = ~paste(histology_group, 
                       '\nCount:', patient_count,
                       '\nPercentage:', percentage, '%'),
          marker = list(colors = brewer.pal(12, "Set3"))) %>%
  layout(title = "Distribution of Thoracic Cancer Histology Types",
         showlegend = TRUE,
         legend = list(orientation = "h", x = 0.5, xanchor = "center", y = -0.1))

# Create detailed subtype plot (initially hidden)
subtype_data <- thr_his %>% filter (patient_count>20) %>%
  group_by(histology_group, histologicTypeIcdO3Description) %>%
  summarize(patient_count = sum(patient_count)) %>%
  mutate(percentage = round(100 * patient_count/sum(patient_count), 1))

# Add event handling for switching between views
p <- onRender(p, "
  function(el, x) {
    var mainPlot = x.mainPlot;
    var currentView = 'pie';
    
    el.on('plotly_click', function(d) {
      var pt = d.points[0];
      if (currentView === 'pie') {
        var selectedGroup = pt.label;
        var filteredData = x.subgroups[selectedGroup];
        
        if (filteredData) {
          currentView = 'bar';
          Plotly.newPlot(el, [{
            type: 'bar',
            x: filteredData.x,
            y: filteredData.y,
            text: filteredData.text,
            textposition: 'auto',
            marker: {color: x.colors[0]}
          }], {
            title: 'Detailed View: ' + selectedGroup + ' (Refresh the page to return)',
            xaxis: {
              title: '',
              tickangle: 45
            },
            yaxis: {title: 'Patient Count'},
            showlegend: false
          });
        }
      } else {
        // Return to main pie chart view
        currentView = 'pie';
        Plotly.newPlot(el, mainPlot.data, mainPlot.layout);
      }
    });
  }
")

# Prepare subgroup data
subgroups <- lapply(unique(thr_his$histology_group), function(group) {
  data <- subtype_data %>%
    filter(histology_group == group) %>%
    arrange(desc(patient_count))
  
  list(
    x = data$histologicTypeIcdO3Description,
    y = data$patient_count,
    text = paste0(data$patient_count, " (", data$percentage, "%)")
  )
})
names(subgroups) <- unique(thr_his$histology_group)

# Add data to the plot's x parameter
p$x$subgroups <- subgroups
p$x$colors <- brewer.pal(12, "Set3")
p$x$mainPlot <- list(
  data = p$x$data,
  layout = p$x$layout
)

# Display the interactive plot
p
```


```{r, message=FALSE, warning=FALSE}
# Create detailed summary table
thr_his_summary <- thr_his_res %>%filter (patient_count>20)%>%
  group_by(histology, histology_group, histologicTypeIcdO3Description) %>%
  summarize(
    patient_count = sum(patient_count),
    histology_codes = paste(sort(unique(histologicTypeIcdO3)), collapse = ", ")
  ) %>%
  mutate(percentage = round(100 * patient_count/sum(patient_count), 1)) %>%
  arrange(desc(patient_count))

tab_his<- thr_his_summary %>% select (histology_group, histologicTypeIcdO3Description, histology_codes, patient_count)


# Display using DT for interactive table
DT::datatable(tab_his,
              colnames = c("Histology","Histology Group", "Original Histology", "ICD-O-3", "Patient Count"),
              options = list(
                pageLength = 10,
                searching = TRUE,
                ordering = TRUE,
                dom = 'f<"clear">lrtip',  #  empty search box
                language = list(
                  search = ""  # This removes the "Search:" label
                )
              ),
              rownames = FALSE) 
```

### Treatment
The following table summarizes the distribution of thoracic cancer treatments including chemotherapy, radiation therapy, and surgery across the patient cohort with completed cases. 

```{r, message=FALSE, warning=FALSE, results='hide'}
sql_tx <- "/workspaces/starr-oncology-data-lake-arpah/src/sql/scr/thoracic/scr_thoracic_completed_cases_tx_summary_person_metrics.sql"
thr_tx <- fetch_data_from_sql_file(sql_tx, yaml_file_path)


library(dplyr)
library(tidyr)

thr_tx_long <- thr_tx %>%
  pivot_longer(
    cols = c(chemo_flag, radiation_flag, surgery_flag),
    names_to = "treatment_type",
    values_to = "treatment_flag"
  ) %>%
  group_by(treatment_type, treatment_flag) %>%
  summarize(
    total_patients = sum(patient_count),
    .groups = "drop"
  ) %>%
  group_by(treatment_type) %>%
  mutate(
    percentage = round(100 * total_patients /sum_thr_c, 1),
    treatment_name = case_when(
      treatment_type == "chemo_flag" ~ "Chemotherapy",
      treatment_type == "radiation_flag" ~ "Radiation Therapy",
      treatment_type == "surgery_flag" ~ "Surgery",
      TRUE ~ treatment_type
    )
  ) %>% 
  filter(treatment_flag=="Yes") %>% 
  arrange(desc(percentage))

thr_tx_long$treatment_type<-NULL
```
```{r, message=FALSE, warning=FALSE}
# Display treatment data in an interactive table
DT::datatable(thr_tx_long %>% 
              select(treatment_name, total_patients, percentage),
              colnames = c("Treatment", "Patient Count", "Percentage (%)"),
              options = list(
                pageLength = 10,
                searching = FALSE,
                ordering = TRUE,
                dom = 'rtip'
              ),
              rownames = FALSE) 
```
### Social History 
From  `r format(thr_den$unique_thoracic_cancer_pts, big.mark = ",")` unique thoracic cancer patients, the majority didn't have any records of smoking status and alcohol use in Neural Frame. 

```{r, message=FALSE, warning=FALSE, results='hide'}
sql_soh <- "/workspaces/starr-oncology-data-lake-arpah/src/sql/scr/thoracic/scr_thoracic_social_history.sql"
thr_soh <- fetch_data_from_sql_file(sql_soh, yaml_file_path)

colnames(thr_soh)<-c("Description", "Patinet_Counts", "Status" )

# Split data into smoking and alcohol
smoking_data <- thr_soh %>% filter(Status=="smoking status")
alcohol_data <-  thr_soh %>% filter(Status=="alcohol status")
## den is all thoracic pts 
```


#### Smoking Status (Neural Frame)
```{r, message=FALSE, warning=FALSE}
smoking_data <- smoking_data %>%
  mutate(Description = replace_na(Description, "Not Available"))

smoking_data=smoking_data %>% select (Description , Patinet_Counts)%>% arrange(desc(Patinet_Counts))
labels <- c(
  "Description"="Description",
  "Patinet_Counts" = "Number of Patients"
)

create_gt_table_v1(
  data = smoking_data, 
  columns = c( "Description","Patinet_Counts"), 
  labels = labels, 
   subtitle_text = "",
  footnote_text = "Note: sum of sungroups exceeds the denominator due to multiple responses by patients")
```


#### Alcohol Use (Neural Frame)
```{r, message=FALSE, warning=FALSE, results='hide'}
alcohol_data <- alcohol_data %>%
  mutate(Description = replace_na(Description, "Not Available"))

alcohol_data=alcohol_data %>% select (Description , Patinet_Counts)%>% arrange(desc(Patinet_Counts))
labels <- c(
  "Description"="Description",
  "Patinet_Counts" = "Number of Patients"
)
```
```{r,message=FALSE, warning=FALSE }
create_gt_table_v1(
  data = alcohol_data, 
  columns = c( "Description","Patinet_Counts"), 
  labels = labels, 
   subtitle_text = "",
  footnote_text = "Note: sum of sungroups exceeds the denominator due to multiple responses by patients")
```

#### Smoking Status (OMOP)
The information regarding the most recent smoking status of patients with thoracic cancer was generated by querying the OMOP database. For patinets diagnosed with lung, bronchus, or thymus cancers, relevant patient records were linked to their smoking status in OMOP observation table by filtering out the smoking concept ID (43054909), which included multiple entries over time. By ranking these entries by date, the analysis focused on retrieving only the most recent smoking status for each patient as follow. 

```{r, message=FALSE, warning=FALSE, results='hide'}
sql_sm_omop <- "/workspaces/starr-oncology-data-lake-arpah/src/sql/scr/thoracic/scr_thoracic_smoking_omop.sql"
thr_smk <- fetch_data_from_sql_file(sql_sm_omop, yaml_file_path)
smk_na= sum_thr - sum(thr_smk$unique_person_count)
```
```{r,message=FALSE, warning=FALSE }

smoking_omop=thr_smk %>% select (smoking_status , unique_person_count)%>% arrange(desc(unique_person_count))
labels <- c(
  "smoking_status"="Description",
  "unique_person_count" = "Number of Patients"
)

create_gt_table_v1(
  data = smoking_omop, 
  columns = c( "smoking_status" , "unique_person_count"), 
  labels = labels, 
   subtitle_text = "",
  footnote_text = paste0("Note: ", format(smk_na, big.mark = ","), " patients didn't have smoking concept ID in the observation table"))
```
### Survival Outcome 
:::

:::
