---
title: "Philips ISPM Metrics"
execute:
  echo: false
---
```{r, message=FALSE, warning=FALSE, results='hide'}
#| context: global

rm(list=ls())
source("/workspaces/starr-oncology-data-lake-arpah/src/R/all_function.R", encoding = "UTF-8")
yaml_file_path <-  "/workspaces/starr-oncology-data-lake-arpah/src/sql_params.yml"
library(RColorBrewer)  
library(plotly)
```

```{r, message=FALSE, warning=FALSE, results='hide'}
philips_den_sql<- "/workspaces/starr-oncology-data-lake-arpah/src/sql/philips/philips_denominator.sql"
ph_den <- fetch_data_from_sql_file(philips_den_sql, yaml_file_path)
ph_den$counts_pts

## aberration denominators 
abr_den_sql<- "/workspaces/starr-oncology-data-lake-arpah/src/sql/philips/philips_aberration_denominator.sql"
abr_den <- fetch_data_from_sql_file(abr_den_sql, yaml_file_path)
abr_den
```
```{r, message=FALSE, warning=FALSE, results='hide'}
test_sql="/workspaces/starr-oncology-data-lake-arpah/src/sql/philips/philips_tests_ordered.sql"
test_type=fetch_data_from_sql_file(test_sql, yaml_file_path)

test_date_sql="/workspaces/starr-oncology-data-lake-arpah/src/sql/philips/philips_tests_date_ordered.sql"
test_date=fetch_data_from_sql_file(test_date_sql, yaml_file_path)

test_type=test_type %>% inner_join(test_date, by="test_type")
test_type=test_type %>% mutate(test_type=ifelse(test_type=="StanfordStampTest", "Stanford Stamp Test", "Foundation One Ckb Test"))

```

### Summary:

Philips IntelliSpace Precision Medicine (ISPM) includes genomics database at Stanford.
This data is geared towards identifying genetic mutations that can be targeted for precision medicine, with the goal of improving therapeutic outcomes through personalized treatment strategies.
As of May 2025, `r format(sum(test_type$test_type_counts), big.mark=",") STAMP tests were ordered for `r format(ph_den$counts_pts[1], big.mark = ",")` patients. 

```{r, message=FALSE, warning=FALSE}
labels <- c(
  "test_type"="Type of Tests",
  "test_type_counts" = "N (Tests)",
   "min_order_date" = "Min (Date)",
   "max_order_date"= "Max (Date)"
)

create_gt_table_v1(
  data = test_type, 
  columns = c( "test_type","test_type_counts", "min_order_date", "max_order_date"), 
  labels = labels, 
   subtitle_text = "",
  footnote_text = paste0("Total number of ordered tests: ", format(sum(test_type$test_type_counts)
, big.mark = ",") ))
```

::: {.columns}

::: {.column width="50%"}
### 🧬 Data Components 
- Stanford Stamp Test
- Aberration 
- Specimen 
:::


::: {.column width="50%"}
### 📊 Data Volume 
- Patients Count: `r format(ph_den$counts_pts[1], big.mark = ",")`
- Patients with Aberration Cases: `r format(abr_den$n_pts, big.mark = ",")`
:::


## Tables: {.tabset}

::: {.panel-tabset}

### Diagnoses Orders
## Site Display and Site Code:
 These fields specifies the specific locations in the genome where mutations are detected. The site_display and site_code being disease or organ-specific suggests that the data is organized to reflect the relevance of mutations to particular conditions or anatomical sites, which is crucial for targeted treatment strategies.

```{r,message=FALSE, warning=FALSE, results='hide' }
site_sql<-"/workspaces/starr-oncology-data-lake-arpah/src/sql/philips/philips_site_display.sql"
site <- fetch_data_from_sql_file(site_sql, yaml_file_path)

```

```{r, message=FALSE, warning=FALSE}
# Calculate percentages using numeric values
site <- site %>% filter (n_pts>30) %>%
  mutate(percentage =round( (n_pts/as.numeric(ph_den$counts_pts)) * 100))

# Create interactive horizontal bar plot with plotly
plot_ly(site, 
        y = ~reorder(site_display, n_pts), # switched to y and reordered
        x = ~n_pts, # switched to x
        type = "bar",
        orientation = 'h', # make bars horizontal
        marker = list(color = brewer.pal(12, "Set3"))) %>%
  layout(
    title = "Distribution of Genomic Sites in ISPM Database",
    yaxis = list( # switched axis properties
      title = "",
      tickangle = 0 # removed angle since it's now on y-axis
    ),
    xaxis = list(title = "Patient Count"), # switched axis properties
    hoverlabel = list(bgcolor = "white"),
    hovertemplate = paste(
      "<b>%{y}</b><br>", # switched to %{y}
      "Percentage: %{x:.1f}%<br>", # switched to %{x}
      "<extra></extra>"
    ),
    margin = list(l = 150) # added left margin for site labels
  )
```

### Aberrations
- Aberration Types: An aberration type describes a specific change to DNA.

- Aberration Classes: An aberration class groups related types of genomic changes into broader categories.

- Chromosomes (chrom): This field refers to chromosomal location. For example if a gene is located on chromosome 17, it will be presented at "chr17".

- HGNC Gene and Detected Mutations (hgnc_gene): The HGNC (HUGO Gene Nomenclature Committee) provides standardized names for human genes. Detected mutations refer to specific changes found in these genes that may be important for specific therapies, such as gene therapy or personalized medicine, to improve patient care.

Chromosome number tells us where in the genome we are looking, HGNC gene names tell us which gene or genes are involved, and aberration type/class tells us what kind of genetic change occurred.

```{r,message=FALSE, warning=FALSE , results='hide'}
gene_ch_sql<-"/workspaces/starr-oncology-data-lake-arpah/src/sql/philips/philips_chrom_genes.sql"
gene_ch <- fetch_data_from_sql_file(gene_sql, yaml_file_path)

# Calculate percentages using the denominator
gene_ch <- gene_ch %>%
  mutate(pts_percent = round((n_pts/as.numeric(ph_den$counts_pts))*100, 1))
```

```{r,message=FALSE, warning=FALSE , results='hide'}
gene_sql<-"/workspaces/starr-oncology-data-lake-arpah/src/sql/philips/philips_gene.sql"
gene<- fetch_data_from_sql_file(gene_sql, yaml_file_path)

# Calculate percentages using the denominator
gene <- gene %>%
  mutate(pts_percent = round((n_pts/as.numeric(abr_den$n_pts))*100, 1))
```

```{r,message=FALSE, warning=FALSE }
gene= gene_ch %>% group_by(hgnc_gene) %>% mutate (total_pts=sum(n_pts))
# Create interactive bar plot for top genes
gene_plot <- gene %>%
  filter(n_pts > 500) %>%
  arrange(desc(n_pts)) %>%
  plot_ly(
    x = ~n_pts,
    y = ~reorder(hgnc_gene, n_pts),
    type = "bar",
    marker = list(color = brewer.pal(15, "Set3")),
    text = ~paste("<br>N:", n_pts, "<br>", pts_percent, "%"),
    hoverinfo = "text"
  ) %>%
  layout(
    title = "Distribution of HGNC Genes in Patient Population",
    yaxis = list(title = "Gene"),
    xaxis = list(title = "Number of Patients"),
    margin = list(b = 100)
  )

gene_plot
```

```{r, message=FALSE, warning=FALSE }
# Update the datatable to include percentage
DT::datatable(
  gene %>% filter (n)
    select(hgnc_gene, n_pts, pts_percent) %>%
    rename(
      "Gene" = hgnc_gene,
      "Patient Count" = n_pts,
      "Percentage" = pts_percent
    ),
  options = list(
    pageLength = 10,
    order = list(list(1, 'desc')),
    dom = 'ftlip'
  ),
  rownames = FALSE
) %>%
  DT::formatStyle(
    'Patient Count',
    background = DT::styleColorBar(gene$n_pts, '#FB8072'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  ) %>%
  DT::formatRound('Percentage', digits = 1)
```

## Genes Distribution by Chromosomes
```{r, message=FALSE, warning=FALSE }
library(dplyr)
library(ggplot2)

gene %>%
  count(chrom, hgnc_gene, sort = TRUE) %>%
  group_by(chrom) %>%
  top_n(10, n) %>%
  ggplot(aes(x = reorder(hgnc_gene, n), y = n, fill = chrom)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  facet_wrap(~ chrom, scales = "free") +
  labs(x = "HGNC Gene", y = "Count", title = "Top Genes by Chromosomes") +
  theme_minimal()

```
```{r,message=FALSE, warning=FALSE}
# Create heatmap
plot_ly(
  data = gene,
  x = ~chrom,
  y = ~hgnc_gene,
  z = ~n_pts,
  type = "heatmap",
  colorscale = "Viridis",
  hovertemplate = paste(
    "<b>Gene:</b> %{y}<br>",
    "<b>Chromosome:</b> %{x}<br>",
    "<b>Patients:</b> %{z}<extra></extra>"
  )
) %>%
  layout(
    title = "Patient Counts by Gene and Chromosome",
    xaxis = list(title = "Chromosome"),
    yaxis = list(title = "Gene"),
    margin = list(l = 100)  # To fit long gene names
  )
```




```{r, message=FALSE, warning=FALSE, results='hide'}
mu_type_sql<-"/workspaces/starr-oncology-data-lake-arpah/src/sql/philips/philips_abr_type.sql"
mutation_data <- fetch_data_from_sql_file(mu_type_sql, yaml_file_path)
mutation_data=mutation_data %>% filter(n_pts>200) 
```

```{r, message=FALSE, warning=FALSE}
gene_order <- mutation_data %>%
  group_by(hgnc_gene) %>%
  summarise(total = sum(n_pts), .groups = "drop") %>%
  arrange(desc(total)) %>%
  pull(hgnc_gene)

mutation_data$hgnc_gene <- factor(mutation_data$hgnc_gene, levels = gene_order)
 
 # mutate(hgnc_gene = fct_reorder(hgnc_gene, total_patients, .desc = TRUE))

# Plotly stacked bar plot
plot_ly(
  mutation_data,
  y = ~hgnc_gene,
  x = ~n_pts,
  color = ~aberration_type,
  colors = "Set3",
  type = 'bar'
) %>%
  layout(
    barmode = 'stack',
    title = "Mutation Type Distribution Across Genes",
    yaxis = list(title = "Gene"),
    xaxis = list(title = "Number of Patients")
  )
```
```{r, message=FALSE, warning=FALSE}

df_bar <- mutation_data %>% filter(n_pts>200)%>%
  arrange(desc(n_pts)) %>% mutate(aberration_type = forcats::fct_reorder(aberration_type, n_pts))

plot_ly(df_bar,
        x = ~hgnc_gene,
       y= ~n_pts,
        type = "bar",
        color = ~aberration_type,
        colors = "Set3") %>%
  layout( barmode = 'stack',
    xaxis = list(title = "Gene"),
         yaxis = list(title = "Person Count")) %>%layout (title = "Mutation Type Distribution Across Genes")
```

:::
