---
title: "Philips ISPM"
execute:
  echo: false
---
```{r, message=FALSE, warning=FALSE, results='hide'}
#| context: global

rm(list=ls())
source("/workspaces/starr-oncology-data-lake-arpah/src/R/all_function.R", encoding = "UTF-8")
yaml_file_path <-  "/workspaces/starr-oncology-data-lake-arpah/src/sql_params.yml"
library(RColorBrewer)  
library(plotly)
```

```{r, message=FALSE, warning=FALSE, results='hide'}
philips_den_sql<- "/workspaces/starr-oncology-data-lake-arpah/src/sql/philips/philips_denominator.sql"
ph_den <- fetch_data_from_sql_file(philips_den_sql, yaml_file_path)
ph_den$counts_pts

## aberration denominators 
abr_den_sql<- "/workspaces/starr-oncology-data-lake-arpah/src/sql/philips/philips_aberration_denominator.sql"
abr_den <- fetch_data_from_sql_file(abr_den_sql, yaml_file_path)
abr_den
```
```{r, message=FALSE, warning=FALSE, results='hide'}
test_sql="/workspaces/starr-oncology-data-lake-arpah/src/sql/philips/philips_tests_ordered.sql"
test_type=fetch_data_from_sql_file(test_sql, yaml_file_path)

test_date_sql="/workspaces/starr-oncology-data-lake-arpah/src/sql/philips/philips_tests_date_ordered.sql"
test_date=fetch_data_from_sql_file(test_date_sql, yaml_file_path)

test_type=test_type %>% inner_join(test_date, by="test_type")
test_type=test_type %>% mutate(test_type=ifelse(test_type=="StanfordStampTest", "Stanford Stamp Test", "Foundation One Ckb Test"))%>%
  mutate(
    min_order_ym = format(min_order_date, "%Y-%m"),
    max_order_ym = format(max_order_date, "%Y-%m")
  )



```

### Summary:

Philips IntelliSpace Precision Medicine (ISPM) includes genomics database at Stanford.
This data is geared towards identifying genetic mutations that can be targeted for precision medicine, with the goal of improving therapeutic outcomes through personalized treatment strategies.
As of May 2025, `r format(sum(test_type$test_type_counts), big.mark=",")` STAMP tests were ordered for `r format(ph_den$counts_pts[1], big.mark = ",")` patients. 

```{r, message=FALSE, warning=FALSE}
labels <- c(
  "test_type"="Test Types",
  "test_type_counts" = "n_tests",
   "min_order_ym" = "min_date",
   "max_order_ym"= "max_date"
)

create_gt_table_v1(
  data = test_type, 
  columns = c( "test_type","test_type_counts", "min_order_ym", "max_order_ym"), 
  labels = labels, 
   subtitle_text = "",
  footnote_text = paste0("Total number of ordered tests: ", format(sum(test_type$test_type_counts)
, big.mark = ",") ))
```

::: {.columns}

::: {.column width="50%"}
### 🧬 Data Components 
- Diagnoses orders: includes patients who had the test ordered
- Aberration: records if a genetic aberration is found in the STAMP panel tests
:::


::: {.column width="50%"}
### 📊 Data Volume 
- Patients Count: `r format(ph_den$counts_pts[1], big.mark = ",")`
- Patients with Aberration Cases: `r format(abr_den$n_pts, big.mark = ",")`
:::


## Aberration Metrics: {.tabset}
The aberration table provides a structured overview of genetic changes, categorizing them into specific types and broader classes. Each entry in the table includes the following key components:

- **Aberration Types:** An aberration type describes a specific change to DNA. These are specific alterations in DNA that can affect gene function and contribute to various diseases. Each type represents a distinct change, such as missense variant, deletions, duplications, or point mutations.

- **Aberration Classes:** This field groups related aberration types into overarching categories, allowing for a more organized understanding of the types of genomic changes and their potential implications in health and disease.

- **Chromosomes:** This field indicates the chromosomal location of the gene associated with the aberration. For example if a gene is located on chromosome 17, it will be presented at "chr17".

- **HGNC Gene and Detected Mutations:** The HGNC (HUGO Gene Nomenclature Committee) provides standardized names for human genes. Detected mutations refer to specific changes found in these genes that may be important for specific therapies, such as gene therapy or personalized medicine, to improve patient care.

::: {.panel-tabset}


```{r,message=FALSE, warning=FALSE, results='hide' }
site_sql<-"/workspaces/starr-oncology-data-lake-arpah/src/sql/philips/philips_site_display.sql"
site <- fetch_data_from_sql_file(site_sql, yaml_file_path)

```

```{r, message=FALSE, warning=FALSE, results='hide'}
# Calculate percentages using numeric values
site <- site %>% filter (n_pts>30) %>%
  mutate(percentage =round( (n_pts/as.numeric(ph_den$counts_pts)) * 100))

# Create interactive horizontal bar plot with plotly
plot_ly(site, 
        y = ~reorder(site_display, n_pts), # switched to y and reordered
        x = ~n_pts, # switched to x
        type = "bar",
        orientation = 'h', # make bars horizontal
        marker = list(color = brewer.pal(12, "Set3"))) %>%
  layout(
    title = "Distribution of Sites in ISPM Database",
    yaxis = list( # switched axis properties
      title = "",
      tickangle = 0 # removed angle since it's now on y-axis
    ),
    xaxis = list(title = "Patient Count"), # switched axis properties
    hoverlabel = list(bgcolor = "white"),
    hovertemplate = paste(
      "<b>%{y}</b><br>", # switched to %{y}
      "Percentage: %{x:.1f}%<br>", # switched to %{x}
      "<extra></extra>"
    ),
    margin = list(l = 150) # added left margin for site labels
  )
```

### All Cancer Types
The following analysis displays genes that have aberration for more than 100 patients. 

```{r,message=FALSE, warning=FALSE , results='hide'}
gene_ch_sql<-"/workspaces/starr-oncology-data-lake-arpah/src/sql/philips/philips_chrom_genes.sql"
gene_ch <- fetch_data_from_sql_file(gene_ch_sql, yaml_file_path)

# Calculate percentages using the denominator
gene_ch <- gene_ch %>%
  mutate(pts_percent = round((n_pts/as.numeric(ph_den$counts_pts))*100, 1))
```

```{r,message=FALSE, warning=FALSE , results='hide'}
gene_sql<-"/workspaces/starr-oncology-data-lake-arpah/src/sql/philips/philips_gene.sql"
gene<- fetch_data_from_sql_file(gene_sql, yaml_file_path)

# Calculate percentages using the denominator
gene <- gene %>%
  mutate(pts_percent = round((n_pts/as.numeric(abr_den$n_pts))*100, 1))
```


## HGNC Gene Distribution

```{r,message=FALSE, warning=FALSE}
## explain the top genes 
## table filtering count>10
df_bar <- gene %>%
  filter(n_pts > 500) %>%
  arrange(desc(n_pts)) %>% mutate(hgnc_gene = forcats::fct_reorder(hgnc_gene, n_pts))

plot_ly(df_bar,
        y = ~reorder(hgnc_gene, n_pts),
        x = ~n_pts,
        type = "bar",
        color = ~hgnc_gene,
        colors = "Set3") %>%
  layout(xaxis = list(title = "Person Count"),
         yaxis = list(title = "")) %>%layout (title = "HGNC Gene Distribution")

```

 ## Summary Table of HGNC Gene Distribution
```{r, message=FALSE, warning=FALSE }
# Update the datatable to include percentage
DT::datatable(
  gene %>% 
    select(hgnc_gene, n_pts, pts_percent) %>%
    rename(
      "Gene" = hgnc_gene,
      "Patient Count" = n_pts,
      "Percentage" = pts_percent
    ),
  options = list(
    pageLength = 10,
    order = list(list(1, 'desc')),
    dom = 'ftlip'
  ),
  rownames = FALSE
) %>%
  DT::formatStyle(
    'Patient Count',
    background = DT::styleColorBar(gene$n_pts, '#FB8072'),
    backgroundSize = '98% 88%',
    backgroundRepeat = 'no-repeat',
    backgroundPosition = 'center'
  ) %>%
  DT::formatRound('Percentage', digits = 1)
```

## Genes Distribution by Chromosomes 

```{r, message=FALSE, warning=FALSE}
  filtered_gene_ch <- gene_ch %>%
  filter(!is.na(hgnc_gene), n_pts > 100) %>%
  mutate(
    # Create proper chromosome ordering
    chrom = factor(chrom, 
                  levels = paste0("chr", c(1:22, "X", "Y")),
                  ordered = TRUE)
  ) %>%
  # Sort by chromosome and patient count
  arrange(chrom, desc(n_pts))


  p <- ggplot(filtered_gene_ch, 
     aes(x = 1, 
         y = reorder(hgnc_gene, n_pts), 
         fill = n_pts,
         text = sprintf(
           "Gene: %s<br>Patient Count: %d",
           hgnc_gene, n_pts
         ))) +
  geom_tile(aes(width = n_pts/max(n_pts), 
                height = 0.8)) +
  facet_wrap(~chrom, scales = "free_y", ncol = 5) +
  scale_fill_distiller(name = "Patient Count", 
                      palette = "Set3", 
                      direction = 1) +
  theme_minimal() +
  labs(title = "Patient Counts by Gene and Chromosome",
       y = "Gene",
       x = NULL) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid = element_blank(),
    strip.text = element_text(angle = 0),
    legend.position = "right",
    panel.spacing = unit(0.3, "cm")
  )

# Make it interactive with custom hover
ggplotly(p, 
         tooltip = "text", # Use only the custom text for hover
         height = 1000, 
         width = 800) %>%
  layout(
    margin = list(l = 150, r = 50, b = 50, t = 50),
    showlegend = TRUE
  )
```




## Gene Distribution across Chromosomes 

```{r, message=FALSE, warning=FALSE}
  p <- ggplot(filtered_gene_ch, 
     aes(x = n_pts, 
         y = chrom,
         text = sprintf(
           "Gene: %s<br>Patient Count: %d<br>Chromosome: %s",
           hgnc_gene, n_pts, chrom
         ))) +
         geom_text(
    data = . %>% filter(n_pts > 500),  # Filter for text labels
    aes(label = hgnc_gene),
    size = 3,
    hjust = -0.2,
    check_overlap = TRUE
  ) +
  geom_jitter(
    aes(size = n_pts, color = n_pts),
    alpha = 0.6,
    width = 0.3,      # No horizontal jitter
    height = 0   # Vertical jitter amount
  ) +
  scale_size_continuous(range = c(3, 10)) +
  scale_color_distiller(name = "Patient Count", 
    palette = "Set3",
    direction = 1
  ) +
  theme_minimal() +
  labs(
    title = "Patient Counts by Gene and Chromosome",
    x = "Patient Count",
    y = "Chromosome"
  ) +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank(),
    panel.spacing = unit(0.3, "cm")
  )

# Make it interactive
ggplotly(p, 
         tooltip = "text"
)
         #height = 800, 
        # width = 1200) 
```



## Distribution of Genes by Aberration Type 

```{r, message=FALSE, warning=FALSE, results='hide'}
mu_type_sql<-"/workspaces/starr-oncology-data-lake-arpah/src/sql/philips/philips_abr_type.sql"
mutation_data <- fetch_data_from_sql_file(mu_type_sql, yaml_file_path)
mutation_data=mutation_data %>% filter(n_pts>100) 
```

```{r, message=FALSE, warning=FALSE}
gene_order <- mutation_data %>%
  group_by(aberration_type) %>%
 mutate(total = sum(n_pts))
```


```{r, message=FALSE, warning=FALSE, results='hide'}
gene_order <- mutation_data %>%
  group_by(hgnc_gene) %>%
  mutate(total_pts = sum(n_pts)) %>%
  ungroup() %>%
  filter(n_pts > 100) %>%
  arrange((total_pts)) %>%
  mutate(
    # Order genes by total patient count
    hgnc_gene = factor(hgnc_gene, levels = unique(hgnc_gene)),
    # Order aberration types by their total counts
    aberration_type = forcats::fct_reorder(aberration_type, n_pts, sum)
  )

plot_ly(
  gene_order,
  y = ~hgnc_gene,  # No need for fct_reorder here since we pre-ordered
  x = ~n_pts,
  type = "bar",
  color = ~aberration_type,
  colors = "Set3"
) %>%
  layout(
    barmode = 'stack',
    yaxis = list(
      title = "Gene",
      automargin = TRUE,  # Ensure labels aren't cut off
      tickfont = list(size = 10)
    ),
    xaxis = list(
      title = "Patient Count",
      tickfont = list(size = 10)
    ),
    title = "Mutation Type Distribution Across Genes",
    legend = list(
      orientation = "h", 
      y = -0.2, 
      x = 0.5, 
      xanchor = "center"
    ),
    height = 1000,  # Increased height for better readability
    margin = list(l = 150, r = 50, t = 50, b = 100)  # Adjusted margins
  )
```
 
```{r, message=FALSE, warning=FALSE}
gene_order %>% 
  plot_ly(
    x = ~aberration_type,
    y = ~hgnc_gene,
    z = ~n_pts,
    type = "heatmap",
    colors = colorRamp(brewer.pal(9, "Set3")),
    hovertemplate = paste(
      "<b>Gene:</b> %{y}<br>",
      "<b>Mutation:</b> %{x}<br>",
      "<b>Count:</b> %{z}<br>",
      "<extra></extra>"
    )
  ) %>%
  layout(
    title = "Mutation Type Distribution Across Genes",
    xaxis = list(title = "Mutation Type"),
    yaxis = list(title = "Gene"),
    height = 800,
    margin = list(l = 150, r = 50, t = 50, b = 100)
  )
```

### Thoracic Cancer 
:::
