---
title: "Data Evolution Overview"
toc: true
toc-depth: 3
toc-location: left
execute:
  echo: false
---

This page provides an overview of the VISTA Oncology Data Lake's evolution across multiple releases, showcasing the progressive expansion and maturation of our comprehensive cancer research platform. The data lake has grown substantially from February 2025 through November 2025, with significant increases in patient coverage, data source integration, and analytical capabilities.

## Release Overview

The VISTA Oncology Data Lake represents a multi-institutional effort to create a comprehensive, longitudinal cancer research resource. Through systematic data releases, we have built an increasingly robust platform that integrates clinical, imaging, genomic, and registry data to support advanced cancer research and outcomes analysis.

**Key Growth Highlights:**

- **Patient Cohort Expansion**: From approximately 200,000 patients in February 2025 to over xxx patients by November 2025

- **Enhanced Data Integration**: Progressive addition of NeuralFrame registry data, image occurrences, and genomic testing results 

- **Improved Clinical Coverage**: Expanding availability of structured clinical data, pathology reports, and tumor board documentation

- **Specialized Sub-Cohorts**: Development of thoracic cancer and other disease-specific research cohorts

### February 2025 - Foundation Release

The inaugural February 2025 release established the core OMOP5.3-CDM foundation of our data lake, providing:

- **Core OMOP Dataset**: Comprehensive patient demographics, encounters, diagnoses, and procedures following OMOP CDM 5.3.1 standards
- **Clinical Infrastructure**: Essential clinical data including medications (RxNorm), laboratory results (LOINC), and procedure codes (CPT)
- **NeuralFrame Diagnoses**: Initial integration of Stanford Cancer Registry diagnosis data
- **Quality Framework**: Initial data validation and PHI scrubbing processes to ensure research readiness
- **Patient Base**: Approximately 200,000+ oncology patients forming the foundational cohort

This release prioritized data quality and standardization, establishing the structural foundation for subsequent data integration efforts.

### May 2025 - Integration Expansion

The May 2025 release marked a significant evolution with the integration of multiple specialized data sources:

- **NeuralFrame Integration**: Addition of Stanford Cancer Registry data, providing comprehensive cancer case information including miscellaneous, treatments, and outcomes
- **Imaging Data**: Integration of OMOP image occurrence data, linking patients to their diagnostic imaging studies
- **Philips ISPM**: Introduction of molecular testing data including STAMP, Heme-STAMP, and FoundationOne results
- **Enhanced Coverage**: Substantial increase in patient coverage across multiple data domains

This release transformed the data lake from a primarily clinical dataset to a multi-modal research platform supporting translational cancer research.

### August 2025 - Depth and Specialization

The August 2025 release focused on increasing data depth and developing specialized research capabilities:

- **STAMP Add-On Dataset**: Integrated detailed genomic testing metadata including assay types (STAMP, Heme-STAMP), pipeline versions, and test identifiers for comprehensive molecular profiling analysis
- **HIPPO Benchmark**: Added 1,394 expert-annotated clinical notes corpus for PHI identification and de-identification methodology validation
- **Lung Resection CAP Forms**: Introduced fully identified structured pathology reports combining standardized College of American Pathologists (CAP) templates with unstructured clinical information
- **Epic Media Server Documents**: Integrated all PDF documents from third-party providers accessible via EPIC Media/Lab tabs and Mulesoft API
- **Enhanced Metadata**: Expanded technical documentation including DICOM dictionary, STAMP gene panels, and genomic BED files

This release established the data lake as a comprehensive resource for both population-level and detailed clinical research, while adding critical benchmarking datasets for data quality and PHI scrubbing validation.

### November 2025 - Advanced Analytics and Standardization

The November 2025 release represents a major technical advancement with comprehensive OMOP upgrade and enhanced data capabilities:

- **OMOP 5.3 to 5.4 Upgrade**: Migrated to OMOP CDM version 5.4.2 with new tables, enhanced field linking capabilities, and updated naming conventions for improved data interoperability
- **Vocabulary Modernization**: Updated from January 2023 to August 2025 OHDSI vocabulary, incorporating 2.5 years of concept mappings, domain changes, and standardization improvements
- **Pipeline Redesign**: Overhauled ETL processes with standardized filtering (events from 2000+, flexible visit requirements) and enhanced JSON-formatted source value fields for improved data traceability
- **Genomic Tests Expansion**: Additional molecular testing data and genomic profiling results
- **Agentic Tide Integration**: Advanced analytics capabilities for automated data processing and analysis

This release modernizes the technical infrastructure while maintaining backward compatibility, positioning the data lake for next-generation research applications and enhanced analytical capabilities.

## Multi-Release Statistical Overview

For each data release (February, May, August, and November 2025), we examine:

- Patient counts across key datasets (ARPAH cohort, NeuralFrame, Philips ISPM, Imaging data)
- Data availability metrics for primary domains (encounters, diagnoses, procedures, observations)
- Coverage percentages showing the proportion of patients with data in each system


```{r setup, message=FALSE, warning=FALSE, results='hide'}
# Load required libraries
library(dplyr)
library(yaml)
library(purrr)

# Source the functions
source("/workspaces/starr-oncology-data-lake-arpah/src/R/all_function.R", encoding = "UTF-8")

# Define YAML configuration path
yaml_file_path <- "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql_params.yml"
```

```{r loop through release,  message=FALSE, warning=FALSE, results='hide'}
# Load config from YAML
config <- yaml::read_yaml(yaml_file_path)

# Extract release names (those with nested structure, not top-level scalars)
all_names <- names(config)
releases <- all_names[sapply(config[all_names], is.list)]

# Build datasets_by_month from YAML dynamically
datasets_by_month <- config[releases]

# Map release names to month abbreviations for plotting
release_to_month <- list(
  feb_2025 = "feb",
  may_2025 = "may",
  aug_2025 = "aug",
  nov_2025 = "nov"
)

```

```{r function by time,  message=FALSE, warning=FALSE, results='hide'}
run_monthly_metrics <- function(datasets_by_month,
                                sql_file_path = NULL,
                                sql_file_paths = NULL,
                                release_to_month = NULL,
                                credentials_path = "/home/rstudio/.config/gcloud/application_default_credentials.json") {
  
  # Validate inputs: either sql_file_path OR sql_file_paths must be provided
  if (is.null(sql_file_path) && is.null(sql_file_paths)) {
    stop("❌ Must provide either sql_file_path (single file) or sql_file_paths (named list)")
  }
  
  if (!is.null(sql_file_path) && !is.null(sql_file_paths)) {
    stop("❌ Provide only ONE of: sql_file_path or sql_file_paths, not both")
  }
  
  Sys.setenv(GOOGLE_APPLICATION_CREDENTIALS = credentials_path)
  
  purrr::map_dfr(names(datasets_by_month), function(release_name) {
    
    dataset_map <- datasets_by_month[[release_name]]
    
    # Handle both list and atomic vector structures from YAML
    if (is.list(dataset_map)) {
      project <- dataset_map[["oncology_prod"]]
    } else {
      stop(glue::glue("❌ Unexpected structure for {release_name} in config"))
    }
    
    # Determine which SQL file to use for this release
    if (!is.null(sql_file_paths)) {
      # Release-specific SQL paths provided
      current_sql_path <- sql_file_paths[[release_name]]
      
      if (is.null(current_sql_path)) {
        message(glue::glue("⚠️ No SQL path defined for {release_name}, skipping"))
        return(NULL)
      }
      
      if (!file.exists(current_sql_path)) {
        message(glue::glue("⚠️ SQL file not found for {release_name}: {current_sql_path}"))
        return(NULL)
      }
    } else {
      # Single SQL file for all releases
      current_sql_path <- sql_file_path
    }
    
    # Read and prepare SQL
    sql_query <- readLines(current_sql_path) %>% paste(collapse = "\n")
    
    # Replace ALL placeholders (including oncology_prod project)
    for (placeholder in names(dataset_map)) {
      sql_query <- stringr::str_replace_all(
        sql_query,
        paste0("@", placeholder),
        dataset_map[[placeholder]]
      )
    }
    
    # Connect to BigQuery using project from config
    conn <- DBI::dbConnect(
      bigrquery::bigquery(),
      project = project,
      use_legacy_sql = FALSE
    )
    
    # Run query safely
    result <- tryCatch({
      DBI::dbGetQuery(conn, sql_query)
    }, error = function(e) {
      message(glue::glue("⚠️ Query failed for {release_name}: {e$message}"))
      NULL
    })
    
    DBI::dbDisconnect(conn)
    
    if (!is.null(result)) {
      # Convert release name to month abbreviation for backward compatibility
      month_abbr <- release_to_month[[release_name]] %||% substr(release_name, 1, 3)
      result <- dplyr::mutate(result, month = month_abbr, release = release_name)
    }
    
    result
  })
}

```

```{r list of sqls, message=FALSE, warning=FALSE, results='hide'}
# Define different SQL paths per release
person_sql_paths <- list(
  feb_2025 = "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql_std/arpah_cohort/feb25/person_metrics.sql",
  may_2025 = "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql_std/arpah_cohort/may_aug25/person_metrics.sql",
  aug_2025 = "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql_std/arpah_cohort/may_aug25/person_metrics.sql",
  nov_2025 = "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql_std/arpah_cohort/nov25/person_metrics.sql"
)

person_res <- run_monthly_metrics(
  datasets_by_month = datasets_by_month,
  sql_file_paths = person_sql_paths  # Note: plural!
)
```


```{r person cnt, message=FALSE, warning=FALSE, results='hide'}
person_res <- run_monthly_metrics(
  datasets_by_month = datasets_by_month,
  sql_file_path = "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql/arpah_cohort/person_metrics.sql"
)
colnames(person_res)<-c("counts", "release")

```


### Demographic Summary ###
The demographic composition of the cohort has remained relatively stable across releases.
```{r demographic , message=FALSE, warning=FALSE }

demog <- run_monthly_metrics(
  datasets_by_month = datasets_by_month,
  sql_file_path = "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql_std/arpah_cohort/may_aug25/demographic_summary.sql"
)
```
```{r demog table, message=FALSE, warning=FALSE}
library(DT)

# Format the demographic data for display
demog_formatted <- demog %>%
  mutate(
    month_label = recode(month,
      "feb" = "February",
      "may" = "May", 
      "aug" = "August",
      "nov" = "November"
    ),
    # Format numbers with commas
    sample_size = scales::comma(sample_size),
    percent_female = paste0(round(percent_female, 1), "%"),
    average_age = round(average_age, 1),
    most_frequent_race_percent = paste0(round(most_frequent_race_percent, 1), "%")
  ) %>%
  select(
    month_label,
    sample_size,
    percent_female,
    average_age,
    most_frequent_race,
    most_frequent_race_percent
  )

# Create interactive table
datatable(
  demog_formatted,
  colnames = c(
    "Release",
    "Sample Size", 
    "% Female",
    "Average Age",
    "Most Frequent Race",
    "Race %"
  ),
  caption = "Demographic Summary Across Data Releases",
  options = list(
    dom = 't',  # Only show table (no search, pagination, etc.)
    pageLength = 10,
    columnDefs = list(
      list(className = 'dt-center', targets = c(1, 2, 3, 5)),  # Center align numeric columns
      list(className = 'dt-left', targets = c(0, 4))  # Left align text columns
    )
  ),
  rownames = FALSE
)
```

::: {.panel-tabset}

### Data Sources Summary 

```{r overlap metrics, message=FALSE, warning=FALSE, results='hide'}
# Use denominators queries which return aggregated counts
overlap_sql_paths <- list(
  feb_2025 = "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql_std/arpah_cohort/feb25/arpah_denominators.sql",
  may_2025 = "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql_std/arpah_cohort/may_aug25/arpah_denominators.sql",
  aug_2025 = "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql_std/arpah_cohort/may_aug25/arpah_denominators.sql",
  nov_2025 = "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql_std/arpah_cohort/nov25/arpah_denominators_update.sql"
)

overlap_res <- run_monthly_metrics(
  datasets_by_month = datasets_by_month,
  sql_file_paths = overlap_sql_paths,
  release_to_month = release_to_month
)

# Preview the data structure
head(overlap_res)
```

```{r check overlap data, message=FALSE, warning=FALSE, results='hide'}
# Check what data sources are available
unique(overlap_res$data_set)
```

```{r overlap plot, message=FALSE, warning=FALSE}
library(plotly)
library(dplyr)

# Prepare data with proper month ordering and labels
overlap_plot_data <- overlap_res %>%
  mutate(
    month = factor(month, levels = c("feb", "may", "aug", "nov"), ordered = TRUE),
    month_label = recode(month,
      "feb" = "February",
      "may" = "May",
      "aug" = "August",
      "nov" = "November"
    ),
    # Clean up data set names for legend
    data_set_clean = case_when(
      data_set == "Oncology OMOP (Cohort)" ~ "OMOP Cohort (100%)",
      data_set == "Neural Frame" ~ "Neural Frame",
      data_set == "With Tumor Board Encounter" ~ "Tumor Board",
      data_set == "With Image Occurrence" ~ "Image Occurrence",
      data_set == "Philips ISPM" ~ "Philips ISPM",
      TRUE ~ data_set
    )
  ) %>%
  # Remove the 100% baseline to focus on actual data sources
  filter(data_set != "Oncology OMOP (Cohort)") %>%
  # Ensure we have data for all months, fill missing values
  tidyr::complete(month, data_set_clean, fill = list(percentage = 0, counts_pts = 0))

# Create interactive line plot with consistent colors
plot_ly(
  data = overlap_plot_data,
  x = ~month,
  y = ~percentage,
  color = ~data_set_clean,
  colors = c('#8dd3c7', '#ffffb3', '#bebada', '#fb8072'),
  type = 'scatter',
  mode = 'lines+markers',
  text = ~paste0(
    data_set_clean, 
    "<br>Month: ", month_label,
    "<br>Coverage: ", round(percentage, 1), "%",
    "<br>Patients: ", scales::comma(counts_pts)
  ),
  hoverinfo = "text",
  marker = list(size = 10),
  line = list(width = 3)
) %>%
  layout(
    title = "Data Source Coverage Across Releases",
    xaxis = list(
      title = "Release Month",
      ticktext = c("February", "May", "August", "November"),
      tickvals = c("feb", "may", "aug", "nov"),
      categoryorder = "array",
      categoryarray = c("feb", "may", "aug", "nov")
    ),
    yaxis = list(
      title = "Patient Coverage (%)",
      range = c(0, 105)
    ),
    legend = list(
      title = list(text = "Data Source"),
      orientation = "v",
      x = 1.02,
      y = 0.5
    ),
    hovermode = 'closest',
    margin = list(r = 150)
  )
```

### Data Source Rankings

```{r bump_chart, message=FALSE, warning=FALSE, fig.width=10, fig.height=6}
library(ggplot2)
library(ggbump)
library(dplyr)

# Prepare data for bump chart - calculate rankings by patient count
bump_data <- overlap_res %>%
  filter(data_set != "Oncology OMOP (Cohort)") %>%
  mutate(
    month = factor(month, levels = c("feb", "may", "aug", "nov")),
    month_label = recode(month,
      "feb" = "Feb",
      "may" = "May",
      "aug" = "Aug",
      "nov" = "Nov"
    ),
    data_set_clean = case_when(
      data_set == "Neural Frame" ~ "NeuralFrame",
      data_set == "With Tumor Board Encounter" ~ "Tumor Board",
      data_set == "With Image Occurrence" ~ "Imaging",
      data_set == "Philips ISPM" ~ "Philips ISPM",
      TRUE ~ data_set
    )
  ) %>%
  # Only include data sources that have data in all months OR handle missing properly
  filter(counts_pts > 0) %>%
  group_by(month) %>%
  mutate(rank = rank(-counts_pts, ties.method = "first")) %>%
  ungroup()

# Create bump chart
p <- ggplot(bump_data, aes(x = month_label, y = rank, color = data_set_clean, group = data_set_clean)) +
  geom_bump(linewidth = 2, smooth = 8) +
  geom_point(size = 5) +
  geom_text(
    data = bump_data %>% filter(month == "feb"),
    aes(label = data_set_clean),
    hjust = 1.1,
    size = 3.5,
    fontface = "bold"
  ) +
  geom_text(
    data = bump_data %>% filter(month == "nov"),
    aes(label = data_set_clean),
    hjust = -0.1,
    size = 3.5,
    fontface = "bold"
  ) +
  scale_y_reverse(breaks = 1:max(bump_data$rank)) +
  scale_color_manual(values = c('#8dd3c7', '#ffffb3', '#bebada', '#fb8072')) +
  labs(
    title = "Data Source Coverage Rankings Across Releases",
    subtitle = "Ranked by number of patients (higher rank = more patients)",
    x = "Release",
    y = "Rank"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10, color = "gray40"),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 11, face = "bold")
  ) +
  expand_limits(x = c(0.5, 4.5))

print(p)
```



### OMOP Clinical Metrics 

```{r ctsa, message=FALSE, warning=FALSE} 


ctsa_res <- run_monthly_metrics(
  datasets_by_month = datasets_by_month,
  sql_file_path = "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql_std/omop/ctsa/metrics_uniq_pts_vital_overtime.sql",
  release_to_month = release_to_month)

# Named vector of variable label mappings
library(dplyr)

ctsa_res <- ctsa_res %>%
  mutate(group_var = case_when(
    f0_ == "total_patients"              ~ "Total patients in cohort",
    f0_ == "total_pt_gt_12"              ~ "Patients aged ≥ 12 years",
    f0_ == "uniq_pt_any_insurance_value" ~ "Patients with any recorded insurance",
    f0_ == "uniq_pt_cpt"                 ~ "Patients with CPT (procedure) codes",
    f0_ == "uniq_pt_icd_dx"              ~ "Patients with ICD diagnosis codes",
    f0_ == "uniq_pt_icd_proc"            ~ "Patients with ICD procedure codes",
    f0_ == "uniq_pt_loinc"               ~ "Patients with LOINC lab results",
    f0_ == "uniq_pt_med_rxnorm"          ~ "Patients with medications (RxNorm)",
    f0_ == "uniq_pt_note"                ~ "Patients with clinical notes",
    f0_ == "uniq_pt_opioid"              ~ "Patients with opioid prescriptions",
    f0_ == "uniq_pt_smoking"             ~ "Patients with documented smoking status",
    f0_ == "uniq_pt_snomed_dx"           ~ "Patients with SNOMED diagnosis codes",
    f0_ == "uniq_pt_snomed_proc"         ~ "Patients with SNOMED procedure codes",
    f0_ == "uniq_pt_with_age"            ~ "Patients with valid age recorded",
    TRUE ~ f0_  # fallback: keep original if not matched
  ))

```

```{r ctsa plot, message=FALSE, warning=FALSE }

# Check if ctsa_res has data
if (nrow(ctsa_res) == 0) {
  print("No CTSA data available")
} else {
  
  # Prepare data with proper factor levels
  ctsa_res <- ctsa_res %>% 
    mutate(month = factor(month, levels = c("feb", "may", "aug", "nov"))) %>%
    filter(!f0_ %in% c("total_pt_gt_12", "uniq_pt_with_age")) %>%
    filter(!group_var %in% c("Patients with opioid prescriptions", "Patients with any recorded insurance", "Patients with documented smoking status"))

  library(ggplot2)
  library(plotly)
  
  # Define consistent colors for the plot
  bar_colors <- c('#8dd3c7', '#ffffb3', '#bebada', '#fb8072')

  # Add formatted release labels with proper ordering
  ctsa_res <- ctsa_res %>%
    mutate(release_label = factor(
      recode(release,
        "feb_2025" = "February",
        "may_2025" = "May", 
        "aug_2025" = "August",
        "nov_2025" = "November"
      ),
      levels = c("February", "May", "August", "November")
    ))

  # Create the plot
  plot_ly(
    data = ctsa_res,
    x = ~f1_,
    y = ~reorder(group_var, f1_),
    color = ~release_label,
    colors = bar_colors,
    type = 'bar',
    orientation = 'h',
    text = ~paste0(group_var, "<br>Release: ", release_label, "<br>Count: ", scales::comma(f1_)),
    hoverinfo = "text"
  ) %>%
    layout(
      title = "Clinical Metrics over Releases",
      xaxis = list(title = "Number of Patients", tickformat = ","),
      yaxis = list(title = ""),
      legend = list(
        traceorder = "normal"
      )
    )
}
```

:::

## **Tumor Board and Thoracic Sub-Cohorts** ##

The VISTA Oncology Data Lake includes specialized sub-cohorts for tumor board patients and thoracic cancer cases, enabling focused research on multidisciplinary care coordination and disease-specific outcomes. These curated cohorts demonstrate the platform's capability to support both population-level analyses and targeted clinical investigations.

**Tumor Board Sub-Cohort**: Patients presented at multidisciplinary tumor boards represent complex cases requiring expert consensus for treatment planning. The data lake captures these encounters and tracks the availability of complementary data sources (imaging, genomics, registry data) for comprehensive case analysis.

**Thoracic Cancer Sub-Cohort**: Focused on lung, bronchus, and thymus malignancies, this subset enables specialized research in thoracic oncology. The cohort tracks patients identified through the Stanford Cancer Registry (NeuralFrame), supporting longitudinal studies of treatment patterns and outcomes in thoracic malignancies.

**Multi-Modal Data Integration**: By examining the overlap between tumor board patients and available data modalities (NeuralFrame registry, imaging studies, genomic testing), we can assess the completeness of clinical documentation for this high-complexity patient population and identify opportunities for enhanced data capture.

::: {.panel-tabset}
### Thoracic Cancer Sub-Cohort

```{r thoracic metrics, message=FALSE, warning=FALSE, results='hide'}
# Define thoracic-specific SQL paths per release
thoracic_sql_paths <- list(
  feb_2025 = "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql_std/thoracic/feb25/nf_thoracic/scr_thoracic_denominator.sql",
  may_2025 = "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql_std/thoracic/may_aug25/nf_thoracic/scr_thoracic_denominator.sql", 
  aug_2025 = "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql_std/thoracic/may_aug25/nf_thoracic/scr_thoracic_denominator.sql",
  nov_2025 = "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql_std/thoracic/nov25/nf_thoracic/scr_thoracic_denominator.sql"
)

# Run thoracic analysis across releases
thoracic_res <- run_monthly_metrics(
  datasets_by_month = datasets_by_month,
  sql_file_paths = thoracic_sql_paths,
  release_to_month = release_to_month
)

# Preview the results
head(thoracic_res)
```




```{r tumor_board_overlap, message=FALSE, warning=FALSE, results='hide'}
# Define tumor board overlap SQL paths per release
tb_overlap_sql_paths <- list(
  feb_2025 = "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql_std/tumor_board/feb25/tumor_board_overlap.sql",
  may_2025 = "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql_std/tumor_board/may25/tumor_board_overlap.sql",
  aug_2025 = "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql_std/tumor_board/aug25/tumor_board_overlap.sql",
  nov_2025 = "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql_std/tumor_board/nov25/tumor_board_overlap.sql"
)

# Run tumor board overlap analysis across releases
tb_res <- run_monthly_metrics(
  datasets_by_month = datasets_by_month,
  sql_file_paths = tb_overlap_sql_paths,
  release_to_month = release_to_month
)

# Preview the results
head(tb_res)
```

```{r combined_plot, message=FALSE, warning=FALSE}
library(plotly)
library(dplyr)

# Prepare thoracic data
thoracic_plot_data <- thoracic_res %>%
  mutate(
    month = factor(month, levels = c("feb", "may", "aug", "nov")),
    month_label = recode(month,
      "feb" = "February", 
      "may" = "May",
      "aug" = "August",
      "nov" = "November"
    )
  ) %>%
  select(month, month_label, patients = unique_thoracic_cancer_pts) %>%
  mutate(metric = "Thoracic Cancer Patients")

# Prepare tumor board data
tb_plot_data <- tb_res %>%
  mutate(
    month = factor(month, levels = c("feb", "may", "aug", "nov")),
    month_label = recode(month,
      "feb" = "February",
      "may" = "May",
      "aug" = "August",
      "nov" = "November"
    ),
    metric = case_when(
      flag == "all_tb" ~ "All Tumor Board Patients",
      flag == "thoracic_tb" ~ "Thoracic Tumor Board Patients",
      flag == "tb_nf" ~ "TB Patients with NeuralFrame Data",
      flag == "tb_imaging_nf" ~ "TB Patients with Imaging & NeuralFrame",
      flag == "tb_genomic_nf_imaging" ~ "TB Patients with Genomic, Imaging & NeuralFrame",
      TRUE ~ flag
    )
  ) %>%
  select(month, month_label, patients = total_patients, metric)

# Combine the datasets
combined_data <- bind_rows(thoracic_plot_data, tb_plot_data)

# Create interactive area plot with consistent colors
plot <- plot_ly()

# Define consistent color palette
metric_colors <- c('#8dd3c7', '#ffffb3', '#bebada', '#fb8072', '#b3de69', '#fccde5')

# Add each metric as a separate trace with fill
metrics <- unique(combined_data$metric)

for(i in seq_along(metrics)) {
  metric_data <- combined_data %>% filter(metric == metrics[i])
  
  plot <- plot %>%
    add_trace(
      data = metric_data,
      x = ~month_label,
      y = ~patients,
      name = metrics[i],
      type = 'scatter',
      mode = 'lines',
      fill = 'tonexty',
      fillcolor = metric_colors[i],
      line = list(color = metric_colors[i]),
      text = ~paste0(
        metric,
        "<br>Month: ", month_label,
        "<br>Patients: ", scales::comma(patients)
      ),
      hoverinfo = "text"
    )
}

plot %>%
  layout(
    title = "Tumor Board and Thoracic Sub-Cohort Trends Across Releases",
    xaxis = list(
      title = "Release Month",
      categoryorder = "array",
      categoryarray = c("February", "May", "August", "November")
    ),
    yaxis = list(
      title = "Number of Patients",
      tickformat = ","
    ),
    legend = list(
      title = list(text = "Metric"),
      orientation = "h",
      x = 0.5,
      xanchor = "center",
      y = -0.2
    ),
    hovermode = 'closest',
    margin = list(b = 150)
  )
```

### Tumor Board Sub-Group 

```{r grouped_bar_plot, message=FALSE, warning=FALSE}
library(plotly)
library(dplyr)

# Create grouped bar chart with consistent colors
plot_ly(
  data = combined_data,
  x = ~month_label,
  y = ~patients,
  color = ~metric,
  colors = c('#8dd3c7', '#ffffb3', '#bebada', '#fb8072', '#b3de69', '#fccde5'),
  type = 'bar',
  text = ~paste0(
    metric,
    "<br>Patients: ", scales::comma(patients)
  ),
  hoverinfo = "text"
) %>%
  layout(
    title = "Tumor Board and Thoracic Sub-Cohort Comparison by Release",
    xaxis = list(
      title = "Release Month",
      categoryorder = "array",
      categoryarray = c("February", "May", "August", "November")
    ),
    yaxis = list(
      title = "Number of Patients",
      tickformat = ","
    ),
    barmode = 'group',
    legend = list(
      title = list(text = "Metric"),
      orientation = "h",
      x = 0.5,
      xanchor = "center",
      y = -0.3
    ),
    margin = list(b = 180)
  )
```
:::

## **Image Occurrence Modalities** ##

The VISTA Oncology Data Lake integrates comprehensive medical imaging data, linking diagnostic images to patient clinical records through the OMOP image occurrence table. This integration enables research connecting imaging studies with clinical outcomes, treatments, and genomic profiles.

**Imaging Coverage Growth**: The imaging dataset has expanded dramatically from approximately 93,000 patients in February 2025 to over 1.6 million patients by November 2025, representing a substantial increase in multi-modal data availability for cancer research.

**Image Series and Studies**: The platform captures detailed imaging metadata including study identifiers, series information, anatomic regions, and imaging modalities (CT, MRI, PET, etc.), supporting comprehensive radiomics and image-based research.

::: {.panel-tabset}

### Patient and Series Coverage

```{r image occ basics, message=FALSE, warning=FALSE, results='hide'}
# Define image person SQL paths per release
img_person_sql_paths <- list(
  feb_2025 = "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql_std/omop/image_occ/feb_aug25/image_occ_person.sql",
  may_2025 = "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql_std/omop/image_occ/feb_aug25/image_occ_person.sql", 
  aug_2025 = "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql_std/omop/image_occ/feb_aug25/image_occ_person.sql",
  nov_2025 = "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql_std/omop/image_occ/nov25/image_occ_person.sql"
)

img_res <- run_monthly_metrics(
  datasets_by_month = datasets_by_month,
  sql_file_paths = img_person_sql_paths,
  release_to_month = release_to_month
)

# Define image basic SQL paths per release
img_basic_sql_paths <- list(
  feb_2025 = "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql_std/omop/image_occ/feb_aug25/image_occ_basic.sql",
  may_2025 = "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql_std/omop/image_occ/feb_aug25/image_occ_basic.sql", 
  aug_2025 = "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql_std/omop/image_occ/feb_aug25/image_occ_basic.sql",
  nov_2025 = "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql_std/omop/image_occ/nov25/image_occ_basic.sql"
)

img_size <- run_monthly_metrics(
  datasets_by_month = datasets_by_month,
  sql_file_paths = img_basic_sql_paths,
  release_to_month = release_to_month
)
```

```{r combined_image_plots, message=FALSE, warning=FALSE, fig.width=12, fig.height=10}
library(plotly)
library(dplyr)

# Prepare img_res data with proper labels
img_res_plot <- img_res %>%
  mutate(
    month = factor(month, levels = c("feb", "may", "aug", "nov")),
    month_label = recode(month,
      "feb" = "February",
      "may" = "May",
      "aug" = "August",
      "nov" = "November"
    ),
    metric = case_when(
      grepl("unique persons", variable_name) ~ "Unique Patients",
      grepl("series", variable_name) ~ "Image Series",
      TRUE ~ variable_name
    )
  )

# Create dual-axis plot for patients and series
fig1 <- plot_ly()

# Add patients line (left y-axis)
patients_data <- img_res_plot %>% filter(metric == "Unique Patients")
fig1 <- fig1 %>%
  add_trace(
    data = patients_data,
    x = ~month_label,
    y = ~counts,
    name = "Unique Patients",
    type = 'scatter',
    mode = 'lines+markers',
    line = list(color = '#8dd3c7', width = 3),
    marker = list(size = 10, color = '#8dd3c7'),
    text = ~paste0("Patients<br>", month_label, "<br>Count: ", scales::comma(counts)),
    hoverinfo = "text"
  )

# Add series line (right y-axis)
series_data <- img_res_plot %>% filter(metric == "Image Series")
fig1 <- fig1 %>%
  add_trace(
    data = series_data,
    x = ~month_label,
    y = ~counts,
    name = "Image Series",
    type = 'scatter',
    mode = 'lines+markers',
    yaxis = "y2",
    line = list(color = '#ffffb3', width = 3),
    marker = list(size = 10, color = '#ffffb3'),
    text = ~paste0("Series<br>", month_label, "<br>Count: ", scales::comma(counts)),
    hoverinfo = "text"
  )

fig1 <- fig1 %>%
  layout(
    title = "Imaging Data Growth: Patient and Series Coverage",
    xaxis = list(
      title = "Release Month",
      categoryorder = "array",
      categoryarray = c("February", "May", "August", "November")
    ),
    yaxis = list(
      title = "Number of Unique Patients",
      tickformat = ",",
      side = "left"
    ),
    yaxis2 = list(
      title = "Number of Image Series",
      tickformat = ",",
      overlaying = "y",
      side = "right"
    ),
    legend = list(
      orientation = "h",
      x = 0.5,
      xanchor = "center",
      y = -0.15
    ),
    hovermode = 'closest',
    margin = list(b = 100)
  )

# Prepare img_size data for metadata plot
img_size_plot <- img_size %>%
  mutate(
    month = factor(month, levels = c("feb", "may", "aug", "nov")),
    month_label = recode(month,
      "feb" = "February",
      "may" = "May",
      "aug" = "August",
      "nov" = "November"
    ),
    metric = case_when(
      flag == "counts_anatomic_type" ~ "Anatomic Regions",
      flag == "counts_modality_type" ~ "Imaging Modalities",
      flag == "counts_study" ~ "Unique Studies",
      flag == "counts_series" ~ "Image Series",
      flag == "size_gb" ~ "Data Size (GB)",
      TRUE ~ flag
    )
  ) %>%
  filter(flag != "size_gb")

# Create metadata bar chart with consistent colors
fig2 <- plot_ly(
  data = img_size_plot,
  x = ~month_label,
  y = ~counts,
  color = ~metric,
  colors = c('#8dd3c7', '#ffffb3', '#bebada', '#fb8072'),
  type = 'bar',
  text = ~paste0(metric, "<br>", month_label, "<br>Count: ", scales::comma(counts)),
  hoverinfo = "text"
) %>%
  layout(
    title = "Imaging Metadata Diversity Across Releases",
    xaxis = list(
      title = "Release Month",
      categoryorder = "array",
      categoryarray = c("February", "May", "August", "November")
    ),
    yaxis = list(
      title = "Count",
      tickformat = ",",
      type = "log"
    ),
    barmode = 'group',
    legend = list(
      title = list(text = "Metric"),
      orientation = "h",
      x = 0.5,
      xanchor = "center",
      y = -0.15
    ),
    margin = list(b = 100)
  )

# Combine plots using subplot
subplot(
  fig1, fig2,
  nrows = 2,
  shareX = FALSE,
  titleY = TRUE,
  margin = 0.1
) %>%
  layout(
    title = list(
      text = "Comprehensive Imaging Data Evolution",
      y = 0.98
    ),
    showlegend = TRUE
  )
```

### Data Volume

```{r image_size_plot, message=FALSE, warning=FALSE}
library(plotly)
library(dplyr)

# Prepare size data
size_data <- img_size %>%
  filter(flag == "size_gb") %>%
  mutate(
    month = factor(month, levels = c("feb", "may", "aug", "nov")),
    month_label = recode(month,
      "feb" = "February",
      "may" = "May",
      "aug" = "August",
      "nov" = "November"
    )
  )

# Create area plot for data size
plot_ly(
  data = size_data,
  x = ~month_label,
  y = ~counts,
  type = 'scatter',
  mode = 'lines+markers',
  fill = 'tozeroy',
  fillcolor = 'rgba(141, 211, 199, 0.3)',
  line = list(color = '#8dd3c7', width = 3),
  marker = list(size = 12, color = '#8dd3c7'),
  text = ~paste0("Data Volume<br>", month_label, "<br>Size: ", counts, " GB"),
  hoverinfo = "text"
) %>%
  layout(
    title = "Imaging Data Volume Growth",
    xaxis = list(
      title = "Release Month",
      categoryorder = "array",
      categoryarray = c("February", "May", "August", "November")
    ),
    yaxis = list(
      title = "Data Size (GB)",
      tickformat = ","
    ),
    showlegend = FALSE,
    hovermode = 'closest'
  )
```

### Imaging Modalities

```{r image_modality, message=FALSE, warning=FALSE, results='hide'}
# Define image modality SQL paths per release
img_mod_sql_paths <- list(
  feb_2025 = "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql_std/omop/image_occ/feb_aug25/image_occ_modality.sql",
  may_2025 = "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql_std/omop/image_occ/feb_aug25/image_occ_modality.sql", 
  aug_2025 = "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql_std/omop/image_occ/feb_aug25/image_occ_modality.sql",
  nov_2025 = "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql_std/omop/image_occ/nov25/image_occ_modality.sql"
)

img_mod <- run_monthly_metrics(
  datasets_by_month = datasets_by_month,
  sql_file_paths = img_mod_sql_paths,
  release_to_month = release_to_month
)

library(dplyr)

img_mod <- img_mod %>%
  mutate(
    month = factor(month, levels = c("feb", "may", "aug", "nov")),
    month_label = recode(month,
      "feb" = "February",
      "may" = "May",
      "aug" = "August",
      "nov" = "November"
    ),
    modality_group = case_when(
      grepl("^CT|CT/|CT-", modality) | grepl("PET", modality) ~ "CT/PET",
      grepl("^MR|MRI|MR/", modality) ~ "MRI",
      grepl("^XR|X-RAY|XA|XC|XD|XR", modality) ~ "X-Ray",
      grepl("^US|ULTRASOUND|MG|MG/", modality) ~ "Ultrasound/Mammography",
      grepl("^NM|NM/", modality) ~ "Nuclear Medicine",
      grepl("^OT|OTHER|DOC|REPORT|REQUEST|FINDINGS|NONE|NULL", modality) ~ "Other/Docs",
      TRUE ~ "Other"
    )
  ) %>%
  filter(modality_group != "Other" & modality_group != "Other/Docs")

# Summarize by modality group and month
img_summary <- img_mod %>%
  group_by(month, month_label, modality_group) %>%
  summarise(counts = sum(unique_person_count), .groups = "drop")
```

```{r modality_plot, message=FALSE, warning=FALSE}
library(plotly)
library(dplyr)

# Create area chart showing modality growth with consistent colors
plot_ly(
  data = img_summary,
  x = ~month_label,
  y = ~counts,
  color = ~modality_group,
  colors = c('#8dd3c7', '#ffffb3', '#bebada', '#fb8072', '#b3de69'),
  type = 'scatter',
  mode = 'lines+markers',
  fill = 'tonexty',
  text = ~paste0(
    modality_group,
    "<br>Month: ", month_label,
    "<br>Patients: ", scales::comma(counts)
  ),
  hoverinfo = "text",
  line = list(width = 2),
  marker = list(size = 8)
) %>%
  layout(
    title = "Imaging Modality Coverage Across Releases",
    xaxis = list(
      title = "Release Month",
      categoryorder = "array",
      categoryarray = c("February", "May", "August", "November")
    ),
    yaxis = list(
      title = "Number of Patients",
      tickformat = ","
    ),
    legend = list(
      title = list(text = "Modality"),
      orientation = "h",
      x = 0.5,
      xanchor = "center",
      y = -0.2
    ),
    hovermode = 'closest',
    margin = list(b = 120)
  )
```

:::

## **Philips ISPM Genomic Testing** ##

The VISTA Oncology Data Lake integrates comprehensive molecular testing data from the Philips IntelliSpace Precision Medicine (ISPM) platform, beginning with the May 2025 release. This dataset includes genomic and molecular profiling results from various testing platforms including STAMP (Stanford Actionable Mutation Panel), Heme-STAMP, and FoundationOne, enabling research on precision oncology and genomic-driven treatment strategies.

**Testing Platform Coverage**: The ISPM dataset captures detailed molecular testing information including test types, accession numbers, order dates, and genomic aberrations across multiple assay platforms.

**Research Applications**: This integration supports studies of actionable mutations, variant frequencies, treatment selection based on genomic profiles, and longitudinal tracking of molecular testing patterns in cancer care.

::: {.panel-tabset}

### Overview Metrics

```{r philips_overview, message=FALSE, warning=FALSE, results='hide'}
# Define Philips denominator SQL paths per release (May onwards only)
philips_denom_sql_paths <- list(
  may_2025 = "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql_std/philips/may_aug25/philips_denominator.sql", 
  aug_2025 = "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql_std/philips/may_aug25/philips_denominator.sql",
  nov_2025 = "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql_std/philips/nov25/philips_denominator.sql"
)

philips_denom <- run_monthly_metrics(
  datasets_by_month = datasets_by_month[c("may_2025", "aug_2025", "nov_2025")],
  sql_file_paths = philips_denom_sql_paths,
  release_to_month = release_to_month
)
```

```{r philips_overview_plot, message=FALSE, warning=FALSE}
library(plotly)
library(dplyr)
library(tidyr)

# Prepare data in long format for grouped bar chart
philips_overview <- philips_denom %>%
  mutate(
    month = factor(month, levels = c("may", "aug", "nov")),
    month_label = recode(month,
      "may" = "May",
      "aug" = "August",
      "nov" = "November"
    )
  ) %>%
  pivot_longer(
    cols = c(counts_pts, n_acc),
    names_to = "metric",
    values_to = "count"
  ) %>%
  mutate(
    metric_label = recode(metric,
      "counts_pts" = "Unique Patients",
      "n_acc" = "Test Accessions"
    )
  )

# Create grouped bar chart
plot_ly(
  data = philips_overview,
  x = ~month_label,
  y = ~count,
  color = ~metric_label,
  colors = c('#8dd3c7', '#ffffb3'),
  type = 'bar',
  text = ~paste0(
    metric_label,
    "<br>", month_label,
    "<br>Count: ", scales::comma(count)
  ),
  hoverinfo = "text"
) %>%
  layout(
    title = "Philips ISPM Coverage: Patients and Test Accessions",
    xaxis = list(
      title = "Release Month",
      categoryorder = "array",
      categoryarray = c("May", "August", "November")
    ),
    yaxis = list(
      title = "Count",
      tickformat = ",",
      range = c(5000, 20000)
    ),
    barmode = 'group',
    legend = list(
      title = list(text = "Metric"),
      orientation = "h",
      x = 0.5,
      xanchor = "center",
      y = -0.15
    ),
    hovermode = 'closest',
    margin = list(b = 100)
  )
```

### Test Types

```{r philips_tests, message=FALSE, warning=FALSE, results='hide'}
# Define Philips tests ordered SQL paths per release
philips_tests_sql_paths <- list(
  may_2025 = "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql_std/philips/may_aug25/philips_tests_ordered.sql", 
  aug_2025 = "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql_std/philips/may_aug25/philips_tests_ordered.sql",
  nov_2025 = "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql_std/philips/nov25/philips_tests_ordered.sql"
)

philips_tests <- run_monthly_metrics(
  datasets_by_month = datasets_by_month[c("may_2025", "aug_2025", "nov_2025")],
  sql_file_paths = philips_tests_sql_paths,
  release_to_month = release_to_month
)
```

```{r philips_tests_plot, message=FALSE, warning=FALSE}
library(plotly)
library(dplyr)

# Prepare test type data
philips_tests_plot <- philips_tests %>%
  mutate(
    month = factor(month, levels = c("may", "aug", "nov")),
    month_label = recode(month,
      "may" = "May",
      "aug" = "August",
      "nov" = "November"
    ),
    # Clean up test type names
    test_type_clean = case_when(
      is.na(test_type) | test_type == "" ~ "Not Specified",
      TRUE ~ test_type
    )
  )

# Create grouped bar chart with consistent colors
plot_ly(
  data = philips_tests_plot,
  x = ~month_label,
  y = ~test_type_counts,
  color = ~test_type_clean,
  colors = c('#8dd3c7', '#ffffb3', '#bebada', '#fb8072', '#b3de69', '#fccde5', '#d9d9d9', '#bc80bd'),
  type = 'bar',
  text = ~paste0(
    test_type_clean,
    "<br>", month_label,
    "<br>Tests: ", scales::comma(test_type_counts)
  ),
  hoverinfo = "text"
) %>%
  layout(
    title = "Molecular Test Types Distribution Across Releases",
    xaxis = list(
      title = "Release Month",
      categoryorder = "array",
      categoryarray = c("May", "August", "November")
    ),
    yaxis = list(
      title = "Number of Tests",
      tickformat = ","
    ),
    barmode = 'group',
    legend = list(
      title = list(text = "Test Type"),
      orientation = "h",
      x = 0.5,
      xanchor = "center",
      y = -0.3
    ),
    margin = list(b = 150)
  )
```

### Test Type Trends

```{r philips_tests_area, message=FALSE, warning=FALSE}
library(plotly)
library(dplyr)

# Create stacked area chart with consistent colors
plot_ly(
  data = philips_tests_plot,
  x = ~month_label,
  y = ~test_type_counts,
  color = ~test_type_clean,
  colors = c('#8dd3c7', '#ffffb3', '#bebada', '#fb8072', '#b3de69', '#fccde5', '#d9d9d9', '#bc80bd'),
  type = 'scatter',
  mode = 'lines+markers',
  fill = 'tonexty',
  text = ~paste0(
    test_type_clean,
    "<br>Month: ", month_label,
    "<br>Tests: ", scales::comma(test_type_counts)
  ),
  hoverinfo = "text",
  line = list(width = 2),
  marker = list(size = 8)
) %>%
  layout(
    title = "Molecular Test Types Growth Over Time",
    xaxis = list(
      title = "Release Month",
      categoryorder = "array",
      categoryarray = c("May", "August", "November")
    ),
    yaxis = list(
      title = "Number of Tests",
      tickformat = ","
    ),
    legend = list(
      title = list(text = "Test Type"),
      orientation = "h",
      x = 0.5,
      xanchor = "center",
      y = -0.3
    ),
    hovermode = 'closest',
    margin = list(b = 150)
  )
```

### Stanford Test Types
```{r setup, include=FALSE}

# Your existing initialization
rm(list=ls())
source("/workspaces/starr-oncology-data-lake-arpah/src/R/all_function.R", encoding = "UTF-8")
yaml_file_path <- "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql_params.yml"
```

```{r philips new test types, message=FALSE, warning=FALSE, results='hide'}
sql_philips<- "/workspaces/starr-oncology-data-lake-arpah/src/nov_2025/sql_std/philips/nov25/philips_stanford_tests_ordered.sql"
philips_order<- fetch_data_from_sql_file(sql_philips, yaml_file_path)
philips_order
philips_order=philips_order %>% filter(!is.na(stanford_test_type))
```

```{r philips_stanford_table, message=FALSE, warning=FALSE}
library(DT)
library(dplyr)

# Prepare data for table
philips_order_table <- philips_order %>%
  arrange(test_type, desc(stanford_test_type_counts)) %>%
  mutate(
    test_type_clean = case_when(
      is.na(test_type) | test_type == "" ~ "Not Specified",
      TRUE ~ test_type
    ),
    stanford_test_type_counts_formatted = scales::comma(stanford_test_type_counts)
  ) %>%
  select(
    test_type_clean,
    stanford_test_type,
    stanford_test_type_counts_formatted
  )

# Create interactive table
datatable(
  philips_order_table,
  colnames = c(
    "Test Category",
    "Stanford Test Type",
    "Number of Tests"
  ),
  caption = "FoundationOne Test Types (November 2025)",
  filter = 'top',
  options = list(
    pageLength = 15,
    dom = 'Bfrtip',
    columnDefs = list(
      list(className = 'dt-left', targets = c(0, 1)),
      list(className = 'dt-right', targets = 2)
    ),
    order = list(list(2, 'desc'))
  ),
  rownames = FALSE
)
```

### Test Distribution
```{r philips_stanford_tests, message=FALSE, warning=FALSE}
library(plotly)
library(dplyr)

# Use the table data and convert counts back to numeric
philips_plot_data <- philips_order_table %>%
  mutate(
    count_numeric = as.numeric(gsub(",", "", stanford_test_type_counts_formatted))
  ) %>% 
  filter(test_type_clean!="StanfordStampTest") %>%
  arrange(desc(count_numeric))

# Create stacked bar chart
plot_ly(
  data = philips_plot_data,
  x = ~reorder(stanford_test_type, count_numeric),
  y = ~count_numeric,
  color = ~test_type_clean,
  colors = c('#8dd3c7', '#ffffb3', '#bebada', '#fb8072', '#b3de69'),
  type = 'bar',
  text = ~paste0(
    "Category: ", test_type_clean,
    "<br>Test: ", stanford_test_type,
    "<br>Count: ", stanford_test_type_counts_formatted
  ),
  hoverinfo = "text"
) %>%
  layout(
    title = "Stanford Test Types Distribution by Category (November 2025)",
    xaxis = list(
      title = "Stanford Test Type",
      tickangle = -45,
      categoryorder = "total descending"
    ),
    yaxis = list(
      title = "Number of Tests",
      tickformat = ","
    ),
    barmode = 'stack',
    legend = list(
      title = list(text = "Test Category"),
      orientation = "v",
      x = 1.02,
      y = 0.5
    ),
    hovermode = 'closest',
    margin = list(b = 150, r = 150)
  )
```

