---
title: "Neural Frame Event Table"
execute:
  echo: false
---

```{r load functions, message=FALSE, warning=FALSE, results='hide'}
#| context: global

rm(list=ls())
source("/workspaces/starr-oncology-data-lake-arpah/src/R/all_function.R", encoding = "UTF-8")
yaml_file_path <- "/workspaces/starr-oncology-data-lake-arpah/src/deid_aug_2026/sql_params.yml"
library(RColorBrewer)
library(plotly)
```


The Neural Frame Event table captures detailed clinical events and treatment information for cancer patients, including chemotherapy, radiation, surgery, clinical protocols, and various treatment outcomes. This comprehensive event-level data enables in-depth analysis of treatment patterns, clinical course, and patient outcomes.

```{r nf event table summary, message=FALSE, warning=FALSE, results='hide'}
sql_event_summary <- "/workspaces/starr-oncology-data-lake-arpah/src/sql_std/common/neural_frame/nf_event_summary.sql"
event_summary <- fetch_data_from_sql_file(sql_event_summary, yaml_file_path, release = "aug_2026")

sql_event_types <- "/workspaces/starr-oncology-data-lake-arpah/src/sql_std/common/neural_frame/nf_event_types.sql"
event_types_data <- fetch_data_from_sql_file(sql_event_types, yaml_file_path, release = "aug_2026")

sql_radiation_details <- "/workspaces/starr-oncology-data-lake-arpah/src/sql_std/common/neural_frame/nf_radiation_therapy.sql"
radiation_details <- fetch_data_from_sql_file(sql_radiation_details, yaml_file_path, release = "aug_2026")

sql_chemo_courses <- "/workspaces/starr-oncology-data-lake-arpah/src/sql_std/common/neural_frame/nf_chemotherapy_courses.sql"
chemo_courses <- fetch_data_from_sql_file(sql_chemo_courses, yaml_file_path, release = "aug_2026")

sql_surgery_details <- "/workspaces/starr-oncology-data-lake-arpah/src/sql_std/common/neural_frame/nf_surgical_procedures.sql"
surgery_details <- fetch_data_from_sql_file(sql_surgery_details, yaml_file_path, release = "aug_2026")

sql_completion_status <- "/workspaces/starr-oncology-data-lake-arpah/src/sql_std/common/neural_frame/nf_completion_status.sql"
completion_status <- fetch_data_from_sql_file(sql_completion_status, yaml_file_path, release = "aug_2026")

sql_recurrence <- "/workspaces/starr-oncology-data-lake-arpah/src/sql_std/common/neural_frame/nf_recurrence_analysis.sql"
recurrence_data <- fetch_data_from_sql_file(sql_recurrence, yaml_file_path, release = "aug_2026")

sql_temporal <- "/workspaces/starr-oncology-data-lake-arpah/src/sql_std/common/neural_frame/nf_event_temporal_trends.sql"
temporal_data <- fetch_data_from_sql_file(sql_temporal, yaml_file_path, release = "aug_2026")
```

```{r treatment overview summary, message=FALSE, warning=FALSE}
treatment_overview <- event_summary %>%
  mutate(
    `Unique Patients` = format(total_patients, big.mark = ","),
    `Total Cases` = format(total_cases, big.mark = ","),
    `Total Events` = format(total_events, big.mark = ","),
    `Unique Event Types` = unique_event_types
  ) %>%
  select(`Unique Patients`, `Total Cases`, `Total Events`, `Unique Event Types`)

create_gt_table_v1(
  data = treatment_overview,
  columns = c("Unique Patients", "Total Cases", "Total Events", "Unique Event Types"),
  labels = c(
    "Unique Patients" = "Patients",
    "Total Cases" = "Cases",
    "Total Events" = "Events",
    "Unique Event Types" = "Event Types"
  ),
  subtitle_text = "Neural Frame Event Table Summary Statistics"
)
```

#### Event Type Distribution Overview
```{r event types visualization, message=FALSE, warning=FALSE}
event_type_plot <- event_types_data %>%
  arrange(desc(event_count)) %>%
  head(12) %>%
  mutate(event_label = ifelse(nchar(eventtypedescription) > 20,
                               substr(eventtypedescription, 1, 20),
                               eventtypedescription))

plot_ly(
  data = event_type_plot,
  x = ~reorder(event_label, -event_count),
  y = ~event_count,
  type = 'bar',
  marker = list(color = '#1f77b4'),
  hovertemplate = ~paste0(
    "<b>", eventtypedescription, "</b>",
    "<br>Events: ", scales::comma(event_count),
    "<br>Cases: ", scales::comma(case_count),
    "<br>Patients: ", scales::comma(patient_count),
    "<extra></extra>"
  )
) %>%
  layout(
    title = "Top 12 Event Types in Neural Frame",
    xaxis = list(title = "Event Type"),
    yaxis = list(title = "Number of Events", tickformat = ","),
    showlegend = FALSE,
    hovermode = 'x unified',
    margin = list(b = 120)
  )
```

#### Event Temporal Trends
```{r event temporal trends, message=FALSE, warning=FALSE}
temporal_plot <- temporal_data %>%
  mutate(year_num = suppressWarnings(as.numeric(year))) %>%
  filter(!is.na(year_num), year_num <= 2026) %>%
  mutate(
    year_display = ifelse(year_num < 2000, "<2000", as.character(year_num)),
    year_sort = ifelse(year_num < 2000, 1999, year_num)
  ) %>%
  group_by(year_sort, year_display) %>%
  summarise(
    distinct_patients = sum(distinct_patients),
    total_events = sum(total_events),
    .groups = "drop"
  ) %>%
  arrange(year_sort)

plot_ly(data = temporal_plot) %>%
  add_trace(
    x = ~year_sort, y = ~distinct_patients,
    type = 'bar',
    name = 'Patients',
    marker = list(color = '#89cbad'),
    hovertemplate = ~paste0(
      "<b>Year: ", year_display, "</b>",
      "<br>Patients: ", scales::comma(distinct_patients),
      "<br>Events: ", scales::comma(total_events),
      "<extra></extra>"
    )
  ) %>%
  add_trace(
    x = ~year_sort, y = ~total_events,
    type = 'scatter', mode = 'lines+markers',
    name = 'Events',
    yaxis = 'y2',
    line = list(color = '#f09090'),
    marker = list(color = '#d62728', size = 4),
    hovertemplate = ~paste0(
      "<b>Year: ", year_display, "</b>",
      "<br>Events: ", scales::comma(total_events),
      "<extra></extra>"
    )
  ) %>%
  layout(
    title = "Neural Frame Events and Patients Over Time",
    xaxis = list(
      title = "Year",
      tickmode = "array",
      tickvals = temporal_plot$year_sort,
      ticktext = temporal_plot$year_display
    ),
    yaxis = list(title = "Distinct Patients", tickformat = ","),
    yaxis2 = list(title = "Total Events", overlaying = "y", side = "right", tickformat = ","),
    legend = list(x = 0.01, y = 0.99),
    hovermode = 'x unified'
  )
```

This section provides a comprehensive overview of treatment patterns in the Neural Frame Event data, including chemotherapy, radiation, surgery, and other therapeutic interventions.

::: {.panel-tabset}

### ARIA Radiation

ARIA (Radiation Oncology Information System) provides per-phase radiation treatment logistics that are not available from general NF abstracting. These include target volumes, lymph node involvement, daily/total dosing, fraction counts, and treatment end dates.

```{r aria phase summary, message=FALSE, warning=FALSE, results='hide'}
sql_aria_summary <- "/workspaces/starr-oncology-data-lake-arpah/src/sql_std/common/neural_frame/nf_aria_phase_summary.sql"
aria_summary <- fetch_data_from_sql_file(sql_aria_summary, yaml_file_path, release = "aug_2026")

sql_aria_detail <- "/workspaces/starr-oncology-data-lake-arpah/src/sql_std/common/neural_frame/nf_aria_phase_detail.sql"
aria_detail <- fetch_data_from_sql_file(sql_aria_detail, yaml_file_path, release = "aug_2026")
```

```{r aria summary table, message=FALSE, warning=FALSE}
aria_overview <- aria_summary %>%
  mutate(
    `Patients` = format(patients_with_aria_phase_data, big.mark = ","),
    `Cases` = format(cases_with_aria_phase_data, big.mark = ","),
    `Events` = format(events_with_aria_phase_data, big.mark = ",")
  ) %>%
  select(`Patients`, `Cases`, `Events`)

create_gt_table_v1(
  data = aria_overview,
  columns = c("Patients", "Cases", "Events"),
  labels = c(
    "Patients" = "Number of Patients",
    "Cases" = "Number of Cases",
    "Events" = "Number of Events"
  ),
  subtitle_text = ""
)
```

#### ARIA Field Coverage
The following table shows how many patients and cases have data for each ARIA-specific field:

```{r aria detail table, message=FALSE, warning=FALSE}
aria_detail_table <- aria_detail %>%
  arrange(desc(patients)) %>%
  mutate(
    `Field` = aria_field,
    `Patients` = format(patients, big.mark = ","),
    `Cases` = format(cases, big.mark = ",")
  ) %>%
  select(`Field`, `Patients`, `Cases`)

create_gt_table_v1(
  data = aria_detail_table,
  columns = c("Field", "Patients", "Cases"),
  labels = c(
    "Field" = "ARIA Field",
    "Patients" = "Patients",
    "Cases" = "Cases"
  ),
  subtitle_text = "Coverage by ARIA Phase-Level Field"
)
```

#### Thoracic Tumor Board: ARIA Metrics 

The following ARIA metrics are restricted to thoracic tumor board patients with radiation events only:

```{r thoracic tumor board aria table, message=FALSE, warning=FALSE, results='hide'}
sql_tb_aria <- "/workspaces/starr-oncology-data-lake-arpah/src/sql_std/common/neural_frame/nf_aria_thoracic_tumor_board.sql"
tb_aria_table <- fetch_data_from_sql_file(sql_tb_aria, yaml_file_path, release = "aug_2026") %>%
  arrange(desc(patients)) %>%
  mutate(
    `Field` = aria_field,
    `Patients` = format(patients, big.mark = ","),
    `Cases` = format(cases, big.mark = ",")
  ) %>%
  select(`Field`, `Patients`, `Cases`)
```

```{r thoracic tumor board aria table output, message=FALSE, warning=FALSE}
create_gt_table_v1(
  data = tb_aria_table,
  columns = c("Field", "Patients", "Cases"),
  labels = c(
    "Field" = "ARIA Field",
    "Patients" = "Patients",
    "Cases" = "Cases"
  ),
  subtitle_text = "Thoracic Tumor Board ARIA Metrics"
)
```

### Event Table

::: {.panel-tabset}

### Chemotherapy Courses
```{r chemotherapy courses table, message=FALSE, warning=FALSE}
chemo_table <- chemo_courses %>%
  arrange(eventcoursenumber) %>%
  filter(patients_receiving_chemo > 5,
         eventcoursenumberdescription != "Non-treatment related") %>%
  mutate(
    `Course` = eventcoursenumberdescription,
    `Cases` = format(cases_receiving_chemo, big.mark = ","),
    `Patients` = format(patients_receiving_chemo, big.mark = ","),
    `Events` = format(chemo_events, big.mark = ",")
  ) %>%
  select(`Course`, `Cases`, `Patients`, `Events`)

create_gt_table_v1(
  data = chemo_table,
  columns = c("Course", "Cases", "Patients", "Events"),
  labels = c(
    "Course" = "Course",
    "Cases" = "Cases",
    "Patients" = "Patients",
    "Events" = "Events"
  ),
  subtitle_text = "Chemotherapy Course Distribution"
)
```

### Surgical Procedures
```{r surgical procedures table, message=FALSE, warning=FALSE}
surg_table <- surgery_details %>%
  arrange(desc(cases_with_surgery)) %>%
  head(15) %>%
  mutate(
    `Procedure` = surgery_type,
    `Cases` = format(cases_with_surgery, big.mark = ","),
    `Patients` = format(patients_with_surgery, big.mark = ",")
  ) %>%
  select(`Procedure`, `Cases`, `Patients`)

create_gt_table_v1(
  data = surg_table,
  columns = c("Procedure", "Cases", "Patients"),
  labels = c(
    "Procedure" = "Surgery Type",
    "Cases" = "Cases",
    "Patients" = "Patients"
  ),
  subtitle_text = "Top 15 Surgical Procedures"
)
```

### Recurrence
```{r recurrence table, message=FALSE, warning=FALSE}
rec_table <- recurrence_data %>%
  filter(patients_with_recurrence >= 10) %>%
  arrange(desc(cases_with_recurrence)) %>%
  mutate(
    `Recurrence Type` = recurrence_type,
    `Cases` = format(cases_with_recurrence, big.mark = ","),
    `Patients` = format(patients_with_recurrence, big.mark = ",")
  ) %>%
  select(`Recurrence Type`, `Cases`, `Patients`)

create_gt_table_v1(
  data = rec_table,
  columns = c("Recurrence Type", "Cases", "Patients"),
  labels = c(
    "Recurrence Type" = "Recurrence Type",
    "Cases" = "Cases",
    "Patients" = "Patients"
  ),
  subtitle_text = "Recurrence Type Distribution"
)
```

:::

:::

### Source Codes:

The source codes for this page can be found [here](https://github.com/susom/starr-oncology-data-lake-arpah/tree/main/src/deid_aug_2026/neural_frame_event_table.qmd), and the sql queries that support the metrics can be found [here](https://github.com/susom/starr-oncology-data-lake-arpah/tree/main/src/sql_std/common/neural_frame).
