---
title: "OMOP Image Occurrence"
execute:
  echo: false
---
```{r setup, include=FALSE}

# Your existing initialization
rm(list=ls())
source("/workspaces/starr-oncology-data-lake-arpah/src/R/all_function.R", encoding = "UTF-8")
yaml_file_path <- "/workspaces/starr-oncology-data-lake-arpah/src/aug_2025/sql_params.yml"
```

```{r modality sql, message=FALSE, warning=FALSE, results='hide'}

library(forcats)
sql_file_path_image_mod <- "/workspaces/starr-oncology-data-lake-arpah/src/aug_2025/sql/image_occ/image_occ_modality.sql"
res_img_mod <- fetch_data_from_sql_file(sql_file_path_image_mod, yaml_file_path)
head(res_img_mod)
res_img_mod$variable_name<-NULL
colnames(res_img_mod)=c("modality_type", "series_count", "person_count")
```

```{r modality dictionary,  warning=FALSE, results='hide'}


Value <- c("AR", "AS", "ASMT", "AU", "BDUS", "BI", "BMD", "CD", "CF", "CP", 
           "CR", "CS", "CT", "DD", "DF", "DG", "DM", "DOC", "DS", "DX", 
           "EC", "ECG", "EPS", "ES", "FA", "FID", "FS", "GM", "HC", "HD", 
           "IO", "IOL", "IVOCT", "IVUS", "KER", "KO", "LEN", "LP", "LS", 
           "MA", "MG", "MR", "MS", "NM", "OAM", "OCT", "OP", "OPM", "OPR", 
           "OPT", "OPV", "OSS", "OT", "PLAN", "PR", "PT", "PX", "REG", 
           "RESP", "RF", "RG", "RTDOSE", "RTIMAGE", "RTPLAN", "RTRECORD", 
           "RTSTRUCT", "RWV", "SEG", "SM", "SMR", "SR", "SRF", "ST", 
           "STAIN", "TG", "US", "VA", "VF", "XA", "XC")

Description <- c("Autorefraction", "Angioscopy", "Content Assessment Results", 
                 "Audio", "Bone Densitometry (ultrasound)", "Biomagnetic imaging", 
                 "Bone Densitometry (X-Ray)", "Color flow Doppler", 
                 "Cinefluorography", "Colposcopy", "Computed Radiography", 
                 "Cystoscopy", "Computed Tomography", "Duplex Doppler", 
                 "Digital fluoroscopy", "Diaphanography", "Digital microscopy", 
                 "Document", "Digital Subtraction Angiography", "Digital Radiography", 
                 "Echocardiography", "Electrocardiography", 
                 "Cardiac Electrophysiology", "Endoscopy", 
                 "Fluorescein angiography", "Fiducials", "Fundoscopy", 
                 "General Microscopy", "Hard Copy", "Hemodynamic Waveform", 
                 "Intra-Oral Radiography", "Intraocular Lens Data", 
                 "Intravascular Optical Coherence Tomography", 
                 "Intravascular Ultrasound", "Keratometry", 
                 "Key Object Selection", "Lensometry", "Laparoscopy", 
                 "Laser surface scan", "Magnetic resonance angiography", 
                 "Mammography", "Magnetic Resonance", 
                 "Magnetic resonance spectroscopy", "Nuclear Medicine", 
                 "Ophthalmic Axial Measurements", 
                 "Optical Coherence Tomography (non-Ophthalmic)", 
                 "Ophthalmic Photography", "Ophthalmic Mapping", 
                 "Ophthalmic Refraction", "Ophthalmic Tomography", 
                 "Ophthalmic Visual Field", "Optical Surface Scan", 
                 "Other", "Plan", "Presentation State", 
                 "Positron emission tomography (PET)", "Panoramic X-Ray", 
                 "Registration", "Respiratory Waveform", 
                 "Radio Fluoroscopy", "Radiographic imaging (conventional film/screen)", 
                 "Radiotherapy Dose", "Radiotherapy Image", 
                 "Radiotherapy Plan", "RT Treatment Record", 
                 "Radiotherapy Structure Set", "Real World Value Map", 
                 "Segmentation", "Slide Microscopy", "Stereometric Relationship", 
                 "SR Document", "Subjective Refraction", 
                 "Single-photon emission computed tomography (SPECT)", 
                 "Automated Slide Stainer", "Thermography", 
                 "Ultrasound", "Visual Acuity", "Videofluorography", 
                 "X-Ray Angiography", "External-camera Photography")

retired <- c(NA, "Retired", NA, NA, NA, NA, NA, "Retired", "Retired", 
             "Retired", NA, "Retired", NA, "Retired", "Retired", NA, 
             "Retired", NA, NA, NA, "Retired", NA, NA, NA, "Retired", 
             NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, "Retired", 
             "Retired", NA, NA, "Retired", NA, "Retired", NA, 
             "Retired", NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, 
             NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, 
             NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, 
             "Retired", NA)

# Create the data frame
dat_modality <- data.frame(Value, Description, retired, stringsAsFactors = FALSE)

```
### Summary:

In the OMOP Common Data Model (CDM), the "Image_occurrence" table provides a structured way to represent medical imaging events, bridging the gap between imaging and observational research by integrating image-based measurements into the CDM. 

The OMOP image_occurrence table captures records of medical imaging events at the series level, representing events where one or more related images are acquired as part of a clinical procedure, and it also provides a path to de-identified DICOM images.

Each record includes metadata such as the modality (e.g., CT, MRI) and the anatomic site of the image. The table maintains foreign key links to the procedure_occurrence and visit_occurrence tables, enabling contextual association with the related clinical procedure and patient encounter. 

At Stanford, this table has been extended to include a link to the note table, allowing integration of imaging data with relevant clinical documentation. We expect to have larger number of images in the coming months. 

```{r, message=FALSE, warning=FALSE, results='hide'}
## number of pts and number of series 
path_img <- "/workspaces/starr-oncology-data-lake-arpah/src/aug_2025/sql/image_occ/image_occ_person.sql"
pt_img <- fetch_data_from_sql_file(path_img, yaml_file_path)
pt_img
den_series=as.numeric(pt_img$counts)[2]
den_pts=as.numeric(pt_img$counts)[1]
```

As of August 2025, this table included images for  `r format(pt_img$counts[1], big.mark = ",")` patients with `r format(pt_img$counts[2], big.mark = ",")` series. **Series** are defined as individual imaging acquisitions within a study.

### Source Queries:
The queries related to the following analysis can be accessed [here](/workspaces/starr-oncology-data-lake-arpah/src/aug_2025/sql/image_occ). 
The queries related to the following tables and graphs can be found [here](/workspaces/starr-oncology-data-lake-arpah/src/aug_2025/image_occ.qmd).
The queries regarding thoracic cancer can be found [here](/workspaces/starr-oncology-data-lake-arpah/src/aug_2025/sql/image_occ/image_occ_thoracic)

```{r, message=FALSE, warning=FALSE, results='hide'}
## basic counts 
sql_file_path_image<- "/workspaces/starr-oncology-data-lake-arpah/src/aug_2025/sql/image_occ/image_occ_basic.sql"
res<- fetch_data_from_sql_file(sql_file_path_image, yaml_file_path)
head(res)


##- **>`r format(res$counts[res$flag=="size_gb"], big.mark = ",")` GB** of imaging data 
```

::: {.columns}

::: {.column width="40%"}
### 📊 Data Volume 
- Patient Count: `r format(pt_img$counts[1], big.mark = ",")`
- Series Count: `r format(pt_img$counts[2], big.mark = ",")`
- Study UID: `r format(res$counts[res$flag=="counts_study"], big.mark = ",")`
:::

::: {.column width="60%"}
### 🧬 Data Components 
- `r format(res$counts[res$flag=="counts_modality_type"], big.mark = ",")` Modality types 
- `r format(res$counts[res$flag=="counts_anatomic_type"], big.mark = ",")` Anatomic sites 
:::



## Image Occurrence Metrics {.tabset}
This visualization summarizes distribution of Modality categories and Anatomic sites by series and patient counts for all cancer types and for thoracic patients. 
Note: We expect the number of patients with imaging data to increase per release. 

::: {.panel-tabset}

### All Cancer Types 
## Modality Types
Top modality  types for series above 5k occurrences are listed as follow.

**Note:** The following modality source values without image pixel were excluded from the analysis:

SR: Structured Report
REG: Registration
KO: Key Object Selection
PR: Presentation State


```{r, message=FALSE, warning=FALSE, results='hide'}
colnames(dat_modality)[1]<-"modality_type"
res_img_mod=res_img_mod %>% left_join(dat_modality,by="modality_type" )
res_img_mod=res_img_mod %>% filter(person_count>10)
res_img_mod$retired<-NULL
mod_ex=c("SR", "PR", "REG", "KO")
res_img_mod= res_img_mod %>% filter (!modality_type %in% mod_ex)
res_img_mod = res_img_mod %>% mutate(percent_pts = round(100 * person_count/den_pts, 1)) %>% mutate(percent_series = round(100 * series_count/den_series, 1)) 
```

```{r, message=FALSE, warning=FALSE}
## Number of series in each modality - bar plot to keep 
library(plotly)
library(plotly)
df_bar <- res_img_mod %>% 
  filter(series_count > 5000) %>%
  arrange(desc(series_count)) %>% 
  mutate(
    modality_type = fct_reorder(modality_type, series_count),
    percent_display = ifelse(percent_series < 2, 
                            "<2", 
                            sprintf("%.1f", percent_series))
  )

plot_ly(df_bar,
        y = ~modality_type,
        x = ~series_count,
        customdata = ~percent_display,  
         text = ~paste0(percent_display, "%"), 
          textposition = "outside", 
        type = "bar",
        color = ~modality_type,
        colors = rev(RColorBrewer::brewer.pal(n = length(unique(df_bar$modality_type)), "Set3")),
        hovertemplate = paste(
          "Modality: %{y}<br>",
          "Series Count: %{x:,.0f}<br>",
          "Series Percentage: %{customdata}%<br>",
          "<extra></extra>"
        )) %>%
  layout(
    xaxis = list(title = "Series Count"),
    yaxis = list(title = ""),
    title = "Number of Series per Modality",
    showlegend = FALSE,
    uniformtext = list(minsize = 8, mode = "show")
  )
```

```{r, message=FALSE, warning=FALSE, results='hide'}

## hiding the pie ## 
df_pie <- res_img_mod %>%
  group_by(modality_type) %>%  filter(series_count>5000)%>%
  arrange(desc(person_count)) 

plot_ly(
  df_pie,
  labels = ~modality_type,
  values = ~series_count,
  type = "pie",
  text = ~paste("Modality:", modality_type, "<br>Series Count:", format(series_count, big.mark = ",")),
  hoverinfo = "text",
  marker = list(colors = RColorBrewer::brewer.pal(n = max(3, nrow(df_pie)), "Set3")),
  textinfo = "label+percent"
) %>%
  layout(
    title = list(text = "Number of Series per Modality", x = 0.5),
    legend = list(
      orientation = "v",
      x = -0.1,     
      y = 0.5,     
      xanchor = "left",
      font = list(size = 12)
    )
  )
```


## Patient Population 
**Note:** This visualization shows modality types with more than 1,000 patients.

```{r, message=FALSE, warning=FALSE}
df_bar <- res_img_mod %>% filter(person_count>1000)%>%
  arrange(desc(person_count)) %>% mutate(modality_type = fct_reorder(modality_type, person_count)) %>% mutate(
    percent_display = ifelse(percent_pts < 2, 
                            "<2", 
                            sprintf("%.1f", percent_pts))
  )




plot_ly(df_bar,
        y = ~modality_type,
        x = ~person_count,
        text = ~paste0(percent_display, "%") ,
        textposition = "outside",  
        customdata = ~percent_display,
        type = "bar",
        color = ~modality_type,
        colors = rev(RColorBrewer::brewer.pal(n = length(unique(df_bar$modality_type)), "Set3")),
        hovertemplate = paste(
          "Modality: %{y}<br>",
          "Patient Count: %{x:,.0f}<br>",
           "Patient Percentage: %{customdata:.1f}%<br>",
          "<extra></extra>"
        )) %>%
  layout(
    xaxis = list(title = "Person Count"),
    yaxis = list(title = ""),
    title = "Number of Patients per Modality",
    showlegend = FALSE,
    uniformtext = list(minsize = 8, mode = "show")
  )

```
## Modality Descriptions and Frequency by Series and Patients 
Please note that the modality descriptions are listed here: [Modality Descriptions (DICOM Library)](https://www.dicomlibrary.com/dicom/modality/).

```{r, message=FALSE, warning=FALSE}
library(DT)
res_img_mod=res_img_mod %>% mutate(modality_description = paste0(Description, " (", modality_type, ")"))
res_img_mod$Description[is.na(res_img_mod$Description)]<-"Not Available"
#res_img_mod$modality_type<-NULL
res_img_mod<-res_img_mod %>% select (Description,modality_type,series_count,person_count)
DT::datatable(as.data.frame(res_img_mod), options = list(scrollX = TRUE, pageLength = 10))%>%
  formatRound(columns = c("person_count", "series_count"), digits = 0)

## https://www.dicomlibrary.com/dicom/modality/

## drop or regrp the documents and patterns by pixel data 
## round to 1 decimal not 2
```


## Anatomic Sites
Top anatomic sites for series above 100k occurrences are listed as follow.

```{r, message=FALSE, warning=FALSE, results='hide'}
#| context: global

sql_file_path_image_an <- "/workspaces/starr-oncology-data-lake-arpah/src/aug_2025/sql/image_occ/image_occ_anatomic.sql"
res_ana <- fetch_data_from_sql_file(sql_file_path_image_an, yaml_file_path)
head(res_ana)
res_ana$variable_name<-NULL
colnames(res_ana)=c("anatomic_site", "series_count", "person_count")
res_ana=res_ana %>% filter(person_count>10)

res_ana = res_ana %>% mutate(percent_pts = round(100 * person_count/den_pts, 1)) %>% mutate(percent_series = round(100 * series_count/den_series, 1)) 

```

```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(plotly)
library(DT)

df_bar_sr <- res_ana %>% filter(series_count>100000)%>%
  arrange(desc(series_count)) %>% mutate(anatomic_site = forcats::fct_reorder(anatomic_site, series_count))


plot_ly(df_bar_sr,
        y = ~anatomic_site,
        x = ~series_count,
        #text = ~sprintf("%.1f%%", percent_series),  
       # textposition = "outside",  
        customdata = ~percent_series,
        type = "bar",
        color = ~anatomic_site,
                colors = rev(RColorBrewer::brewer.pal(n = length(unique(df_bar_sr$anatomic_site_source_value)), "Set3")),
        hovertemplate = paste(
          "Anatomic: %{y}<br>",
          "Series Count: %{x:,.0f}<br>",
          "Series Percentage: %{customdata:.1f}%<br>",
          "<extra></extra>"
        )) %>%
  layout(
    xaxis = list(title = "Series Count"),
    yaxis = list(title = ""),
    title = "Series Count by Anatomic Site",
    showlegend = FALSE,
    uniformtext = list(minsize = 8, mode = "show"), 
    margin = list(r = 150) 
  )         

```
## Patient Population 
**Note:** This visualization shows anatomic sites with more than 100k series. 

```{r, message=FALSE, warning=FALSE}
library(forcats)
df_bar <- res_ana %>% filter(series_count>100000)%>%
  arrange(desc(person_count)) %>% mutate(anatomic_site = forcats::fct_reorder(anatomic_site, person_count))

plot_ly(df_bar,
        y = ~anatomic_site,
        x = ~person_count,
       # text = ~sprintf("%.1f%%", percent_pts),  
      #  textposition = "outside",  
       customdata = ~percent_pts,
        type = "bar",
        color = ~anatomic_site,
        colors = rev(RColorBrewer::brewer.pal(n = length(unique(df_bar$anatomic_site_source_value)), "Set3")),
        hovertemplate = paste(
          "Anatomic: %{y}<br>",
          "Patient Count: %{x:,.0f}<br>",
          "Patient Percentage: %{customdata:.1f}%<br>",
          "<extra></extra>"
        )) %>%
  layout(
    xaxis = list(title = "Person Count"),
    yaxis = list(title = ""),
    title = "Number of Patients per Anatomic Site",
    showlegend = FALSE,
    uniformtext = list(minsize = 8, mode = "show")
  )

```
## Anatomic Site Descriptions and Frequency by Series and Patients 

```{r, message=FALSE, warning=FALSE}
library(DT)

res_ana=res_ana %>% select (anatomic_site,series_count,person_count)

DT::datatable(as.data.frame(res_ana), options = list(scrollX = TRUE, pageLength = 10))%>%
  formatRound(columns = c("person_count", "series_count"), digits = 0)

```
### Thoracic Cancer 
Thoracic cancer patients are identified based on their primary site descriptions in the Neural Frame diagnoses data, which include diagnoses of lung, bronchus, or thymus cancers.
```{r, message=FALSE, warning=FALSE, results='hide'}
## basic counts 
sql_thr<- "/workspaces/starr-oncology-data-lake-arpah/src/aug_2025/sql/image_occ/image_occ_thoracic/image_occ_person_thr.sql"
res_den_thr<- fetch_data_from_sql_file(sql_thr, yaml_file_path)
head(res_den_thr)
```

As of August 2025, from `r format(res_den_thr$counts[res_den_thr$variable_name=="Number of unique persons with thoracic cancer"], big.mark = ",")` patients with thoracic cancer, `r format(res_den_thr$counts[res_den_thr$variable_name=="Number of unique persons in image_occurrence table"], big.mark = ",")` had imaging data. 



## Modality Types
Top modality types for series above 5k occurrences are listed as follow.
```{r modality thoracic, message=FALSE, warning=FALSE, results='hide'}
## modality
sql_thr_mod<- "/workspaces/starr-oncology-data-lake-arpah/src/aug_2025/sql/image_occ/image_occ_thoracic/image_occ_modality_thr.sql"
res_thr<- fetch_data_from_sql_file(sql_thr_mod, yaml_file_path)
head(res_thr)
res_thr
thr_pts_den=res_den_thr$counts[2]
thr_ser_den=res_den_thr$counts[3]

res_thr <- res_thr %>% mutate(percent_pts = round(100 * unique_person_count/thr_pts_den, 1),
    percent_series = round(100 * series_count/thr_ser_den, 1))
```

```{r modality plot, message=FALSE, warning=FALSE}
# Prepare the data
res_thr <- res_thr %>% filter (series_count>5000) %>%
  arrange(desc(series_count)) %>%
  mutate(
    modality_source_value = factor(modality_source_value, levels = unique(modality_source_value))
  )


```

```{r, message=FALSE, warning=FALSE}

plot <- plot_ly() %>%
  add_trace(
    data = res_thr,
    x = ~modality_source_value,
    y = ~series_count,
    type = "bar",
    name = "Series Count",
    marker = list(color = "#8B0000"),
    hovertemplate = paste(
      "Series Count: %{y:,.0f}<br>",
      "Series (%): %{text}%<br>",
      "<extra></extra>"
    ),
    text = ~percent_series
  ) %>%
  # Add line for person count
  add_trace(
    data = res_thr,
    x = ~modality_source_value,
    y = ~unique_person_count,
    type = "scatter",
    mode = "lines+markers",
    name = "Person Count",
    yaxis = "y2",
    line = list(color = "gray", width = 2),
    marker = list(color = "gray", size = 8),
    hovertemplate = paste(
      "Patient Count: %{y:,.0f}<br>",
      "Patient (%): %{text}%<br>",
      "<extra></extra>"
    ),
    text = ~percent_pts
  ) %>%
  layout(
    title = "Imaging Modalities: Series Count vs Patient Count",
    xaxis = list(
      title = "Modality Type",
      tickangle = 45
    ),
    yaxis = list(
      title = "Series Count",
      side = "left"
    ),
    yaxis2 = list(
      title = "",
      side = "right",
      overlaying = "y",
      showline = FALSE,
      showgrid = FALSE
    ),
    legend = list(x = 0.8, y = 0.9),
    margin = list(b = 90),
    hovermode = "x unified"
  )
plot 

```


## Anatomic Sites 
Top anatomic sites for series above 5k occurrences are listed as follow.
```{r, message=FALSE, warning=FALSE, results='hide'}
## anatomic
sql_thr_ana<- "/workspaces/starr-oncology-data-lake-arpah/src/aug_2025/sql/image_occ/image_occ_thoracic/image_occ_anatomic._thr.sql"
res_ana_thr<- fetch_data_from_sql_file(sql_thr_ana, yaml_file_path)
head(res_ana_thr)

res_ana_thr <- res_ana_thr %>% mutate(percent_pts = round(100 * unique_person_count/thr_pts_den, 1),
    percent_series = round(100 * series_count/thr_ser_den, 1))

```
```{r anatomic, message=FALSE, warning=FALSE}
# Prepare the data
res_ana_thr <- res_ana_thr %>% filter (series_count>5000) %>%
  arrange(desc(series_count)) %>%
  mutate(
    anatomic_site_source_value = factor(anatomic_site_source_value, levels = unique(anatomic_site_source_value))
  )



plot <- plot_ly() %>%
  add_trace(
    data = res_ana_thr,
    x = ~anatomic_site_source_value,
    y = ~series_count,
    type = "bar",
    name = "Series Count",
    marker = list(color = "#8B0000"),
    hovertemplate = paste(
      "Series Count: %{y:,.0f}<br>",
      "Series (%): %{text}%<br>",
      "<extra></extra>"
    ),
    text = ~percent_series
  ) %>%
  # Add line for person count
  add_trace(
    data = res_ana_thr,
    x = ~anatomic_site_source_value,
    y = ~unique_person_count,
    type = "scatter",
    mode = "lines+markers",
    name = "Person Count",
    yaxis = "y2",
    line = list(color = "gray", width = 2),
    marker = list(color = "gray", size = 8),
    hovertemplate = paste(
      "Patient Count: %{y:,.0f}<br>",
      "Patient (%): %{text}%<br>",
      "<extra></extra>"
    ),
    text = ~percent_pts
  ) %>%
  layout(
    title = "Imaging Anatomic Site: Series Count vs Patient Count",
    xaxis = list(
      title = "Anatomic Sites",
      tickangle = 45
    ),
    yaxis = list(
      title = "Series Count",
      side = "left"
    ),
    yaxis2 = list(
      title = "",
      side = "right",
      overlaying = "y",
      showline = FALSE,
      showgrid = FALSE
    ),
    legend = list(x = 0.8, y = 0.9),
    margin = list(b = 90),
    hovermode = "x unified"
  )
plot 



```
:::

