---
title: "OMOP Variant Occurrence"
execute:
  echo: false
---
```{r setup, include=FALSE}
# Your existing initialization
rm(list=ls())
source("/workspaces/starr-oncology-data-lake-arpah/src/R/all_function.R", encoding = "UTF-8")
yaml_file_path <- "/workspaces/starr-oncology-data-lake-arpah/src/feb_2026/sql_params.yml"
```

```{r variant sql, message=FALSE, warning=FALSE, results='hide'}
library(forcats)
library(dplyr)
library(DT)
library(plotly)
library(ggplot2)


# Define unified color palette
color_palette <- c('#8dd3c7', '#ffffb3', '#bebada', '#fb8072', '#b3de69', '#fcd7cd')

# Basic variant counts
sql_file_path_pts <- "/workspaces/starr-oncology-data-lake-arpah/src/feb_2026/sql_std/omop/variant_occ/feb26/variant_patient_counts.sql"
n_pts<- fetch_data_from_sql_file(sql_file_path_pts, yaml_file_path, release = "feb_2026")

# Cohort overlap
sql_overlap <- "/workspaces/starr-oncology-data-lake-arpah/src/feb_2026/sql_std/omop/variant_occ/feb26/variant_cohort_overlap.sql"
cohort_overlap <- fetch_data_from_sql_file(sql_overlap, yaml_file_path, release = "feb_2026")

# Additional queries for visualizations
sql_by_gene <- "/workspaces/starr-oncology-data-lake-arpah/src/feb_2026/sql_std/omop/variant_occ/feb26/variant_by_gene.sql"
variant_genes <- fetch_data_from_sql_file(sql_by_gene, yaml_file_path, release = "feb_2026")

sql_by_type <- "/workspaces/starr-oncology-data-lake-arpah/src/feb_2026/sql_std/omop/variant_occ/feb26/variant_by_type.sql"
variant_types <- fetch_data_from_sql_file(sql_by_type, yaml_file_path, release = "feb_2026")

sql_by_consequence <- "/workspaces/starr-oncology-data-lake-arpah/src/feb_2026/sql_std/omop/variant_occ/feb26/variant_by_consequence.sql"
variant_consequence <- fetch_data_from_sql_file(sql_by_consequence, yaml_file_path, release = "feb_2026")

sql_clinical_sig <- "/workspaces/starr-oncology-data-lake-arpah/src/feb_2026/sql_std/omop/variant_occ/feb26/variant_by_clinical_significance.sql"
variant_clinical <- fetch_data_from_sql_file(sql_clinical_sig, yaml_file_path, release = "feb_2026")

sql_specimen_type <- "/workspaces/starr-oncology-data-lake-arpah/src/feb_2026/sql_std/omop/variant_occ/feb26/variant_by_specimen_type.sql"
variant_specimen_type <- fetch_data_from_sql_file(sql_specimen_type, yaml_file_path, release = "feb_2026")

sql_specimen_source <- "/workspaces/starr-oncology-data-lake-arpah/src/feb_2026/sql_std/omop/variant_occ/feb26/variant_by_specimen_source.sql"
variant_specimen_source <- fetch_data_from_sql_file(sql_specimen_source, yaml_file_path, release = "feb_2026")

sql_ordered_tests <- "/workspaces/starr-oncology-data-lake-arpah/src/feb_2026/sql_std/omop/variant_occ/feb26/variant_ordered_tests_counts.sql"
ordered_tests_counts <- fetch_data_from_sql_file(sql_ordered_tests, yaml_file_path, release = "feb_2026")

sql_ordered_trend <- "/workspaces/starr-oncology-data-lake-arpah/src/feb_2026/sql_std/omop/variant_occ/feb26/variant_ordered_tests_trend.sql"
ordered_tests_trend <- fetch_data_from_sql_file(sql_ordered_trend, yaml_file_path, release = "feb_2026")


sql_frequency <- "/workspaces/starr-oncology-data-lake-arpah/src/feb_2026/sql_std/omop/variant_occ/feb26/variant_freqency.sql"
variant_frequency <- fetch_data_from_sql_file(sql_frequency, yaml_file_path, release = "feb_2026")

sql_phenotype_class <- "/workspaces/starr-oncology-data-lake-arpah/src/feb_2026/sql_std/omop/variant_occ/feb26/variant_by_phenotype_class.sql"
variant_phenotype_class <- fetch_data_from_sql_file(sql_phenotype_class, yaml_file_path, release = "feb_2026")
```




### Summary

The **`_variant_occurrence`** table is a custom extension to the OMOP Common Data Model (CDM) that captures detailed genetic variant information identified in patients through genomic testing. This table enables precision oncology research by providing granular variant-level data including genomic coordinates, variant types, molecular consequences, and clinical interpretations.

Currently, the table contains variants identified via the **Stanford Actionable Mutation Panel for Solid Tumors (STAMP)** test, sourced from the Epic Genomics Suite. STAMP tests processed through Epic Genomics Suite starting in August 2025 are included in this table.


Each record in the **`_variant_occurrence`** table includes:

- Genomic information (chromosome, position, DNA/amino acid changes)
- Variant classification (type, molecular consequence)
- Clinical interpretation and assessment (tier classification)
- Test metadata (test name, specimen information)
- Linkage to clinical context (visit, procedure, provider)

As of February 2026, this table includes **`r format(n_pts$n_patients, big.mark = ",")`** patients with **`r format(n_pts$n_variants, big.mark = ",")`** genetic variants.


### Key Considerations

- Only simple variants detected in STAMP tests are present in the source data; fusion variants and copy number variants are not currently reported. We expect that additional variant types will be incorporated in future releases as the dataset continues to evolve.

- The OMOP variant occurrence table does not currently contain Foundation One tests or certain types of variants detected by STAMP (both of which were previously present in the Philips data). We expect that additional STAMP-detected variant types will be available later this year.


### Cohort Distribution

The following table shows the distribution of patients with variant data across key clinical cohorts:

```{r cohort table, message=FALSE, warning=FALSE, echo=FALSE}
# Create summary table
summary_table <- cohort_overlap %>%
  mutate(
    category = case_when(
      category == "Total variant patients" ~ "Total Patients with Variants",
      category == "Variants with tumor board" ~ "Patients with Tumor Board Encounters",
      category == "Variants with thoracic cancer" ~ "Patients with Thoracic Cancer",
      category == "Variants with both tumor board and thoracic" ~ "Patients with Both TB & Thoracic",
      TRUE ~ category
    )
  ) %>%
  rename(
    "Patient Cohort" = category,
    "Number of Patients" = n_patients
  )

DT::datatable(
  summary_table,
  options = list(
    dom = 't',
    ordering = FALSE,
    pageLength = 10
  ),
  rownames = FALSE
) %>%
  formatRound(columns = "Number of Patients", digits = 0, mark = ",")
```

**Note:** Tumor board patients are identified by visit records containing "tumor board" (case-insensitive). Thoracic cancer patients are identified from NeuralFrame diagnoses with primary sites including lung, bronchus, or thymus. 


## Variant Occurrence Metrics {.tabset}
This section presents key metrics and visualizations of genetic variants identified in our patient population through STAMP genomic testing.

::: {.columns}

::: {.column width="40%"}
### 📊 Data Volume 
- **Patient Count:** `r format(n_pts$n_patients, big.mark = ",")`
- **Variant Count:** `r format(n_pts$n_variants, big.mark = ",")`
:::

::: {.column width="60%"}
### 🧬 Data Components 
- **Unique Genes:** `r format(nrow(variant_genes), big.mark = ",")`
- **Molecular Consequences:** `r format(nrow(variant_consequence), big.mark = ",")`
:::

:::


## Genomic Test Metrics {.tabset}

::: {.panel-tabset}

### Overview

This section provides metrics on the number of unique genomic tests ordered that generated variant data.

```{r test counts, message=FALSE, warning=FALSE, echo=FALSE}
# Create summary table
test_summary <- data.frame(
  Metric = c("Unique Ordered Tests", "Patients with Tests", "Total Variants Detected"),
  Count = c(
    format(ordered_tests_counts$n_unique_tests, big.mark = ","),
    format(ordered_tests_counts$n_patients, big.mark = ","),
    format(ordered_tests_counts$n_variants, big.mark = ",")
  )
)

DT::datatable(
  test_summary,
  options = list(
    dom = 't',
    ordering = FALSE,
    pageLength = 10
  ),
  rownames = FALSE
)
```

**Note:** Each unique `procedure_occurrence_id` represents a distinct genomic test order. A single test may detect multiple variants in a patient.

### Calendar Trend

The following visualization shows the monthly trend of genomic test orders over time.

```{r test trend, message=FALSE, warning=FALSE}
# Format data for visualization - aggregate daily data to monthly
library(lubridate)
df_trend <- ordered_tests_trend %>%
  mutate(order_date = as.Date(order_date),
         year_month = floor_date(order_date, "month")) %>%
  group_by(year_month) %>%
  summarise(
    n_tests = sum(n_tests),
    n_patients = sum(n_patients),
    n_variants = sum(n_variants),
    .groups = "drop"
  )

# Create the plot
plot_ly(df_trend, x = ~year_month, y = ~n_tests, type = "bar",
        marker = list(color = color_palette[1]),
        hovertemplate = paste(
          "<b>%{x|%B %Y}</b><br>",
          "Tests: %{y:,.0f}<br>",
          "Patients: %{customdata:,.0f}<br>",
          "<extra></extra>"
        ),
        customdata = ~n_patients) %>%
  layout(
    xaxis = list(title = "Order Month"),
    yaxis = list(title = "Number of Unique Tests"),
    title = "Genomic Test Orders by Month",
    hovermode = "x"
  )
```

:::

## Variant Characteristics {.tabset}

::: {.panel-tabset}

### Top Genes

The following visualization shows the most frequently mutated genes in the cohort.

```{r top genes, message=FALSE, warning=FALSE}
# Top 20 genes by variant count
df_genes <- variant_genes %>% 
  filter(!is.na(gene_name)) %>%
  arrange(desc(n_variants)) %>%
  head(20) %>%
  mutate(gene_name = fct_reorder(gene_name, n_variants))

plot_ly(df_genes,
        y = ~gene_name,
        x = ~n_variants,
        type = "bar",
        marker = list(color = color_palette[1]),
        hovertemplate = paste(
          "<b>%{y}</b><br>",
          "Variants: %{x:,.0f}<br>",
          "Patients: %{customdata:,.0f}<br>",
          "<extra></extra>"
        ),
        customdata = ~n_patients) %>%
  layout(
    xaxis = list(title = "Number of Variants"),
    yaxis = list(title = ""),
    title = "",
    margin = list(l = 100)
  )
```

### Molecular Consequences

```{r consequences, message=FALSE, warning=FALSE}
df_consequence <- variant_consequence %>% 
  filter(!is.na(variant_molecular_consequence)) %>%
  arrange(desc(n_variants)) %>%
  head(15) %>%
  mutate(variant_molecular_consequence = fct_reorder(variant_molecular_consequence, n_variants))

plot_ly(df_consequence,
        y = ~variant_molecular_consequence,
        x = ~n_variants,
        type = "bar",
        marker = list(color = color_palette[3]),
        customdata = ~n_patients,
        hovertemplate = paste(
          "<b>%{y}</b><br>",
          "Variants: %{x:,.0f}<br>",
          "Patients: %{customdata:,.0f}<br>",
          "<extra></extra>"
        )) %>%
  layout(
    xaxis = list(title = "Number of Variants"),
    yaxis = list(title = ""),
    title = "",
    margin = list(l = 150)
  )
```

**Key Consequence Types:**

- **Missense:** Amino acid change
- **Nonsense:** Premature stop codon
- **Frameshift:** Reading frame disruption
- **Splice site:** Affects RNA splicing

### Clinical Classification

Distribution of variants by phenotype-specific variant classification.

```{r phenotype class, message=FALSE, warning=FALSE}
df_phenotype <- variant_phenotype_class %>% 
  filter(!is.na(phenotype_spec_var_class)) %>%
  arrange(desc(n_variants)) %>%
  mutate(phenotype_spec_var_class = fct_reorder(phenotype_spec_var_class, n_variants))

plot_ly(df_phenotype,
        y = ~phenotype_spec_var_class,
        x = ~n_variants,
        type = "bar",
        marker = list(color = color_palette[5]),
        customdata = ~n_patients,
        hovertemplate = paste(
          "<b>%{y}</b><br>",
          "Variants: %{x:,.0f}<br>",
          "Patients: %{customdata:,.0f}<br>",
          "<extra></extra>"
        )) %>%
  layout(
    xaxis = list(title = "Number of Variants"),
    yaxis = list(title = ""),
    title = "",
    margin = list(l = 200)
  )
```

**Clinical Significance:**

- **Pathogenic/Likely Pathogenic:** Variants with established or strong evidence for disease association and potential therapeutic implications
- **Benign/Likely Benign:** Variants with no expected clinical impact
- **VUS (Variant of Uncertain Significance):** Variants requiring further investigation to determine clinical relevance
- **Other classifications:** May include context-dependent or conditional pathogenicity assessments

:::

## Specimen Type and Specimen Source {.tabset}

::: {.panel-tabset}

### Specimen Type
Distribution of variants by specimen type.

```{r specimen type, message=FALSE, warning=FALSE}
df_specimen_type <- variant_specimen_type %>% 
  filter(!is.na(specimen_type)) %>%
  arrange(desc(n_variants))

plot_ly(df_specimen_type,
        labels = ~specimen_type,
        values = ~n_variants,
        type = "pie",
        textposition = "inside",
        textinfo = "label",
        customdata = ~n_patients,
        hovertemplate = paste(
          "<b>%{label}</b><br>",
          "Variants: %{value:,.0f}<br>",
          "Patients: %{customdata:,.0f}<br>",
          "Percent: %{percent}<br>",
          "<extra></extra>"
        ),
        marker = list(colors = rep(color_palette, length.out = nrow(df_specimen_type)))) %>%
  layout(title = "Variants by Specimen Type",
         showlegend = TRUE)
```

**Common Specimen Types:**

- **Tissue:** Solid tumor biopsies
- **Blood:** Liquid biopsies, circulating tumor DNA
- **Other sources:** Bone marrow, fluid samples

### Specimen Source

Distribution of variants by specimen anatomic source.

```{r specimen source, message=FALSE, warning=FALSE}
df_specimen_source <- variant_specimen_source %>% 
  filter(!is.na(specimen_source)) %>%
  arrange(desc(n_variants)) %>%
  head(15) %>%
  mutate(specimen_source = fct_reorder(specimen_source, n_variants))

plot_ly(df_specimen_source,
        y = ~specimen_source,
        x = ~n_variants,
        type = "bar",
        marker = list(color = color_palette[6]),
        hovertemplate = paste(
          "<b>%{y}</b><br>",
          "Variants: %{x:,.0f}<br>",
          "Patients: %{customdata:,.0f}<br>",
          "<extra></extra>"
        ),
        customdata = ~n_patients) %>%
  layout(
    xaxis = list(title = "Number of Variants"),
    yaxis = list(title = ""),
    title = "",
    margin = list(l = 150)
  )
```


:::

## Allelic Frequency Distribution {.tabset}

::: {.panel-tabset}

### By Molecular Consequence

Distribution of allelic frequencies across different variant molecular consequences.

```{r allelic frequency, message=FALSE, warning=FALSE, fig.width=10, fig.height=6}
# Filter and prepare data
df_freq <- variant_frequency %>%
  filter(!is.na(variant_molecular_consequence),
         !is.na(allelic_frequency),
         allelic_frequency > 0) %>%
  # Get top consequence types for readability
  mutate(consequence_count = n(), .by = variant_molecular_consequence) %>%
  filter(consequence_count >= 10) %>%
  mutate(variant_molecular_consequence = fct_reorder(variant_molecular_consequence, allelic_frequency, .fun = median))

# Create sina plot
ggplot(df_freq, aes(x = variant_molecular_consequence, y = allelic_frequency)) +
  geom_violin(fill = color_palette[4], alpha = 0.3, color = NA) +
  geom_jitter(alpha = 0.3, size = 1, width = 0.2, color = color_palette[4]) +
  stat_summary(fun = median, geom = "point", size = 3, color = "black", shape = 18) +
  labs(
    x = "Variant Molecular Consequence",
    y = "Allelic Frequency",
    title = ""
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold")
  ) +
  scale_y_continuous(labels = scales::percent_format())
```

**Interpretation:** Each point represents a variant's allelic frequency. The diamond markers indicate the median allelic frequency for each consequence type. Allelic frequency represents the proportion of DNA reads containing the variant allele.

:::



### Source Code
The source code for this page can be found [here](https://github.com/susom/starr-oncology-data-lake-arpah/tree/main/src/feb_2026/variant_occ.qmd), and the SQL queries that support these metrics can be found [here](https://github.com/susom/starr-oncology-data-lake-arpah/tree/main/src/feb_2026/sql_std/variant).

